---
title: "length of stay in itu parameterisation"
author: "Rob Challen"
date: "14 April 2020"
output: html_document
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
devtools::load_all("~/Git/uk-covid-datatools/")
devtools::load_all("~/Git/standard-print-output/")
source("~/Git/uk-covid-datatools/data-raw/rawDataFunctions.R")
library(survival)
library(patchwork)

paths=findDatafile()
standardPrintOutput::setDefaults()
```

```{r}
CHESS = getCHESS(paths$chess)
CHESSClean = CHESS %>% chessAdmissionSubset()
CHESSClean %>% select(
  trustcode, 
  trustname,
  records,
  patients,
  updatedRecords,
  knownOutcomes,
  outcomeWithoutDates,
  knownAdmittedItu,
  knownAdmittedItuPercent,
  knownOutcomePercent
) %>% distinct() %>% write.csv("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/QC_trustsIncluded.csv")

  

```



## Survival model - Admission to ITU Admission to Death or

```{r}

# admissionToDischargeOrDeathUnadj = CHESSClean %>% filter(age>10) %>%
#   mutate(status = ifelse(!is.na(finaloutcome),1,0), time)
# meanKnownOutcome = mean(admissionToDischargeOrDeathUnadj$status)
# dataQualityCheck = admissionToDischargeOrDeathUnadj %>% group_by(trustcode, status) %>% summarise(n=n(), sumTime = sum(time)) %>% 
#   pivot_wider(names_from = status, values_from = c("n", "sumTime")) %>% 
#   mutate(
#     N=n_0+n_1,
#     knownOutcome = n_1/(N),
#     meanTime=(sumTime_0+sumTime_1)/(N)
#   )
# excl2 = dataQualityCheck %>% filter(knownOutcome < (meanKnownOutcome) - sd(dataQualityCheck$knownOutcome))
# CHESSClean = CHESSClean %>% anti_join(excl2, by="trustcode")

admissionToItuAdmission = CHESSClean %>% filter(age>10) %>% generateSurvivalData(
  idVar = caseid,
  startDateVar = hospitaladmissiondate, 
  endDateExpr = if_else(is.na(dateadmittedicu), finaloutcomedate, dateadmittedicu),
  statusExpr = case_when(
    !is.na(dateadmittedicu) ~ 1,
    !is.na(finaloutcomedate) ~ 2,
    TRUE ~ 0),
  ageVar = "age", 
  censoredDateExpr = censorDate,
  statusLabels = c("censored","admitted ITU", "not admitted ITU"),
  ageBreaks = c(-Inf,50,70,80,Inf), 
  ageLabels = c("10-50","50-70","70-80","80+"),
  ageReferenceCat = "10-50"
) %>% select(
     caseid, status, trustcode, finaloutcome, time, ageCat, sex
 )
```

```{r}
CHESSClean %>% filter(age>10) %>% mutate(
  ageCat = cut(age,
    breaks = c(-Inf,50,70,80,Inf), 
    labels = c("10-50","50-70","70-80","80+")
    )) %>% group_by(ageCat) %>% summarise(
  admitted = n(),
  knownOutcome = sum(if_else(is.na(finaloutcomedate),0,1)),
  knownItu = sum(if_else(is.na(dateadmittedicu),0,1))
)
```

# summarise data

```{r}
summary = bind_rows(
  admissionToItuAdmission %>% mutate(value = status) %>% group_by(value) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(admissionToItuAdmission)*100)) %>% mutate(category="Outcome"),
  
  admissionToItuAdmission %>% group_by(value = sex) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(admissionToItuAdmission)*100)) %>% mutate(category="Gender", value = as.character(value)),
  admissionToItuAdmission %>% group_by(value = ageCat) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(admissionToItuAdmission)*100)) %>% mutate(category="Age", value = as.character(value)) %>% arrange(value),
  
#  tibble(category="Time",value = sprintf("%1.2f days [%1.2f; %1.2f]", mean(admissionToItuAdmission$time), ))
  
  admissionToItuAdmission %>% summarise(N=n()) %>% mutate(category="Total", value = "", `%age`="" )
)

summary %>% group_by(category) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/Table1_summary")
```

# fit a survival model

```{r}
fitByAgeCat = survfit(Surv(time,status, type="mstate") ~ age, admissionToItuAdmission %>% rename(age = ageCat))
plot = survminer::ggcompetingrisks(
  fitByAgeCat, pval=TRUE, conf.int = TRUE, palette = "Accent"
)

(plot+standardPrintOutput::defaultFigureLayout()) %>% 
  standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/Fig1_cumIncidenceByAge")

# summary(admissionToDischargeOrDeath)

survModel1 = coxph(Surv(time, status == "admitted ITU") ~ sex + age, admissionToItuAdmission %>% rename(age=ageCat))
# survModel2 = coxph(Surv(time, status, type="mstate") ~ sex + age, (admissionToItuAdmission %>% rename(age=ageCat)), id=caseid)
# survModel1 %>% gtsummary::tbl_regression(exp = TRUE) 


survminer::ggforest(survModel1) %>% 
  standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/Fig2_coxModel")

#summary(survModel2)

```

# Parameter fitting

So we are going to use the same sample - i.e. admissions to look into probabilities of admission to ITU
by fitting a survival model and looking at the admission probability at some distant future time.

This should give us an estimate of the overall probability of admission that accounts for some of the right censoring.

```{r}
censoredAdmissionToItuAdmission = CHESSClean %>% filter(age>15) %>% generateSurvivalData(
  idVar = caseid,
  startDateVar = hospitaladmissiondate, 
  endDateExpr = if_else(is.na(dateadmittedicu),finaloutcomedate,dateadmittedicu),
  statusExpr = case_when(
      !is.na(dateadmittedicu) ~ 1,
      !is.na(finaloutcomedate) ~ 2,
      TRUE ~ 0),
  censoredDateExpr = censorDate,
  statusLabels = c("censored","admit ITU","no ITU")
) %>% select(
  age, ageCat, time, status
)

tmp = survfit(Surv(time,status, type="mstate") ~ ageCat, censoredAdmissionToItuAdmission)
tmp2 = summary(tmp, times=1000, extend=TRUE)

ituAdmissionByAge = tibble(
  ageCat = ordered(levels(censoredAdmissionToItuAdmission$ageCat)[-3:0]),
  p_admit_itu = tmp2$pstate[,2],
  low_ci = tmp2$lower[,2],
  high_ci = tmp2$upper[,2]
)

write.csv(ituAdmissionByAge, "~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/probability-admission-by-age.csv")

(ggplot(ituAdmissionByAge, aes(x=ageCat,y=p_admit_itu,ymin=smooth(low_ci),ymax=smooth(high_ci), group=1))+geom_line(stat="smooth")+geom_ribbon(alpha=0.2)+xlab("age")+ylab("P(admission to ITU given inpatient)")) %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/Fig3_pITUAdmByAgeGivenHospAdm")

```

# Parameter fitting goodness of fit for single age group

temporal distribution of admission to ITU (given inpatient) is done using a different set of data - the subset of patients
who are already admitted to ITU and fitting the difference between hospital admission and ITU admission


```{r}

CHESSItu = CHESS %>% chessItuSubset()

uncensoredAdmissionToItuAdmission = CHESSItu %>% filter(age>15) %>% generateSurvivalData(
  idVar = caseid,
  startDateVar = hospitaladmissiondate, 
  endDateExpr = dateadmittedicu,
  statusExpr = 1,
  censoredDateExpr = censorDate,
  statusLabels = c("admit ITU")
) %>% select(
  age, ageCat, time, status
)

#ggplot(uncensoredAdmissionToItuAdmission, aes(x=time))+geom_histogram(binwidth=1)
fitdistrplus::plotdist(uncensoredAdmissionToItuAdmission$time, histo=TRUE, demp=TRUE)
fitdistrplus::descdist(uncensoredAdmissionToItuAdmission$time, discrete=FALSE, boot=500)

models = standardModels(c("gamma","weibull", "exp"))
 
subGroup = uncensoredAdmissionToItuAdmission %>% filter(ageCat=="60-64") %>% mutate(
    left=time,
    right=time
  ) %>% filter(status=="admit ITU") %>% select(left,right) %>% as.data.frame()

#fit = fitdistrplus::fitdistcens(subGroup, "exp", start=list(rate=0.5))
#fit

dists = suppressWarnings(lapply(models, function(m) fitdistrplus::fitdistcens(subGroup, m$name, start = m$start))) 
# p1 = fitdistrplus::denscomp(dists, plotstyle = "ggplot") +standardPrintOutput::narrowAndTall()
p2 = fitdistrplus::qqcompcens(dists, plotstyle = "ggplot") +standardPrintOutput::narrowAndTall()
p3 = fitdistrplus::ppcompcens(dists, plotstyle = "ggplot") +standardPrintOutput::narrowAndTall()
(p2 + p3 + patchwork::plot_layout(ncol=2)) %>% standardPrintOutput::saveQuarterPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/Fig3_fitDistributions")
```


# select log normal models

```{r}
out = uncensoredAdmissionToItuAdmission %>% fitModelsByAge(models = standardModels(c("exp", "weibull", "gamma")))
```

```{r}
out %>% mutate(value = sprintf("%1.4f [%1.4f; %1.4f]", value, low_ci, if_else(high_ci>100,as.double(NA),high_ci))) %>% select(ageCat,n,model,aic,bic,loglik,param, value) %>% 
  group_by(ageCat,n,model,aic,bic,loglik) %>% standardPrintOutput::saveTableLandscape("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/SuppTable1_fullBreakdownFittedModels")

out %>% filter(model=="exp") %>% ungroup() %>% mutate(value = sprintf("%1.4f [%1.4f; %1.4f]", value, low_ci, if_else(high_ci>100,as.double(NA),high_ci))) %>%
  select(-low_ci,-high_ci) %>% 
  pivot_wider(names_from = param, values_from = value) %>% select(
  -model,-bic,-aic,age=ageCat, N = n, `log(likelihood)`=loglik, `rate [95% CrI]` = rate
  ) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/Table2_exponentialParameterisation")
```

# generate exp curves

```{r}
survSurf = out %>% createSurvivalSurfaces(days=5)
surf = plotSurvivalSurface(survSurf, "exp")
surf %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/Fig4_losDensity")
        
```

# incidence surves more important for other transitions

```{r}
surf2 = plotIncidenceSurface(survSurf, "exp")
surf2
```

```{r}

lvl = levels(survSurf$ageCat)

survSurfcsv = out %>% createSurvivalSurfaces(timepoints=c(1:10)) %>% filter(model=="exp")
survSurfcsv = survSurfcsv %>% mutate(as.character(ageCat)) %>% ungroup() %>% inner_join(ituAdmissionByAge %>% mutate(as.character(ageCat)), by="ageCat") %>% mutate(
  ageCat = ordered(ageCat, levels = lvl),
  p_given_inpatient = p*p_admit_itu,
  cdf_given_inpatient = cdf*p_admit_itu
)
write.csv(
  survSurfcsv,
  file="~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/admission-to-itu-admission.csv")
```

# adjust for probability of ITU admission

```{r}
plotProbabilityMatrix(survSurfcsv,"exp", p_given_inpatient) %>% saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/daily_probability_by_age")

```

# validate against Bristol data

Do a set of Kolmogorov-Smirnov tests on different combinations of parameters and log normal distributions

```{r}
# bri = getBristolData(paths$bristol)
# bri = bri %>% bristolSurvivalSubset() 
# bri2 = bri %>% generateSurvivalData(
#     admission_date,
#     discharge_date
# )
# 
# doKSTest = function(d,g,paramNames,params,n) {
#   sample = d$time
#   params = as.list(params)
#   names(params) = paramNames
#   params$x = jitter(sample)
#   params$y = "plnorm"
#   #browser()
#   ks = do.call("ks.test",params)
#   return(tibble(
#     testN = length(sample),
#     referenceN = max(n),
#     pValue = ks$p.value,
#     dStatistic = ks$statistic
#   ))
# }
# 
# gofBri = bri2 %>% filter(status == 1) %>% group_by(ageCat) %>% group_modify(function(d,g,...) {
#   reference = out %>% filter(model == "lnorm" & ageCat == g$ageCat)
#   out = bind_rows(
#     doKSTest(d,g,reference$param, reference$value, reference$n) %>% mutate(estimate = "central"),
#     doKSTest(d,g,reference$param, reference$low_ci, reference$n) %>% mutate(estimate = "low credible"),
#     doKSTest(d,g,reference$param, reference$high_ci, reference$n) %>% mutate(estimate = "high credible")
#   )
# })
# 
# numbers = gofBri %>% select(ageCat,estimate,test=testN,fit=referenceN) %>% pivot_longer(cols=c("test","fit"),names_to = "source", values_to = "N")
# 
# top = ggplot(gofBri, aes(x=ageCat,y=pValue,fill=estimate))+geom_bar(stat="identity", position="dodge", width=0.7, colour="black")+standardPrintOutput::narrowAndTall()+theme(axis.text.x = element_blank(), axis.title.x = element_blank())+ylab("P value")
# bottom = ggplot(numbers, aes(x=ageCat,y=source,label=N, colour = source))+geom_text()+guides(colour="none")+theme(panel.grid = element_blank(),axis.title.x = element_blank())+ylab("N")
# 
# (top+bottom+patchwork::guide_area()+patchwork::plot_layout(ncol=1, heights = c(9,2,1), guides = "collect")) %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/Fig5_validateBristol")
```




# fit all models

```{r}
# multistateAdmissionToItuAdmission = CHESSClean %>% filter(age>15) %>% group_by(caseid, sex) %>% generateMultistateSurvivalData(
#   idVar = caseid,
#   transMatrix = transitionMatrix(
#     admitted=`ITU admission`,
#     admitted=`discharge death`
#     #`ITU admission`=`discharge death`
#   ),
#   ageBreaks = c(-Inf,50,70,80,Inf), 
#   ageLabels = c("10-50","50-70","70-80","80+"),
#   censoringDateExpr = censorDate,
#   startDateCol = "admitted",
#   admitted = hospitaladmissiondate, 
#   `ITU admission` = dateadmittedicu,
#   `discharge death` = if_else(is.na(dateadmittedicu),finaloutcomedate,as.Date(NA))
# )
```

```{r}
# 
# crweib <- flexsurv::flexsurvreg(Surv(time, status) ~ ageCat + trans + shape(trans), data = multistateAdmissionToItuAdmission, dist = "weibull")
# crgamma <- flexsurv::flexsurvreg(Surv(time, status) ~ ageCat + trans + shape(trans), data = multistateAdmissionToItuAdmission, dist = "gamma")
# # crnorm <- flexsurv::flexsurvreg(Surv(time, status) ~ ageCat + trans + sdlog(trans), data = multistateAdmissionToItuAdmission, dist = "lnorm")
# crlnorm <- flexsurv::flexsurvreg(Surv(time, status) ~ ageCat + trans + sdlog(trans), data = multistateAdmissionToItuAdmission, dist = "lnorm")
# 
# plot(crweib)
# plot(crgamma)
# plot(crlnorm)
# 
# # https://www.rdocumentation.org/packages/flexsurv/versions/1.1.1/topics/msfit.flexsurvreg
# 
# survcheck(Surv(time, status, type = "mstate") ~ sex + ageCat, multistateAdmissionToItuAdmission, id=id)
# survModel = coxph(Surv(time, status, type = "mstate") ~ sex + ageCat, multistateAdmissionToItuAdmission, id=id)
# survminer::ggforest(survModel)
# 
#crcox = coxph(Surv(time, status) ~ strata(trans), data = multistateAdmissionToItuAdmission)
# out = censoredAdmissionToItuAdmission %>% fitModelsByAge(c("lnorm"))
# 
# out %>% mutate(value = sprintf("%1.4f [%1.4f; %1.4f]", value, low_ci, high_ci)) %>% select(ageCat,n,model,aic,bic,loglik,param, value) %>% 
#   group_by(ageCat,n,model,aic,bic,loglik) %>% standardPrintOutput::saveTableLandscape("~/Dropbox/covid19/ventilator-demand/parameterisation/admission-to-itu-admission/SuppTable1_fullBreakdownFittedModels")
```
