---
title: "Current R(t): Data selection and QA"
author: "Rob Challen"
date: '`r format(Sys.Date(), "%d-%m-%Y")`'
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
    css: ~/Git/uk-covid-datatools/vignettes/ioslides.css
    # self_contained: false
    # lib_dir: ~/Dropbox/covid19/current-rt/data/libs
knit: (function(inputFile, encoding,...) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "~/Dropbox/covid19/current-rt/data", output_file=paste0('current-rt-data-',Sys.Date(),'.html')) })
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)
devtools::load_all("~/Git/standard-print-output/")
standardPrintOutput::setDefaults()
setwd("~/Git/uk-covid-datatools/vignettes/")
devtools::load_all("~/Git/uk-covid-datatools//")
source("cron-estimate-rt.R")

datasrc = function(ct, st) {
  (currentRt$rationale %>% filter(codeType==ct & statistic == st))$sources
}

proc = function(ct, st) {
  (currentRt$rationale %>% filter(codeType==ct & statistic == st))$processing
}

```

## Background

- Complex selection of data for different geographical resolutions
-- Triage; Cases; Hospital Admissions; ICU Admissions; Deaths
-- CTRY; UK; NHSER
- Inconsistencies & other quality issues
- Key dates

```{r}
events = dpc$datasets$getSignificantDates() %>% filter(Label %in% c("Lockdown","VE day","Anti-racism demos"))
events %>% select(`Start date`,Classification,Label) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/current-rt/data/Table1_events")
```

# Data sources

## CTRY

- Many feeds only are available for England
- 4 public health sites publishing data
- Not all data is collected in comparable way

## Triage calls - CTRY

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(triageCTRY, denominatorExpr = population/1000000, events = events, colour=subgroup)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig1_triageCTRY")
})
```

Sources: `r datasrc("CTRY","triage")`

## Triage calls - CTRY (continued...)

```{r}
with(currentRt$source, {
  p=tsp$plotIncidenceQuantiles(finalTriageCTRY, denominatorExpr = population/1000000, events = events)+facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig2_finalTriageCTRY")
})
```

Processing: `r proc("CTRY","triage")`

## Cases - CTRY

```{r}
with(currentRt$source, {
  p=tsp$plotIncidenceQuantiles(finalCasesCTRY, denominatorExpr = population/1000000, events = events)+facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150,500))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig3_casesCTRY")
})
```

Sources: `r datasrc("CTRY","case")`

Processing: `r proc("CTRY","case")`

## Admissions - CTRY

```{r}
with(currentRt$source, {
  p=tsp$plotIncidenceRollmean(admissionsCTRY, denominatorExpr = population/1000000, colour=source, events = events) +
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,2,5,10,20,50))+
    ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig4_admissionsCTRY")
})
```

Sources: `r datasrc("CTRY","hospital admission")`

## Admissions - CTRY (continued...)

```{r}
with(currentRt$source, {
  p=tsp$plotIncidenceQuantiles(finalAdmissionsCTRY, denominatorExpr = population/1000000, events = events) +facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,2,5,10,20))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig5_finalAdmissionsCTRY")
})
```

Processing: `r proc("CTRY","hospital admission")`

## ICU Admissions - CTRY

- incidence only available for England?
- prevalence figures for rest of UK
- not followed up on at present

## Deaths - CTRY

```{r}
with(currentRt$source, {
  p=tsp$plotIncidenceRollmean(deathsCTRY, denominatorExpr = population/1000000, events = events, colour=source) +facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,2,5,10,20))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig6_deathsCTRY")
})
```

Sources: `r datasrc("CTRY","death")`

## Deaths - CTRY (continued...)

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(finalDeathsCTRY, denominatorExpr = population/1000000, events = events) +facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,2,5,10,20))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig7_finalDeathsCTRY")
})
```

Processing: `r proc("CTRY","death")`

## UK

- No 111 & 999 calls at UK level
- Summary statistics usually best inferred from 4 nations totals
- Only UK level statistic is top level cases and deaths published on PHE coronavirus website 
-- number differs from sum of 4 nations

## Cases - UK

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(casesUK, denominatorExpr = population/1000000, events = events, colour = source)+scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig8_casesUK")
})
```

Sources: `r datasrc("UK","case")`

## Admissions - UK

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(finalAdmissionsUK, denominatorExpr = population/1000000, events = events)+scale_y_continuous(trans="log1p", breaks = c(0,2,5,10,20,50))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig9_admissionsUK")
})
```

Processing: `r proc("UK","case")`

## Deaths - UK

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(deathsUK, denominatorExpr = population/1000000, events = events, colour = source)+scale_y_continuous(trans="log1p", breaks = c(0,2,5,10,20))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig10_deathsUK")
})
```

Sources: `r datasrc("UK","death")`

Processing: `r proc("UK","death")`

## NHSER

- principal source is line lists for death and cases - England specific source
- cases are broken down by Pillar 1 and Pillar 2
-- can be by date of specimen or date of report
- deaths are broken down by setting, 
-- can be by date of death or date of report

## Triage calls - NHSER

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceRollmean(triageNHSER, denominatorExpr = population/1000000, events = events, colour=subgroup)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig11_triageNHSER")
})
```

Sources: `r datasrc("NHSER","triage")`

## Triage calls - NHSER (continued...)

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(finalTriageNHSER, denominatorExpr = population/1000000, events = events)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+
    ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig12_finalTriageNHSER")
})
```

Processing: `r proc("NHSER","triage")`

## Cases - NHSER

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceRollmean(casesNHSER, denominatorExpr = population/1000000, events = events, colour=subgroup)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+
    ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig13_casesNHSER")

  # tsp$plotIncidenceRollmean(cases, colour=name)+facet_wrap(vars(subgroup))+ scale_y_continuous(trans="log1p")
  # tmp = cases %>% tsp$imputeAndWeeklyAverage() %>% tsp$estimateRtQuick()
  # p1 = tsp$plotRt(cases,colour=name, events = events)+facet_grid(cols = vars(subgroup), rows=vars(name))
  # p1
  # p2 = tsp$plotGrowthRate(cases,colour=name, events = events)+facet_grid(cols = vars(subgroup), rows=vars(name))
  # p2

})
```

Sources: `r datasrc("NHSER","case")`

## Cases - NHSER (continued...)

```{r}
with(currentRt$source, {
  p=tsp$plotIncidenceQuantiles(finalCasesNHSER, denominatorExpr = population/1000000, events = events)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+
    ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig14_finalCasesNHSER")
})
```

Processing: `r proc("NHSER","case")`

## Admissions - NHSER

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceRollmean(admissionsNHSER, denominatorExpr = population/1000000, colour=source, events = events)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+
    ylab("per 100K per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig15_admissionsNHSER")

#tsp$plotGrowthRate(admissions %>% tsp$aggregateAge(), events = events)+facet_grid(cols=vars(source), rows=vars(name))
#tsp$plotRt(admissions %>% tsp$aggregateAge(), events = events)+facet_grid(cols=vars(source), rows=vars(name))


})
```

Sources: `r datasrc("NHSER","hospital admission")`

## Admissions - NHSER (continued...)

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(finalAdmissionsNHSER, denominatorExpr = population/1000000, events = events)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p",breaks=c(0,5,15,50,150))+
    ylab("per 100K per day")
  
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig16_finalAdmissionsNHSER")

})
```

Processing: `r proc("NHSER","hospital admission")`

## ICU Admissions - NHSER

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(finalIcuAdmissionsNHSER, denominatorExpr = population/1000000, events = events)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,2,5,10))+
    ylab("per 100K per day")
  
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig17_finalIcuAdmissionsNHSER")
})
```

Sources: `r datasrc("NHSER","icu admission")`

Processing: `r proc("NHSER","icu admission")`
  
## Deaths - NHSER

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(deathsNHSER, denominatorExpr = population/1000000, events = events, colour=subgroup)+
    facet_wrap(vars(name))+
    scale_y_continuous(trans="log1p", breaks = c(0,1,3,10,30))+
    ylab("per 1M per day")
  
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig18_deathsNHSER")
})
```

Sources: `r datasrc("NHSER","death")`

## Deaths - NHSER (continued...)

```{r}
with(currentRt$source, {
  p = tsp$plotIncidenceQuantiles(finalDeathsNHSER, denominatorExpr = population/1000000, events = events)+facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+ylab("per 1M per day")
  p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig19_finalDeathsNHSER")
})
```

Processing: `r proc("NHSER","death")`

## Combined data set

```{r}
p = tsp$plotIncidenceQuantiles(currentRt$rt, denominatorExpr = population/1000000, events = events, colour=statistic, ylim=c(0,150))+
  facet_wrap(vars(name),ncol=4)+
  theme(legend.position = "bottom")+
  scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150))+ylab("per 1M per day")
p %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig20_AllGeogsAllModels")

```  

## Combined data set - 7 day $R_t$

```{r}
events2 = dpc$datasets$getSignificantDates() %>% filter(Label %in% c("VE day"))

p2 = tsp$plotRt(
    currentRt$rt %>% filter(statistic != "icu admission"), 
    events = events2,dates = "2020-05-01", 
    colour=statistic, 
    rtlim = c(0.5,1.5), ribbons=TRUE)+
  facet_wrap(vars(name),ncol=4)+
  theme(legend.position = "bottom")
p2 %>% 
  standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig21_Rt7_AllGeogsAllModels")

```
 
## Combined data set - 7 day growth rates

```{r}
p3 = tsp$plotWindowedGrowthRate(
    currentRt$rt %>% filter(statistic != "icu admission"), 
    colour=statistic, 
    events = events2,
    dates = "2020-05-01",
    ribbons = TRUE, 
    rlim = c(-0.125,0.125))+
  facet_wrap(vars(name),ncol=4)+
  theme(legend.position = "bottom")
p3 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig22_GrowthRate7_AllGeogsAllModels")

```

## Combined data set - 28 day $R_t$

```{r}
events2 = dpc$datasets$getSignificantDates() %>% filter(Label %in% c("VE day"))

p2 = tsp$plotRt(
    currentRt$rt28 %>% filter(statistic != "icu admission"), 
    events = events2,dates = "2020-05-01", 
    colour=statistic, 
    rtlim = c(0.5,1.5), ribbons=TRUE)+
  facet_wrap(vars(name),ncol=4)+
  theme(legend.position = "bottom")
p2 %>% 
  standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig23_Rt28_AllGeogsAllModels")

```
 
## Combined data set - 28 day growth rates

```{r}
p3 = tsp$plotWindowedGrowthRate(
    currentRt$rt28 %>% filter(statistic != "icu admission"), 
    colour=statistic, 
    events = events2,
    dates = "2020-05-01",
    ribbons = TRUE, 
    rlim = c(-0.125,0.125))+
  facet_wrap(vars(name),ncol=4)+
  theme(legend.position = "bottom")
p3 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/current-rt/data/Fig24_GrowthRate28_AllGeogsAllModels")

```

```{r}

export7day = currentRt$rt %>% filter(!(name == "Wales" & statistic == "case")) %>% filter(source %in% c("4NationsCases","Deaths","Triage"))
export28day = currentRt$rt28 %>% filter(!(name == "Wales" & statistic == "case")) %>% filter(source %in% c("4NationsCases","Deaths","Triage"))

  ## create SPI-M export ----
  exportSPIM7day = bind_rows(
    export7day %>%
      ungroup() %>%
      filter(!is.na(`Median(R)`)) %>% 
      convertRtToSPIM(name,modelExpr = paste0("EpiEstim/",source), version="0.01"),
    tsp$rtConfig %>% convertSerialIntervalToSPIM(modelName = "EpiEstim/4NationsCases", version="0.01"),
    tsp$rtConfig %>% convertSerialIntervalToSPIM(modelName = "EpiEstim/Deaths", version="0.01"),
    tsp$rtConfig %>% convertSerialIntervalToSPIM(modelName = "EpiEstim/Triage", version="0.01"),
    
    export7day %>%
      ungroup() %>%
      filter(!is.na(`Median(R)`)) %>% 
      convertGrowthRateToSPIM(name,modelExpr = paste0("EpiEstim/",source), version="0.01")
  )
  
  excelOutput <- openxlsx::createWorkbook()
  openxlsx::addWorksheet(excelOutput, "Template")
  openxlsx::writeData(excelOutput, sheet = "Template", x = exportSPIM7day)
  openxlsx::saveWorkbook(excelOutput, paste0("~/Dropbox/covid19/current-rt/SPIMestimates-",Sys.Date(),"-7Day.xlsx"), overwrite = TRUE)
  
  ## create SPI-M export ----
  exportSPIM28Day = bind_rows(
    export28day %>%
      ungroup() %>%
      filter(!is.na(`Median(R)`)) %>% 
      convertRtToSPIM(name,modelExpr = paste0("EpiEstim/",source), version="0.02"),
    tsp$rtConfig %>% convertSerialIntervalToSPIM(modelName = "EpiEstim/4NationsCases", version="0.02"),
    tsp$rtConfig %>% convertSerialIntervalToSPIM(modelName = "EpiEstim/Deaths", version="0.02"),
    tsp$rtConfig %>% convertSerialIntervalToSPIM(modelName = "EpiEstim/Triage", version="0.02"),
    
    export28day %>%
      ungroup() %>%
      filter(!is.na(`Median(R)`)) %>% 
      convertGrowthRateToSPIM(name,modelExpr = paste0("EpiEstim/",source), version="0.02")
    # cfg %>% convertSerialIntervalToSPIM(modelName = "EpiEstim/PHETrackerCases")
  )
  
  excelOutput <- openxlsx::createWorkbook()
  openxlsx::addWorksheet(excelOutput, "Template")
  openxlsx::writeData(excelOutput, sheet = "Template", x = exportSPIM28Day)
  openxlsx::saveWorkbook(excelOutput, paste0("~/Dropbox/covid19/current-rt/SPIMestimates-",Sys.Date(),"-28Day.xlsx"), overwrite = TRUE)
```