---
title: "Age and COVID-19 transmission during Autumn 2020 in England"
author: "Rob Challen"
date: '`r format(Sys.Date(), "%d-%m-%Y")`'
output:
  pdf_document
knit: (function(inputFile, encoding,...) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "~/Dropbox/covid19/by-age", output_file=paste0('covid-growth-by-age-',Sys.Date(),'.pdf') ) })
# output:
#   word_document :
#     fig_caption: yes
#     fig_width: 7
# knit: (function(inputFile, encoding,...) {
#   rmarkdown::render(
#     inputFile,
#     encoding = encoding,
#     output_dir = "~/Dropbox/covid19/by-age", output_file=paste0('covid-growth-by-age-',Sys.Date(),'.docx') ) 
#     })
# TODO: https://www.reed.edu/data-at-reed/software/R/markdown_multiple_reports.html
header-includes:
 \usepackage{float}
 \floatplacement{figure}{H}    
fig_width: 7
fig_height: 5
out.width: "100%"
bibliography: current-rt.bib
csl: sage-vancouver.csl
---
```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  warning = FALSE,
  message = FALSE
)

devtools::load_all("~/Git/standard-print-output/")
devtools::load_all("~/Git/uk-covid-datatools/")
library(tidyverse)
library(patchwork)
library(rgdal)
library(ggplot2)
library(ggspatial)
library(rgeos)
library(maptools)
library(patchwork)
library(sp)
library(sf)

ggplot2::theme_set(standardPrintOutput::defaultFigureLayout())

standardPrintOutput::setDefaults()

cap = list(
  fig = captioner::captioner(prefix="Figure"),
  tab = captioner::captioner(prefix="Table"),
  sfig = captioner::captioner(prefix="Supplemental figure"),
  stab = captioner::captioner(prefix="Supplemental table")
)

ref = list(
  fig = function(name) cap$fig(name,display="cite"),
  tab = function(name) cap$tab(name,display="cite"),
  sfig = function(name) cap$sfig(name,display="cite"),
  stab = function(name) cap$stab(name,display="cite")
)

dpc = tsp = chp = srv = mwp = NULL

reload = function() {
  devtools::load_all("~/Git/uk-covid-datatools/")
  dpc <<- DataProviderController$setup("~/Data/maps/", "~/S3/encrypted/")
  tsp <<- dpc$timeseriesProcessor()
  # tsp$printSerialInterval()
  chp <<- dpc$chessProcessor()
  srv <<- dpc$survivalProcessor()
  mwp <<- dpc$metawardProcessor()
}

reload()
```

# Authors

Robert Challen^1,2^;  Krasimira Tsaneva-Atanasova^1,3,5^; Ellen Brooks-Pollock^4^; Leon Danon^3,5^

1) EPSRC Centre for Predictive Modelling in Healthcare, University of Exeter, Exeter, Devon, UK.
2) Taunton and Somerset NHS Foundation Trust, Taunton, Somerset, UK.
3) The Alan Turing Institute, British Library, 96 Euston Rd, London NW1 2DB, UK.
4) Bristol Medical School, Population Health Sciences, University of Bristol, Bristol, UK
5) Data Science Institute, College of Engineering, Mathematics and Physical Sciences, University of Exeter, Exeter, UK. 

```{r}
# casesPerPillar = dpc$spim$getLineListIncidence(ageBreaks = c(5,15,25,35,45,55,65,75,85),subgroup = pillar)
# casesPerSymptomatic = dpc$spim$getLineListIncidence(ageBreaks = c(5,15,25,35,45,55,65,75,85),subgroup = asymptomatic_indicator)
# casesPerEthnicity = dpc$spim$getLineListIncidence(ageBreaks = c(5,15,25,35,45,55,65,75,85),subgroup = ethnicity_final)
# casesPerIMD = dpc$spim$getLineListIncidence(ageBreaks = c(5,15,25,35,45,55,65,75,85),subgroup = imd_decile)
# 
# breakdown111 = dpc$spim$getOneOneOneIncidence(dateFrom = as.Date("2020-07-01"))
# 
# deathsPerResidence = dpc$spim$getDeathsLineListIncidence(ageBreaks = c(5,15,25,35,45,55,65,75,85),subgroup = residence_type)
# deathsPerEthicity = dpc$spim$getDeathsLineListIncidence(ageBreaks = c(5,15,25,35,45,55,65,75,85),subgroup = ethnicity_final)
# deaths = dpc$spim$getDeathsLineListIncidence(ageBreaks = c(5,15,25,35,45,55,65,75,85))

#others = dpc$spim$getSPIMextract()
#icuAdmissions = others %>% filter(source == "chess_icu_admissions" & !is.na(ageCat) & codeType %in% c("CTRY","NHSER"))admissions = others %>% filter(source %in% c("hospital_inc","hospital_inc_new") & !is.na(ageCat) & codeType %in% c("CTRY","NHSER"))

ageBreaks = c(6,11,18,23,60,75);

cases = dpc$spim$getLineListIncidence(ageBreaks = ageBreaks,subgroup = asymptomatic_indicator) %>% filter(ageCat !="unknown")
deaths = dpc$spim$getDeathsLineListIncidence(ageBreaks = ageBreaks)

events = dpc$datasets$getSignificantDates(nocache=TRUE) %>% 
  filter(Label %in% c("Lockdown","VE day","Hotels and bars reopen","Relaxation shielding","Shielding ends (excl Wales)","End summer term","Start autumn term","Start univerity term","England 2nd lockdown"))

events2 = dpc$datasets$getSignificantDates(nocache=TRUE) %>% 
  filter(Label %in% c("Start autumn term","Start univerity term","Testing capacity limited","England 2nd lockdown","School half term"))

events3 = dpc$datasets$getSignificantDates(nocache=TRUE) %>% 
  filter(Label %in% c("Start autumn term","North West England tier 3","School half term","England 2nd lockdown"))


```

# Introduction

We conducted an analysis of the evolving second wave of the COVID-19 outbreak in the England, with a focus on the interaction between age and regional variation. We focused our analysis on pre-school, school (primary and secondary), university, working, early and later retirement age groups and on the various administrative regions of the NHS in England. The purpose of this was to gather information on how the second wave evolved, and improve our view of the current state of the outbreak across the country. To do this we investigated case and death line list data where we have a detail on the age and location of individuals affected by COVID-19.

# Materials and methods

* Case data is pillar 1 & 2 test data from the line list provided by PHE[@publichealthenglandPHEDataSeries2020]. We excluded Pillar 2 cases marked a "asymptomatic". These data are potentially affected by saturation when test demand outstrips availability. This bias is discussed below and see `r ref$sfig("SymptomaticVsAsymptomatic")` for details.
* Deaths data is from the line list provided by PHE[@publichealthenglandPHEDataSeries2020].
* Maps & demographic information is from the ONS, and based on mid year 2019 estimates.[@PopulationEstimatesOffice]
* Daily exponential growth rates calculated from the gradient of a one sided local linear regression on log of incidence of cases or deaths.
* Daily growth rate estimates are combined over a 14 day window up to the date of the estimate.

# Results

## England wide

```{r}
totalCases = cases %>% filter(subgroup!="Y") %>% tsp$aggregateSubgroup() %>% tsp$aggregateGender()
totalDeaths = deaths %>% tsp$aggregateSubgroup() %>% tsp$aggregateGender() %>% filter(ageCat %in% c("23-59","60-74","75+"))
combined = bind_rows(totalCases,totalDeaths)  %>% tsp$logIncidenceStats(growthRateWindow = 14) %>% tsp$adjustGrowthRateDates() %>% dpc$demog$findDemographics()

p1 = (combined %>% filter(codeType=="CTRY") %>% tsp$plotIncidenceQuantiles(dates = "2020-08-01", events = events2, colour=statistic, denominatorExpr = population/1000000)) + facet_wrap(vars(ageCat)) + scale_y_continuous(trans="log1p", breaks = c(0,2,5,20,50,200,500,2000))+ylab("rate per 1M per day")

p1 %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/by-age/CaseAndDeathCountsByAge")
```

`r cap$fig("incidCTRY","Incidence and death rates per day per 1M people, for pre-school, primary school, secondary school, university, working age, late middle age and elderly age groups, over the course of Autumn.")`

* Incidence in early August was highest per-capita in university and working age groups.
* A fast increase in university age groups lead to a peak in late September / early October, with a slow decline since.
* True size of university age peak may be underestimated due to saturation of testing capacity in September.
* Rapid increase from low level in primary, and less so secondary, school age groups started before the beginning of term. 
* Size of this outbreak is again likely to be an underestimate due to testing capacity issues.
* Deaths in older age groups reached nadir around late August, after which cases have been growing. 
* Exponential growth in older age groups seen to slow during half term.
* Following second lockdown curves begin to flatten in all age groups except those in school which return to growth after half term break.

```{r}
p2 = combined %>% filter(codeType=="CTRY") %>% tsp$plotWindowedGrowthRate(dates = "2020-08-01", events = events2, colour=statistic, rlim = c(-0.1,0.2)) + facet_wrap(vars(ageCat))
# tsp$estimateRt(window = 14) %>%  tsp$plotRt(dates = "2020-08-01", events = events2) + facet_wrap(vars(ageCat))
p2 %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/by-age/CaseAndDeathGrowthByAge")
```

`r cap$fig("growthCTRY","Exponential growth rates of cases and deaths in various age groups. Growth rates are shifted in time to align to estimated date of infection, rather than date of case identification or death.")`

* Most age groups show a transition to positive growth around the mid-late August, university age groups are seen to be growing from as early as the beginning of August.
* True peak of growth rate in younger age groups is unknown as potentially affected by testing capacity issues. 
* Growth has become negative in university age groups, although this is could be affected by access to and compliance with testing programmes.
* Growth rate in cases in older age groups has been positive since late August, and this finding is reinforced by death statistics. If the relationship between cases and deaths stayed constant this suggests the ascertainment bias due to saturation of test availability has not affected this group, possibly as older age groups will have been prioritised for testing.
* Growth rates in older age groups are reducing over half term but remain well above zero.
* Pick up in growth rates post half term in school age groups.

## Regional

```{r}
#p2 = 
tmp = combined %>% filter(codeType=="NHSER" & statistic=="case" & name!="unknown")

tmp %>% tsp$plotWindowedGrowthRate(dates = "2020-08-01", events = events3, group=name, rlim = c(-0.1,0.2),colour="grey60") +
  plotRibbons(data = tmp %>% filter(name=="North West"),meanVar = `Growth.windowed.value`,sdVar = `Growth.windowed.SE.value`,colourExpr = "red") +
  plotRibbons(data = tmp %>% filter(name=="South West"),meanVar = `Growth.windowed.value`,sdVar = `Growth.windowed.SE.value`,colourExpr = "blue") +
  facet_wrap(vars(ageCat))
# tsp$estimateRt(window = 14) %>%  tsp$plotRt(dates = "2020-08-01", events = events2) + facet_wrap(vars(ageCat))
#p2 %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/by-age/CaseAndDeathGrowthByAge")
```

`r cap$fig("growthNHSER","Exponential growth rates of cases in various age groups in the different NHS regions in England. Red is the North West, Blue is the South West. Other regions not highlighted. Growth rates are shifted in time to align to estimated date of infection.")`

* Growth shows different dynamics in different regions and national picture in `r ref$fig("growthCTRY")` is a superposition of the different regions.
* For example the bi-phasic peak in primary school ages seen in `r ref$fig("growthCTRY")` is driven by 2 regional peaks initially in North West, and latterly in the South West.
* Growth in school ages initially very much higher in North West compared to South West until late September, when South West begun to grow faster.
* Growth rate in North West now seen to be approaching zero in all age groups.
* Growth rate has slowed since early to mid October in all regions.
* In the most vulnerable age groups growth currently highest in the South West. 
* increases in growth rate in school children after lockdown seen across all regions but particularly in the South West.

```{r}

dmg = dpc$demog$getDemographicsForShape(mapId = "LAD19",ageBreaks = ageBreaks, combineGenders = TRUE)

tmp = dpc$spim$getLineListIncidence(ageBreaks = ageBreaks, codeTypes = "LAD",subgroup=asymptomatic_indicator) %>% 
  filter(ageCat !="unknown" & subgroup!="Y") %>% mutate(ageCat=as.character(ageCat)) %>%
  tsp$aggregateGender() %>% tsp$aggregateSubgroup() %>% inner_join(dmg %>% mutate(ageCat=as.character(ageCat)), by=c("code","ageCat"), suffix=c("","rhs")) %>% rename(population=count)

other = tmp %>% filter(ageCat %in% c("<6","23-59")) %>% tsp$aggregateAge() %>% mutate(ageCat="other family")
school = tmp %>% filter(ageCat %in% c("6-10","11-17")) %>% tsp$aggregateAge() %>% mutate(ageCat="school")
university = tmp %>% filter(ageCat %in% c("18-22")) %>% tsp$aggregateAge() %>% mutate(ageCat="university")
youngOlder = tmp %>% filter(ageCat %in% c("60-74")) 
oldOlder = tmp %>% filter(ageCat %in% c("75+")) 
all = tmp %>% tsp$aggregateAge() %>% mutate(ageCat="all")

tmp2 = bind_rows(
  other,school,university,youngOlder, oldOlder,all
) %>% mutate(ageCat = factor(ageCat,levels=c("school","university","other family","60-74","75+","all"),ordered=TRUE))
```

```{r}
casesGeog = tmp2 %>% tsp$logIncidenceStats(growthRateWindow = 7)
maxDate = (tmp2 %>% pull(date) %>% max())

map = dpc$geog$getMap("LAD19") %>% inner_join(casesGeog %>% filter(date == maxDate) %>% select(-name,-codeType), by=c("code"))
```

```{r}
p3 = ggplot(map %>% ungroup() %>% sf::st_as_sf()) + geom_sf(aes(fill=Growth.windowed.value),size = 0.05) + facet_wrap(vars(ageCat), ncol = 3) + standardPrintOutput::defaultMapLayout() + standardPrintOutput::narrower() + scale_fill_distiller(
  palette = "RdGy", name="growth rate", oob=scales::squish, lim=c(-0.10,0.10))
p3 %>% standardPrintOutput::saveTwoThirdPageFigure("~/Dropbox/covid19/by-age/CaseGrowthByAgeAndLAD")
```


`r cap$fig("growthLAD",paste0("Exponential growth rates of symptomatic cases by local authority district in functional age categories as of ",maxDate,". For clarity several age groups have been combined. School age comprises 6-17 year olds, University is from 18-22. Other family is 0-6 and 23-59, Retirement is 60 and over"))`

* Day to day variation, and very localized variation in growth rates makes trends identified in `r ref$fig("growthNHSER")` difficult to see. Only general conclusions can be made from a single day. 
* Prior to second lockdown
  * School age growth overall decreasing but with evidence of isolated outbreaks.
  * University age decreasing almost everywhere, but this is most obviously seen in university towns. This has the potential to mask growth in other age groups in those regions.
  * Young children and working age adults are still largely increasing across the country, but qualitatively appear to be levelling off in areas which have seen most significant tier 3 restrictions.
  * Older age groups appears similar to broader family age patterns but with possibly more evidence of evolving outbreaks in specific locations.
* After lockdown:
  * pattern of isolated outbreaks in school age and family age persists with growth strongest in South East.
  * Reductions in university age fairly homogeneous. 
  * lockdown seems effective at reducing transmission to the vulnerable older age gropus.

# Limitations

* The bias in the case data due to saturation is well described. However, where available death data provides a reassuring correlation to case data. Thankfully there are very few deaths in younger age groups for comparison. 
* We conducted a similar analysis including hospital case data however, we do not have admission line lists and this was based on aggregated hospital admission data, only available in larger geographical units (NHS trust) and incompatible age bands. An accurate line list of hospital admissions would be valuable.
* We do not have negative test data at the level of individual test, so we cannot get true test positivity rates. A full breakdown of negative tests at the individual test level would be helpful.

# Discussion

* Growth dynamics within regional age groups suggest the timing of the second wave has been driven by infections in school and university age people, but that growth was established in these groups well before the return their respective terms started.
* The national level is a superposition of various regional age group outbreaks, and resolving outbreaks can mask worsening picture in other age groups within the same area. The different NHS regions are not homogenous and local outbreaks in care homes coupled with falling cases in universities will not be reflected in national metrics such as the reproduction number.
* There seems to be a delay in the outbreak in regions such as the South West, compared to the North West, possibly as a result of the tiering system, however this delay is apparent before the introduction of tiering, and could reflect more fundamental differences in the nature of social and workplace contacts between the regions.
* It is possible the timing of the second lockdown will benefit regions that have had lower case loads to date, but have currently higher growth rates, such as the South West, more than those with high rates and low current growth such as the North West.

# References

<div id="refs"></div>

# Supplemental Material

```{r}
p = (cases %>% tsp$aggregateGender() %>% dpc$demog$findDemographics() %>% filter(codeType=="CTRY") %>% tsp$plotIncidenceQuantiles(colour = subgroup, denominatorExpr = population/1000000, dates = "2020-08-01", events = events2)) + facet_wrap(vars(ageCat)) + scale_y_continuous(trans="log1p", breaks = c(0,5,15,50,150,500,1500,5000,15000))+guides(color=guide_legend("asymptomatic"))
p %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/by-age/SymptomaticVsAsymptomatic")
```

`r cap$sfig("SymptomaticVsAsymptomatic","Symptomatic (red) and asymptomatic (blue) pillar 2 cases, and pillar 1 cases (green) over time. During the time period of limited testing capacity asymptomatic cases were deprioritised leading to a bias in the data. This was less pronounced (but still present) in the symptomatic pillar 2 and pillar 1 cases which we combine for our case estimate. The case count we use still shows a satuation effect particularly in younger age groups which were deprioritised, and we interpret with caution.")`

```{r}
totalCases2 = totalCases %>% tsp$logIncidenceStats()
totalCases2 = totalCases2 %>% group_by(date,codeType,code,name) %>% mutate(totalAllAges = sum(Est.value), percentAllAges = Est.value/sum(Est.value))

ggplot(totalCases2 %>% filter(name=="England" & date > "2020-03-15"),aes(x=date,y=percentAllAges*100,fill=ageCat)) + geom_area()+facet_wrap(vars(name))+ylab("%age of total") + scale_x_date(date_breaks = "2 weeks",date_labels = "%d-%m")+tsp$plotEvents(events,labelSize = 10,labelY=75)
```

`r cap$sfig("propCTRY","The relative proportion of each age group as a percentage of total cases per day for pre-school, primary school, secondary school, university, working age, late middle age and elderly age groups, over the course of the epidemic.")`

* Daily proportion of different age groups initially dominated by elderly as they were the ones being tested in hospitals.
* Proportion of younger age groups being detected as a fraction of total cases has increased steadily - much of this will be due to increased testing.
* Resulting decrease in proportion of elderly until early September but this age group not making up greater proportion of total.
* These figures not adjusted by population in age group and slightly hard to interpret.






