---
title: "Regional infectivity of COVID-19"
output: html_document
fig_width: 7
fig_height: 5
vignette: >
  %\VignetteIndexEntry{Regional infectivity of COVID-19}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyverse)

devtools::load_all("~/Git/uk-covid-datatools/")
# devtools::install_github("terminological/uk-covid-datatools")
# library(ukcovidtools)
devtools::load_all("~/Git/standard-print-output/")
library(rgdal)
library(ggplot2)
library(ggspatial)
library(rgeos)
library(maptools)
library(lubridate)
library(patchwork)
library(sp)

ggplot2::theme_set(standardPrintOutput::defaultFigureLayout())
source("./lockdown-impact-data.R")
```

## Serial interval estimation from the literature:

```{r}
SItable1 %>% group_by(`Reference`) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/lockdown-impact/Table1_serialIntervals", defaultFontSize = 8, colWidths=c(4.5,2,2,0.5,1))
```

## Used serial intervals

```{r}

#  wtSIs is the serial intervals
tdp = function(x,y,z) sprintf("%1.2f (%1.2f - %1.2f)", x ,y, z)
paste0("Serial interval mean plus 95% credible interval: ",tdp(wtSIs$mean_si, wtSIs$min_mean_si, wtSIs$max_mean_si))
paste0("Serial interval standard deviation plus 95% credible interval: ",tdp(wtSIs$std_si, wtSIs$min_std_si, wtSIs$max_std_si))

```

## Significant dates in epididemic UK

https://en.wikipedia.org/wiki/2020_coronavirus_pandemic_in_the_United_Kingdom

* Surveillance testing: https://www.gov.uk/government/news/new-surveillance-system-for-early-detection-of-covid-19: 2020-02-26
* Sport fixtures cancelled: 2020-03-13
* Tests only for inpatients: https://www.gov.uk/government/publications/wuhan-novel-coronavirus-initial-investigation-of-possible-cases/investigation-and-initial-clinical-management-of-possible-cases-of-wuhan-novel-coronavirus-wn-cov-infection: 2020-03-13
* Voluntary self isolation of any symptoms & vunerable people: https://www.gov.uk/government/publications/covid-19-guidance-on-social-distancing-and-for-vulnerable-people: 2020-03-16
* Non essential travel worldwide banned: https://www.gov.uk/guidance/travel-advice-novel-coronavirus#history: 2020-03-19
* School closures: https://www.gov.uk/government/publications/closure-of-educational-settings-information-for-parents-and-carers#history: 2020-03-19
* Full lockdown: https://www.gov.uk/government/publications/full-guidance-on-staying-at-home-and-away-from-others - 2020-03-23

```{r}
glimpse(keyDates)

```

### UK coutries

The R(t) in the different countries of the UK in figure 1. This has fallen from a high initial level (2-3) to just over 1 by 2020-04-07 in all regions.

```{r}
r0ukregionalplot = ggplot(ts$r0UKRegional, aes(x=date, y=`Median(R)`, ymin=`Quantile.0.025(R)`, ymax=`Quantile.0.975(R)`))+
  geom_ribbon(alpha=0.2)+geom_line()+
  geom_hline(yintercept = 1, colour="grey75")+
  
  coord_cartesian(ylim=c(0, 3))+
  geom_vline(aes(xintercept=date,colour=event),data=keyDates, show.legend = FALSE)+
  ggrepel::geom_text_repel(
          aes(x=date, y=Inf, colour=event, label=event),data=keyDates, hjust=0,vjust=1, angle=90, show.legend = FALSE,box.padding=0.05,inherit.aes = FALSE,
          size=(10/ggplot2:::.pt/(96/72)))+
  scale_x_date(date_breaks="2 day", date_labels = "%d-%b")+
  theme(axis.text.x=element_text(angle = 90, vjust =0.5))+
  facet_wrap(vars(uk_region))

r0ukregionalplot %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/Fig1_ukRegionRt")
```

```{r}
ts$r0EnglandNHS %>% filter(event == "Latest") %>% mutate(
  `Median(R)` = sprintf("%1.2f", `Median(R)`),
  `95% confidence interval`=sprintf("(%1.2f; %1.2f)",`Quantile.0.025(R)`,`Quantile.0.975(R)`)
) %>% select(`NHS region`= england_nhs_region,`Median(R)`,`95% confidence interval`) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/lockdown-impact/Table2_Rt_by_NHSregion", defaultFontSize = 8)
```

The rate of change of R(t) in the different countries of the UK. This shows the viral reproduction has been decelerating since 2020-03-31 in all regions.

```{r}
deltaR0ukregionalplot = ggplot(ts$r0UKRegional, aes(x=date, y=slope, ymin=slopeLowerCi, ymax=slopeUpperCi))+
  geom_ribbon(alpha=0.2)+geom_line(linetype="dashed")+
  geom_hline(yintercept = 0, colour="grey75")+
  coord_cartesian(ylim=c(-0.2, 0.2))+
  geom_vline(aes(xintercept=date,colour=event),data=keyDates, show.legend = FALSE)+
  ggrepel::geom_text_repel(
          aes(x=date, y=Inf, colour=event, label=event),data=keyDates, hjust=0,vjust=1, angle=90, show.legend = FALSE,box.padding=0.05,inherit.aes = FALSE,
          size=(10/ggplot2:::.pt/(96/72)))+
  scale_x_date(date_breaks="2 day", date_labels = "%d-%b")+
  theme(axis.text.x=element_text(angle = 90, vjust =0.5))+
  facet_wrap(vars(uk_region))

deltaR0ukregionalplot %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/Fig5_deltaUkRegionRt")
```

## NHS regions

The same analysis as above but looking at NHS regions. there is some evidence that viral replication is different in different areas.

```{r}

tmp = ts$r0UKRegional %>% filter(uk_region == "england") %>% select(date,baseline=`Median(R)`)
englandnhsFromBaseline = ts$r0EnglandNHS %>% left_join(tmp, by="date") %>% mutate(
  `R(t) from baseline`=`Median(R)`-baseline, ymin=`Quantile.0.025(R)`-baseline, ymax=`Quantile.0.975(R)`-baseline
)

#glimpse(ts$tidyEnglandNHS)
englandnhsplot = ggplot(englandnhsFromBaseline, aes(x=date, y=`R(t) from baseline`, ymin=ymin, ymax=ymax))+
  geom_ribbon(alpha=0.2)+geom_line(linetype="dotted")+
  geom_hline(yintercept = 0, colour="grey75")+
  facet_wrap(vars(england_nhs_region)) + 
  standardPrintOutput::narrowAndTall()+
  coord_cartesian(ylim=c(-1,1))+
  geom_vline(aes(xintercept=date,colour=event),data=keyDates)+
  scale_x_date(date_breaks="2 day", date_labels = "%d-%b")+
  theme(axis.text.x=element_text(angle = 90, vjust = 0.5))

englandnhsplot %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/Fig2-englandNHSRt")
```
Significance tests:

* consider using a time series specific meausure:
* e.g. https://cran.r-project.org/web/packages/TSdist/TSdist.pdf

```{r}
englandnhsFromBaselineSignificance = englandnhsFromBaseline %>% 
  filter(date > as.Date("2020-03-27")) %>% 
  group_by(england_nhs_region) %>%
  group_modify(function(d,g,...) {
    res = t.test(d$`Median(R)`,d$baseline,paired = TRUE)
    return(tibble(
      `R(t) effect size` = sprintf("%1.2f", res$estimate),
      `95% CI` = sprintf("(%1.2f; %1.2f)", res$conf.int[1], res$conf.int[2]),
      pValue = ifelse(res$p.value < 0.001, "<0.001", sprintf("%1.3f", res$p.value))
    ))
  }) %>% rename(`England NHS region`=england_nhs_region) 
englandnhsFromBaselineSignificance %>% standardPrintOutput::saveTable("~/Dropbox/covid19/lockdown-impact/Table2-regionalSignificance", defaultFontSize = 8)

```


## Unitary Authority Timeseries

The same analysis demonstrating R(t) but on a smaller regional level, plotted on a map at various time points
Create static maps

```{r}
data("UKCovidMaps")

# https://github.com/tidyverse/ggplot2/issues/3391
# some issues joining tibble onto sf - which 

r0shapes = UKCovidMaps$reportingRegions %>% 
  # fill in missing dates to prevent the map having disappearing / reappearing regions
  crossing(tibble(date=unique(ts$r0CombinedUK$date))) %>% 
  left_join(keyDates, by="date") %>%
  left_join(
    ts$r0CombinedUK,
    by=c("code","date"), suffix=c("",".dup")) %>% 
  mutate(ago=difftime(date,lubridate::now(),units="days")) %>% 
  sf::st_as_sf()



```

```{r}
defaultR0Map = function(data) {
  return(
    ggplot(data)+
    geom_sf(aes(fill=`Median(R)`))+
    scale_fill_gradient2(
      low="green",
      mid="white",
      high="red",
      midpoint=0,
      trans="log",
      na.value = "grey80", 
      limits=c(0.2,5), 
      breaks=c(0.2,0.5,1,2,5), 
      labels=c("<0.2","0.5","1","2",">5"),
      oob=scales::squish)+
    standardPrintOutput::narrowAndTall()+
    standardPrintOutput::mapTheme()
  )
}


defaultDeltaR0Map = function(data) {
  return(
    ggplot(data)+
    geom_sf(aes(fill=slope))+
    scale_fill_gradient2(
      low="cyan",
      mid="white",
      high="magenta",
      midpoint=0,
      na.value = "grey80",
      limits=c(-0.5,0.5), 
      breaks=c(-0.5,-0.25,0,0.25,0.5), 
      labels=c("<-0.5","-0.25","0","0.25",">0.5"),
      oob=scales::squish)+
    standardPrintOutput::narrowAndTall()+
    standardPrintOutput::mapTheme()
  )
}
```


```{r}
ukwide = 
  defaultR0Map(r0shapes %>% filter(!is.na(label)))+
  facet_wrap(vars(label), nrow = 1)
ukwide %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/Fig3_englandMap")

```
```{r}
london = ukwide + coord_sf(crs = 4326,xlim = c(-0.7, 0.5), ylim = c(51.25, 51.75), expand = FALSE) + standardPrintOutput::mapTheme()
london %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/lockdown-impact/Fig4_londonMap")
```
```{r}
# birmingham = ukwide + coord_sf(crs = 4326,xlim = c(-2.3, -1.5), ylim = c(52.25, 52.75), expand = FALSE) + standardPrintOutput::mapTheme()
# birmingham %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/lockdown-impact/Fig4_birminghamMap")
```


```{r}
ukwideDelta = 
  defaultDeltaR0Map(r0shapes %>% filter(!is.na(label)))+
  facet_wrap(vars(label), nrow = 1)
ukwideDelta %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/Fig6_englandMapDelta")
```


```{r}
londonDelta = ukwideDelta + coord_sf(crs = 4326,xlim = c(-0.7, 0.5), ylim = c(51.25, 51.75), expand = FALSE) + standardPrintOutput::mapTheme()
londonDelta %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/lockdown-impact/Fig7_londonMapDelta")
```
```{r}
# birminghamDelta = ukwideDelta + coord_sf(crs = 4326,xlim = c(-2.3, -1.5), ylim = c(52.25, 52.75), expand = FALSE) + standardPrintOutput::mapTheme()
# birminghamDelta %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/lockdown-impact/Fig7_birminghamMapDelta")
```

## Create and save animated map

Animated maps are viewable:

* https://github.com/terminological/uk-covid-datatools/blob/master/vignettes/UK_Rt_over_time.gif

```{r eval=FALSE}
dates = tibble(date1=as.Date(min(r0shapes$date):max(r0shapes$date),origin=as.Date("1970-01-01")))
barAnim = ggplot(dates,aes(xmax=date1+1,xmin=date1))+
  geom_rect(ymin=0,ymax=1) + 
  ylim(c(0,1))+
  geom_vline(aes(xintercept=date,colour=event),data=keyDates,size=4,show.legend = FALSE)+
  # ggrepel::geom_text_repel(
  #   aes(x=date, y=1.25, colour=event, label=event),data=keyDates, hjust=0.5,vjust=0, show.legend = FALSE,box.padding=0.05,inherit.aes = FALSE,
  #         size=(10/ggplot2:::.pt/(96/72)))+
  theme(
    axis.text.y = element_blank(),
    axis.ticks.y = element_blank(), 
    panel.grid.major.y = element_blank(), 
    panel.grid.minor.y = element_blank(), 
    axis.title.x = element_blank()
  )+
  coord_cartesian(ylim = c(0, 1), clip = 'off')+
  gganimate::transition_time(date1)

timeGif = gganimate::animate(barAnim, renderer=gganimate::magick_renderer(), width=800, height=60)


ukwideAnim = 
  defaultR0Map(r0shapes)+
  guides(fill="none")+labs(subtitle="A")+
  gganimate::transition_time(date)

londonAnim = 
  defaultR0Map(r0shapes)+
  coord_sf(crs = 4326,xlim = c(-0.7, 0.5), ylim = c(51.25, 51.75), expand = FALSE) + 
  standardPrintOutput::mapTheme()+labs(subtitle="C")+
  gganimate::transition_time(date)

ukwideDeltaAnim = 
  defaultDeltaR0Map(r0shapes)+
  guides(fill="none")+labs(subtitle="B")+
  gganimate::transition_time(date)

londonDeltaAnim = 
  defaultDeltaR0Map(r0shapes)+
  coord_sf(crs = 4326,xlim = c(-0.7, 0.5), ylim = c(51.25, 51.75), expand = FALSE) + 
  standardPrintOutput::mapTheme()+labs(subtitle="D")+
  gganimate::transition_time(date)

ukWideGif = gganimate::animate(ukwideAnim, renderer=gganimate::magick_renderer(), width=400, height=800)
londonGif = gganimate::animate(londonAnim, renderer=gganimate::magick_renderer(), width=400,height=300)
ukWideDeltaGif = gganimate::animate(ukwideDeltaAnim, renderer=gganimate::magick_renderer(), width=400, height=800)
londonDeltaGif = gganimate::animate(londonDeltaAnim, renderer=gganimate::magick_renderer(), width=400,height=300)

captionTif = magick::image_read(path = "~/Git/uk-covid-datatools/vignettes/suppFig1Caption.png")

layoutGifs = function(i) {
  magick::image_append(c(
  magick::image_append(c(ukWideGif[i], ukWideDeltaGif[i])),
  magick::image_append(c(londonGif[i], londonDeltaGif[i])),
  timeGif[i], captionTif[1]), stack=TRUE)
}

new_gif <- layoutGifs(1)
for(i in 2:100){
  combined <- layoutGifs(i)
  new_gif <- c(new_gif, combined)
}

#gganimate::anim_save("~/Git/uk-covid-datatools/vignettes/UK_Rt_over_time.gif",new_gif)
magick::image_write_gif(new_gif,path="~/Git/uk-covid-datatools/vignettes/UK_Rt_over_time.gif")
# ggplot(r0shapes, aes(x=date,y=`Median(R)`,colour=code))+geom_line(alpha=0.2,show.legend = FALSE)

```

```{r}

UKCovidNeghbours = UKCovidMaps$reportingRegions %>% createNeighbourNetwork(code)
write.csv(UKCovidNeghbours,"~/Git/uk-covid-datatools/vignettes/Rt_Timeseries_Neighbours.csv")

```

```{r}

# captionTif = magick::image_read(path = "~/Git/uk-covid-datatools/vignettes/suppFig1Caption.png")
# manGif = magick::image_read(path = "~/Dropbox/covid19/lockdown-impact/eurosurveillanceSubmission/UK_Rt_over_time_no_label.gif")
# 
# new_gif <- magick::image_append(c(manGif[1], captionTif[1]), stack=TRUE)
# for(i in 2:100){
#   combined <- magick::image_append(c(manGif[i], captionTif[1]), stack=TRUE)
#   new_gif <- c(new_gif, combined)
# }
# 
# magick::image_write_gif(new_gif,path="~/Git/uk-covid-datatools/vignettes/UK_Rt_over_time_2.gif")


# rateOfChangeDist = ggplot(rateOfChange, aes(x=slope))+geom_density()+geom_rug()+geom_vline(xintercept = 0, colour="blue")+coord_cartesian(xlim=c(-0.5,0.5))
# 
# ukRateOfChange = ggplot(rateOfChangeUKregion, aes(x=uk_region,y=slope,fill=slope))+
#   scale_fill_gradient2(
#     low="cyan",
#     mid="white",
#     high="magenta",
#     midpoint=0,
#     #trans="log",
#     na.value = "grey80",
#     limits=c(-1,1), 
#     breaks=c(-1,-0.5,0,0.5,1), 
#     labels=c("<-1","-0.5","0","0.5",">1")
#     ) + 
#   geom_bar(stat="identity", colour="black")+geom_errorbar(aes(ymin=slopeLowerCi,ymax=slopeUpperCi), width=0.2)
# 
# (ukwideRateOfChange + londonRateofChange  + ukRateOfChange + rateOfChangeDist + plot_annotation(tag_levels = 'A')  + plot_layout(nrow = 2, guides="collect")) %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/Fig4_RateOfChangeMap")
# 

```

```{r}

# rateOfChange %>% ungroup() %>% filter(slopeLowerCi > 0) %>% arrange(desc(slope)) %>% head(5) %>% mutate(`95% CI` = sprintf("%1.2f; %1.2f",slopeLowerCi,slopeUpperCi)) %>% select(`Unitary authority`=name,`dR/dt`=slope, `95% CI` ,`R^2`=r_squared) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/lockdown-impact/Table2_Top5UnitaryAuthoritiesByDeltaR_t")

```

```{r}

# rateOfChange %>% ungroup() %>% filter(slopeUpperCi < 0) %>% arrange(slope) %>% head(5) %>% mutate(`95% CI` = sprintf("%1.2f; %1.2f",slopeLowerCi,slopeUpperCi)) %>% select(`Unitary authority`=name,`dR/dt`=slope, `95% CI` ,`R^2`=r_squared) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/lockdown-impact/Table3_Bottom5UnitaryAuthoritiesByDeltaR_t")

```




```{r}

# tmpDeltaR0timeseries = 
# 
# rateOfChangeShapes = UKCovidMaps$reportingRegions %>%
#   crossing(tibble(date=unique(tmpDeltaR0timeseries$date))) %>% 
#   left_join(tmpDeltaR0timeseries, by=c("code","date")) %>%
#   sf::st_as_sf() 
# 
# ukwideRateOfChange = ggplot(rateOfChangeShapes)+
#   geom_sf(aes(fill=slope))+
#   scale_fill_gradient2(
#     low="cyan",
#     mid="white",
#     high="magenta",
#     midpoint=0,
#     trans="pseudo_log",
#     na.value = "grey80",
#     limits=c(-1,1), 
#     
#     breaks=c(-1,-0.5,0,0.5,1), 
#     labels=c("<-1","-0.5","0","0.5",">1")
#     )+
#   standardPrintOutput::mapTheme()
# 
# anim5 = ukwideRateOfChange + guides(fill="none")+#labs(title = 'Date: {frame_time}')+
#   gganimate::transition_time(date)
# gif5 = gganimate::animate(anim5, renderer=gganimate::magick_renderer(), width=400, height=800)
# 
# anim6 = ukwideRateOfChange + coord_sf(crs = 4326,xlim = c(-0.7, 0.5), ylim = c(51.25, 51.75), expand = FALSE)+
#   gganimate::transition_time(date)
# gif6 = gganimate::animate(anim6, renderer=gganimate::magick_renderer(), duration = 10, fps=10, width=400, height=300)

```

```{r}
# dates = tibble(date1=as.Date(min(rateOfChangeShapes$date):max(rateOfChangeShapes$date),origin=as.Date("1970-01-01")))
# barAnim = ggplot(dates,aes(xmax=date1+1,xmin=date1))+
#   geom_rect(ymin=0,ymax=1) + 
#   ylim(c(0,1))+
#   geom_vline(aes(xintercept=date,colour=event),data=keyDates,size=4,show.legend = FALSE)+
#   # ggrepel::geom_text_repel(
#   #   aes(x=date, y=1.25, colour=event, label=event),data=keyDates, hjust=0.5,vjust=0, show.legend = FALSE,box.padding=0.05,inherit.aes = FALSE,
#   #         size=(10/ggplot2:::.pt/(96/72)))+
#   theme(
#     axis.text.y = element_blank(),
#     axis.ticks.y = element_blank(), 
#     panel.grid.major.y = element_blank(), 
#     panel.grid.minor.y = element_blank(), 
#     axis.title.x = element_blank()
#   )+
#   coord_cartesian(ylim = c(0, 1), clip = 'off')+
#   gganimate::transition_time(date1)
# 
# timeGif = gganimate::animate(barAnim, renderer=gganimate::magick_renderer(), width=800, height=60)
```

```{r eval=FALSE}



# ukRateOfChange = ggplot(deltaR0Regionaltimeseries, aes(x=uk_region,y=slope,fill=slope))+
#   scale_fill_gradient2(
#     low="cyan",
#     mid="white",
#     high="magenta",
#     midpoint=0,
#     #trans="log",
#     na.value = "grey80",
#     limits=c(-0.5,0.5), 
#     breaks=c(-0.5,-0.25,0,0.25,0.5), 
#     labels=c("<-0.5","-0.25","0","0.25",">0.5")
#     ) + geom_bar(stat="identity", colour="black")+geom_errorbar(aes(ymin=slopeLowerCi,ymax=slopeUpperCi), width=0.2)+
#   coord_cartesian(ylim=c(-0.25,0.25))#+
#   #labs(title = 'Date: {frame_time}')
# 
# anim3 = ukRateOfChange+gganimate::transition_time(date)
# gif3 = gganimate::animate(anim3, renderer=gganimate::magick_renderer(), width=400, height = 300)
# #gganimate::anim_save("~/Git/uk-covid-datatools/vignettes/delta_Rt_by_uk_region_over_time.gif",gif)
# 
# rateOfChangeDist = ggplot(deltaR0timeseries %>% filter(!is.na(slope)), aes(x=slope))+geom_density()+geom_rug()+geom_vline(xintercept = 0, colour="blue")+coord_cartesian(xlim=c(-0.5,0.5))+labs(title = " ")
# anim4 = rateOfChangeDist+gganimate::transition_time(date)
# gif4 = gganimate::animate(anim4, renderer=gganimate::magick_renderer(), width=400, height = 200)
# 
# new_gif <- magick::image_append(c(
#       magick::image_append(c(
#         gif5[1],
#         magick::image_append(c(
#           gif3[1], 
#           gif4[1],
#           gif6[1]
#         ),stack=TRUE)
#       )),
#       timeGif[1]
# ),stack=TRUE) 
# for(i in 2:100){
#   combined <- magick::image_append(c(
#       magick::image_append(c(
#         gif5[i],
#         magick::image_append(c(
#           gif3[i], 
#           gif4[i],
#           gif6[i]
#         ),stack=TRUE)
#       )),
#       timeGif[i]
# ),stack=TRUE)
#     
#   new_gif <- c(new_gif, combined)
# }
# 
# magick::image_write_gif(new_gif,path="~/Git/uk-covid-datatools/vignettes/delta_Rt_over_time.gif")
#gganimate::anim_save("~/Git/uk-covid-datatools/vignettes/delta_Rt_over_time.gif",new_gif)
#gganimate::anim_save("~/Git/uk-covid-datatools/vignettes/delta_Rt_distribution_over_time.gif",gif)
```


```{r}
# tmp = tempfile("google",fileext = ".pdf")
# pdf = "https://www.gstatic.com/covid19/mobility/2020-03-29_GB_Mobility_Report_en.pdf"
# download.file(pdf,tmp)
```

```{r}
# pdfExtract = pdftools::pdf_text(tmp)
# pdfExtract = pdfExtract[4:length(pdfExtract)-1]
# pdfExtract[[1]] %>% stringr::str_split("\n")

```