---
title: "Impact of the social distancing on regional infectivity of COVID-19"
output: rmarkdown::html_vignette
fig_width: 7
fig_height: 7
vignette: >
  %\VignetteIndexEntry{Impact of the social distancing on regional infectivity of COVID-19}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(tidyverse)

devtools::load_all("~/Git/uk-covid-datatools/")
# devtools::install_github("terminological/uk-covid-datatools")
# library(ukcovidtools)
devtools::load_all("~/Git/standard-print-output/")
library(rgdal)
library(ggplot2)
library(ggspatial)
library(rgeos)
library(maptools)
library(lubridate)
library(patchwork)
library(sp)

ggplot2::theme_set(standardPrintOutput::defaultFigureLayout())
```

## Serial interval estimation from the literature:

```{r}
serialIntervals = tibble(
  mean_si_estimate = c(3.96, 6.3, 4.22, 4.56, 3.95, 5.21, 4.7, 7.5,6.6),
  mean_si_estimate_low_ci = c(3.53, 5.2, 3.43, 2.69,-4.47, -3.35, 3.7, 5.3, 0.7),
  mean_si_estimate_high_ci = c(4.39, 7.6, 5.01, 6.42, 12.51,13.94, 6.0, 19.0, 19.0),
  std_si_estimate = c(4.75,4.2, 0.4, 0.95, 4.24, 4.32, 2.3, 3.4, NA),
  std_si_estimate_low_ci = c(4.46, 3.1, NA, NA, 4.03, 4.06, 1.6, NA, NA),
  std_si_estimate_high_ci = c(5.07, 5.3, NA, NA, 4.95, 5.58, 3.5, NA, NA),
  sample_size = c(468,48,135,93,45,54,28,16,90),
  population = c("China", "Shenzhen","Taijin","Singapore","Taijin","Singapore", "SE Asia", "Wuhan","Italy"),
  source = c(
    "Zhanwei Du et al. Serial Interval of COVID-19 among Publicly Reported Confirmed Cases. Emerging Infectious Disease journal 26, (2020)",
    "Bi, Q. et al. Epidemiology and Transmission of COVID-19 in Shenzhen China: Analysis of 391 cases and 1,286 of their close contacts. Infectious Diseases (except HIV/AIDS) (2020) doi:10.1101/2020.03.03.20028423",
    "Tindale, L. et al. Transmission interval estimates suggest pre-symptomatic spread of COVID-19. Epidemiology (2020) doi:10.1101/2020.03.03.20029983",
    "Tindale, L. et al. Transmission interval estimates suggest pre-symptomatic spread of COVID-19. Epidemiology (2020) doi:10.1101/2020.03.03.20029983",
    "Ganyani, T. et al. Estimating the generation interval for COVID-19 based on symptom onset data. Infectious Diseases (except HIV/AIDS) (2020) doi:10.1101/2020.03.05.20031815",
    "Ganyani, T. et al. Estimating the generation interval for COVID-19 based on symptom onset data. Infectious Diseases (except HIV/AIDS) (2020) doi:10.1101/2020.03.05.20031815",
    "Nishiura, H., Linton, N. M. & Akhmetzhanov, A. R. Serial interval of novel coronavirus (COVID-19) infections. Int. J. Infect. Dis. (2020) doi:10.1016/j.ijid.2020.02.060",
    "Li, Q. et al. Early Transmission Dynamics in Wuhan, China, of Novel Coronavirus-Infected Pneumonia. N. Engl. J. Med. (2020) doi:10.1056/NEJMoa2001316",
    "Cereda, D. et al. The early phase of the COVID-19 outbreak in Lombardy, Italy. arXiv [q-bio.PE] (2020)")
)

unk=function(x) ifelse(is.na(x),"unk",x)

table = serialIntervals %>% mutate(
  `Mean SI\n(95% CrI) days`=paste0(mean_si_estimate,"\n(",unk(mean_si_estimate_low_ci),"-",
                                   unk(mean_si_estimate_high_ci),")"),
  `Std SI\n(95% CrI) days`=paste0(unk(std_si_estimate),"\n(",unk(std_si_estimate_low_ci),"-",unk(std_si_estimate_high_ci),")")
) %>% select(-contains("estimate")) %>% select(
  `Reference`=source,
  `Mean SI\n(95% CrI) days`,
  `Std SI\n(95% CrI) days`,
  `N`=sample_size,
  `Population`=population
)

table %>% group_by(`Reference`) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/lockdown-impact/serialIntervals", defaultFontSize = 8, colWidths=c(4.5,2,2,0.5,1))
```

```{r}

wtSIs = serialIntervals %>% summarise(
  mean_si = weighted.mean(mean_si_estimate,sample_size,na.rm = TRUE),
  min_mean_si = weighted.mean(mean_si_estimate_low_ci,sample_size,na.rm = TRUE),
  max_mean_si = weighted.mean(mean_si_estimate_high_ci,sample_size,na.rm = TRUE),
  std_si  = weighted.mean(ifelse(is.na(std_si_estimate_low_ci),NA,1)*std_si_estimate,sample_size,na.rm = TRUE),
  min_std_si  = weighted.mean(std_si_estimate_low_ci,sample_size,na.rm = TRUE),
  max_std_si  = weighted.mean(std_si_estimate_high_ci,sample_size,na.rm = TRUE)
  #total = sum(sample_size)
) %>% mutate(
  std_mean_si = (max_mean_si - min_mean_si) / 3.92, # TODO: fit gamma
  std_std_si = (max_std_si - min_std_si) / 3.92
)

tdp = function(x,y,z) sprintf("%1.2f (%1.2f - %1.2f)", x ,y, z)

tdp(wtSIs$mean_si, wtSIs$min_mean_si, wtSIs$max_mean_si)
tdp(wtSIs$std_si, wtSIs$min_std_si, wtSIs$max_std_si)

```

## Significant dates in epididemic UK

https://en.wikipedia.org/wiki/2020_coronavirus_pandemic_in_the_United_Kingdom

* Surveillance testing: https://www.gov.uk/government/news/new-surveillance-system-for-early-detection-of-covid-19: 2020-02-26
* Sport fixtures cancelled: 2020-03-13
* Tests only for inpatients: https://www.gov.uk/government/publications/wuhan-novel-coronavirus-initial-investigation-of-possible-cases/investigation-and-initial-clinical-management-of-possible-cases-of-wuhan-novel-coronavirus-wn-cov-infection: 2020-03-13
* Voluntary self isolation of any symptoms & vunerable people: https://www.gov.uk/government/publications/covid-19-guidance-on-social-distancing-and-for-vulnerable-people: 2020-03-16
* Non essential travel worldwide banned: https://www.gov.uk/guidance/travel-advice-novel-coronavirus#history: 2020-03-19
* School closures: https://www.gov.uk/government/publications/closure-of-educational-settings-information-for-parents-and-carers#history: 2020-03-19
* Full lockdown: https://www.gov.uk/government/publications/full-guidance-on-staying-at-home-and-away-from-others - 2020-03-23

```{r}

keyDates = tibble(
  date = as.Date(c("2020-03-13","2020-03-16","2020-03-19","2020-03-23")), #max(r0shapes$date-1, na.rm=TRUE)),
  impactDate = as.Date(c("2020-03-14","2020-03-21","2020-03-24","2020-03-28")), #max(r0shapes$date-1, na.rm=TRUE)),
  event = c("Inpatient only testing","Social isolation of vulnerable","Travel ban / school closure","Stay at home") #,"Latest")
) %>% mutate(label = paste0(date,": \n",event))

```

## Construct a R_t timeseries for Regional breakdowns 

```{r}
ts = ukcovidtools::getUKCovidTimeseries()
#TODO: update these estimates of SI and use credible interval
# cfg = EpiEstim::make_config(list(
#   mean_si = wtSIs$mean_si,
#   std_si = wtSIs$std_si
# ))

cfg = EpiEstim::make_config(list(
  mean_si = wtSIs$mean_si, 
  std_mean_si = wtSIs$std_mean_si,
  min_mean_si = wtSIs$min_mean_si, 
  max_mean_si = wtSIs$max_mean_si,
  std_si = wtSIs$std_si, 
  std_std_si = wtSIs$std_si,
  min_std_si = wtSIs$min_std_si, 
  max_std_si = wtSIs$max_std_si), method="uncertain_si")
```

### UK coutries

```{r}
#glimpse(ts$tidyUKRegional)
R0regionaltimeseries = ts$tidyUKRegional %>% group_by(uk_region) %>% normaliseAndCleanse() %>% tidyEstimateRt(cfg, window = 5) 
R0regionaltimeseries = R0regionaltimeseries %>% filter(!is.na(`Median(R)`))
ukregionalplot = ggplot(R0regionaltimeseries, aes(x=date, y=`Median(R)`, ymin=`Quantile.0.025(R)`, ymax=`Quantile.0.975(R)`))+ #, colour=uk_region, fill=uk_region))+
  geom_ribbon(alpha=0.2)+geom_line()+geom_hline(yintercept = 1, colour="grey50", linetype="dashed")+facet_wrap(vars(uk_region)) + 
  coord_cartesian(ylim=c(0, 5))+
  geom_vline(aes(xintercept=date,colour=event),data=keyDates, show.legend = FALSE)+
  #geom_vline(aes(xintercept=impactDate,colour=event,linetype="dashed"),data=keyDates, show.legend = FALSE)+
  ggrepel::geom_text_repel(
          aes(x=date, y=Inf, colour=event, label=event),data=keyDates, hjust=0,vjust=1, angle=90, show.legend = FALSE,box.padding=0.05,inherit.aes = FALSE,
          size=(10/ggplot2:::.pt/(96/72)))+
  scale_x_date(date_breaks="1 day", date_labels = "%d-%b")+theme(axis.text.x=element_text(angle = 90, vjust =0.5))
ukregionalplot %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/ukRegionRt")
```

### NHS regions

```{r}
#glimpse(ts$tidyEnglandNHS)
R0nhstimeseries = ts$tidyEnglandNHS %>% group_by(england_nhs_region) %>% normaliseAndCleanse() %>% tidyEstimateRt(cfg, window = 5) 
R0nhstimeseries = R0nhstimeseries %>% filter(!is.na(`Median(R)`))
englandnhsplot = ggplot(R0nhstimeseries, aes(x=date, y=`Median(R)`, ymin=`Quantile.0.025(R)`, ymax=`Quantile.0.975(R)`))+ #, colour=uk_region, fill=uk_region))+
  geom_ribbon(alpha=0.2)+geom_line()+
  geom_hline(yintercept = 1, colour="grey50", linetype="dashed")+facet_wrap(vars(england_nhs_region)) + 
  standardPrintOutput::narrowAndTall()+
  coord_cartesian(ylim=c(0, 5))+
  geom_vline(aes(xintercept=date,colour=event),data=keyDates)+
  #geom_vline(aes(xintercept=impactDate,colour=event,linetype="dashed"),data=keyDates, show.legend = FALSE) #, show.legend = FALSE) #+
  # ggrepel::geom_text_repel(
  #         aes(x=date, y=Inf, colour=event, label=event),data=keyDates, hjust=0,vjust=1, angle=90, show.legend = FALSE,box.padding=0.05,inherit.aes = FALSE,
  #         size=(10/ggplot2:::.pt/(96/72)))
  scale_x_date(date_breaks="1 day", date_labels = "%d-%b")+theme(axis.text.x=element_text(angle = 90, vjust = 0.5))
englandnhsplot %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/englandNHSRt")
```

## Unitary Authoriy Timeseries

Initial focus is on localised Rt at Unitary Authority level timeseries in England:

```{r}

R0timeseries = ts$tidyEnglandUnitAuth %>% normaliseAndCleanse() %>% group_by(GSS_CD, GSS_NM) %>% tidyEstimateRt(cfg, window=5)

write_csv(R0timeseries, "~/Git/uk-covid-datatools/vignettes/Supplementary_Rt_Timeseries_by_Unitary_Authority.csv")

```

Create static maps

```{r}
data("UKCovidMaps")

# https://github.com/tidyverse/ggplot2/issues/3391
# some issues joining tibble onto sf - which 

r0shapes = UKCovidMaps$unitaryAuthority %>% 
  left_join(R0timeseries, by=c("ctyua19cd"="GSS_CD")) %>% 
  mutate(ago=difftime(date,lubridate::now(),units="days")) %>% 
  filter(!is.na(date))

r0shapes = r0shapes %>% mutate(`Median(R)` = ifelse(`Median(R)`>10, 9.999,`Median(R)`))

r0shapes_key = r0shapes %>% inner_join(keyDates, by="date")

ukwide = ggplot(r0shapes_key)+
  geom_sf(aes(fill=`Median(R)`),data=r0shapes_key)+
  scale_fill_gradient2(
    low="green",
    mid="white",
    high="red",
    midpoint=0,
    trans="log",
    na.value = "grey80", 
    limits=c(0.1,10), 
    breaks=c(0.1,0.4,1,2.5,10), 
    labels=c("<0.1","0.4","1","2.5",">10"))+
  standardPrintOutput::narrowAndTall()+
  standardPrintOutput::mapTheme()+
  facet_wrap(vars(label), nrow = 1)
  
london = ukwide + coord_sf(crs = 4326,xlim = c(-0.7, 0.5), ylim = c(51.25, 51.75), expand = FALSE)
  
  # layout <- c(
  #  patchwork::area(t = 1, l = 1, b = 10, r = 10),
  #  patchwork::area(t = 1, l = 1, b = 3, r = 3)
  # )
  # combined = ukwide / london + plot_layout(guides="collect")

ukwide %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/lockdown-impact/englandMap")
london %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/lockdown-impact/londonMap")

```


create and save animated map

```{r}

ukwide = ggplot(r0shapes)+
  geom_sf(aes(fill=`Median(R)`), data=r0shapes)+
  scale_fill_gradient2(
    low="green",
    mid="white",
    high="red",
    midpoint=0,
    trans="log",
    na.value = "grey80", 
    limits=c(0.1,10), 
    breaks=c(0.1,0.4,1,2.5,10), 
    labels=c("<0.1","0.4","1","2.5",">10"))+
  standardPrintOutput::narrowAndTall()+
  standardPrintOutput::mapTheme()

london = ukwide + coord_sf(crs = 4326,xlim = c(-0.7, 0.5), ylim = c(51.25, 51.75), expand = FALSE)

anim = ukwide+gganimate::transition_time(date)
gif = gganimate::animate(anim, renderer=gganimate::gifski_renderer())
gganimate::anim_save("~/Git/uk-covid-datatools/vignettes/Rt_by_unitary_authority_over_time.gif",gif)

anim2 = london+gganimate::transition_time(date)
gif2 = gganimate::animate(anim2, renderer=gganimate::gifski_renderer())
gganimate::anim_save("~/Git/uk-covid-datatools/vignettes/London_Rt_by_unitary_authority_over_time.gif",gif2)


```

## Unitary authority trajectories

* Basically select the last 5 days
* fit a linear model
* get the gradient

```{r}
# UA level
lastDate = max(R0timeseries$date)-5

toFit = R0timeseries %>% filter(date >= lastDate) %>% select(GSS_CD, GSS_NM, date, r=`Median(R)`)
rateOfChange = toFit %>% group_by(GSS_CD,GSS_NM) %>% group_modify(function(d,g,...) {
  lmResult = lm(r~date, d)
  slope = coef(lmResult)["date"]
  return(tibble(slope=slope))
})

r0rateOfChange = UKCovidMaps$unitaryAuthority %>% 
  inner_join(rateOfChange, by=c("ctyua19cd"="GSS_CD")) %>% mutate(slope = ifelse(slope < -1, -1, ifelse( slope > 1, 1, slope)))

# National level
toFit = R0regionaltimeseries %>% filter(date >= lastDate) %>% select(uk_region, date, r=`Median(R)`)
rateOfChangeUKregion = toFit %>% group_by(uk_region) %>% group_modify(function(d,g,...) {
  lmResult = lm(r~date, d)
  slope = coef(lmResult)["date"]
  return(tibble(slope=slope))
})

```

```{r}
ukwideRateOfChange = ggplot(r0rateOfChange)+
  geom_sf(aes(fill=slope))+
  scale_fill_gradient2(
    low="cyan",
    mid="white",
    high="magenta",
    midpoint=0,
    #trans="log",
    na.value = "grey80",
    limits=c(-1,1), 
    breaks=c(-1,-0.5,0,0.5,1), 
    labels=c("<-1","-0.5","0","0.5",">1")
    )+
  standardPrintOutput::mapTheme()

londonRateofChange = ukwideRateOfChange + coord_sf(crs = 4326,xlim = c(-0.7, 0.5), ylim = c(51.25, 51.75), expand = FALSE)

rateOfChangeDist = ggplot(rateOfChange, aes(x=slope))+geom_density()+geom_rug()+geom_vline(xintercept = 0, colour="blue")+coord_cartesian(xlim=c(-1,1))

ukRateOfChange = ggplot(rateOfChangeUKregion, aes(x=uk_region,y=slope,fill=slope))+
  scale_fill_gradient2(
    low="cyan",
    mid="white",
    high="magenta",
    midpoint=0,
    #trans="log",
    na.value = "grey80",
    limits=c(-1,1), 
    breaks=c(-1,-0.5,0,0.5,1), 
    labels=c("<-1","-0.5","0","0.5",">1")
    ) + geom_bar(stat="identity", colour="black")

(ukwideRateOfChange + londonRateofChange  + ukRateOfChange + rateOfChangeDist + plot_annotation(tag_levels = 'A')  + plot_layout(nrow = 2, , guides="collect")) %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/lockdown-impact/RateOfChangeMap")


```

```{r}
rateOfChange %>% ungroup() %>% arrange(desc(slope)) %>% head(5) %>% select(`Unitary authority`=GSS_NM,`dR/dt`=slope) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/lockdown-impact/Top5UnitaryAuthoritiesByDeltaR_t")
```