---
title: "Proportion and timing of Symptomatic to admission"
author: "Rob Challen"
date: '`r format(Sys.Date(), "%d-%m-%Y")`'
output: 
  pdf_document :
    fig_caption: yes
header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}
knit: (function(inputFile, encoding,...) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission", output_file="symptomatic-to-admission.pdf" })
fig_width: 7
fig_height: 5
out.width: "100%"
bibliography: current-rt.bib
csl: current-rt.csl
vignette: >
  %\VignetteIndexEntry{COVID-19 Serial Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
devtools::load_all("~/Git/uk-covid-datatools/")
devtools::load_all("~/Git/standard-print-output/")
library(survival)
library(patchwork)
standardPrintOutput::setDefault()
dpc = DataProviderController$setup("~/Data/maps")
dpc$loadSpimSources("~/S3/encrypted/")
tsp = dpc$timeseriesProcessor()
chp = dpc$chessProcessor()
srv = dpc$survivalProcessor()
```

```{r}
CHESSClean = dpc$spim$getCHESS() %>% chp$chessAdmissionSubset()

CHESSClean %>% select(
  trustcode, 
  trustname,
  records,
  patients,
  updatedRecords,
  knownOutcomes,
  outcomeWithoutDates,
  knownAdmittedItu,
  knownAdmittedItuPercent,
  knownOutcomePercent
) %>% distinct() %>% write.csv("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/QC_trustsIncluded.csv")

  

```



## Survival model - Symptomatic to admission

```{r}

# admissionToDischargeOrDeathUnadj = CHESSClean %>% filter(age>10) %>%
#   mutate(status = ifelse(!is.na(finaloutcome),1,0), time)
# meanKnownOutcome = mean(admissionToDischargeOrDeathUnadj$status)
# dataQualityCheck = admissionToDischargeOrDeathUnadj %>% group_by(trustcode, status) %>% summarise(n=n(), sumTime = sum(time)) %>% 
#   pivot_wider(names_from = status, values_from = c("n", "sumTime")) %>% 
#   mutate(
#     N=n_0+n_1,
#     knownOutcome = n_1/(N),
#     meanTime=(sumTime_0+sumTime_1)/(N)
#   )
# excl2 = dataQualityCheck %>% filter(knownOutcome < (meanKnownOutcome) - sd(dataQualityCheck$knownOutcome))
# CHESSClean = CHESSClean %>% anti_join(excl2, by="trustcode")

symptomaticToAdmission = CHESSClean %>% filter(age>10 & !is.na(estimateddateonset) & !is.na(hospitaladmissiondate)) %>% srv$generateSurvivalData(
  idVar = caseid,
  startDateVar = estimateddateonset, 
  endDateExpr = hospitaladmissiondate,
  statusExpr = 1,
  ageVar = "age", 
  censoredDateExpr = censorDate,
  statusLabels = c("admitted"),
  ageBreaks = c(-Inf,50,70,80,Inf), 
  ageLabels = c("10-50","50-70","70-80","80+"),
  ageReferenceCat = "10-50"
) %>% select(
     caseid, status, trustcode, finaloutcome, time, ageCat, sex
 )
```

# summarise data

```{r}
summary = bind_rows(
  symptomaticToAdmission %>% mutate(value = status) %>% group_by(value) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(symptomaticToAdmission)*100)) %>% mutate(category="Outcome"),
  symptomaticToAdmission %>% group_by(value = sex) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(symptomaticToAdmission)*100)) %>% mutate(category="Gender", value = as.character(value)),
  symptomaticToAdmission %>% group_by(value = ageCat) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(symptomaticToAdmission)*100)) %>% mutate(category="Age", value = as.character(value)) %>% arrange(value),
  
#  tibble(category="Time",value = sprintf("%1.2f days [%1.2f; %1.2f]", mean(symptomaticToAdmission$time), ))
  
  symptomaticToAdmission %>% summarise(N=n()) %>% mutate(category="Total", value = "", `%age`="" )
)

summary %>% group_by(category) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/Table1_summary")
```

# fit a survival model

```{r}
fitByAgeCat = survfit(Surv(time) ~ age, symptomaticToAdmission %>% rename(age = ageCat))
plot = survminer::ggsurvplot(
  fitByAgeCat, pval=TRUE, conf.int = TRUE, palette = "Accent"
)

(plot$plot+standardPrintOutput::defaultFigureLayout()) %>% 
  standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/Fig1_cumIncidenceByAge")

# summary(admissionToDischargeOrDeath)

survModel1 = coxph(Surv(time) ~ sex + age, symptomaticToAdmission %>% rename(age=ageCat))
# survModel2 = coxph(Surv(time, status, type="mstate") ~ sex + age, (symptomaticToAdmission %>% rename(age=ageCat)), id=caseid)
# survModel1 %>% gtsummary::tbl_regression(exp = TRUE) 


survminer::ggforest(survModel1) %>% 
  standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/Fig2_coxModel")

#summary(survModel2)

```

# Parameter fitting

So we are going to use the same sample - i.e. admissions to look into probabilities of admission to ITU
by fitting a survival model and looking at the admission probability at some distant future time.

This should give us an estimate of the overall probability of admission that accounts for some of the right censoring.

```{r}
uncensoredSymptomaticToAdmission = CHESSClean %>% filter(age>15) %>% srv$generateSurvivalData(
  idVar = caseid,
  startDateVar = estimateddateonset, 
  endDateExpr = hospitaladmissiondate,
  statusExpr = 1,
  censoredDateExpr = censorDate,
  statusLabels = c("admitted")
) %>% select(
  age, ageCat, time, status
)

#ggplot(uncensoredsymptomaticToAdmission, aes(x=time))+geom_histogram(binwidth=1)
fitdistrplus::plotdist(uncensoredSymptomaticToAdmission$time, histo=TRUE, demp=TRUE)
fitdistrplus::descdist(uncensoredSymptomaticToAdmission$time, discrete=FALSE, boot=500)

models = srv$standardModels(c("gamma","weibull", "lnorm"))
 
subGroup = uncensoredSymptomaticToAdmission %>% mutate(
    left=time,
    right=time
  ) %>% filter(status=="admitted") %>% select(left,right) %>% as.data.frame()

#fit = fitdistrplus::fitdistcens(subGroup, "exp", start=list(rate=0.5))
#fit

dists = suppressWarnings(lapply(models, function(m) fitdistrplus::fitdistcens(subGroup, m$name, start = m$start))) 
# p1 = fitdistrplus::denscomp(dists, plotstyle = "ggplot") +standardPrintOutput::narrowAndTall()
p2 = fitdistrplus::qqcompcens(dists, plotstyle = "ggplot") +standardPrintOutput::narrowAndTall()
p3 = fitdistrplus::ppcompcens(dists, plotstyle = "ggplot") +standardPrintOutput::narrowAndTall()
(p2 + p3 + patchwork::plot_layout(ncol=2)) %>% standardPrintOutput::saveQuarterPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/Fig3_fitDistributions")
```

#TODO: get output of unage stratified models.

# select log normal models

```{r}
out = uncensoredSymptomaticToAdmission %>% fitModelsByAge(models = srv$standardModels(c("lnorm", "weibull", "gamma")))
```

```{r}
out %>% mutate(value = sprintf("%1.4f [%1.4f; %1.4f]", value, low_ci, if_else(high_ci>100,as.double(NA),high_ci))) %>% select(ageCat,n,model,aic,bic,loglik,param, value) %>% 
  group_by(ageCat,n,model,aic,bic,loglik) %>% standardPrintOutput::saveTableLandscape("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/SuppTable1_fullBreakdownFittedModels")

out %>% filter(model=="weibull") %>% ungroup() %>% mutate(value = sprintf("%1.4f [%1.4f; %1.4f]", value, low_ci, if_else(high_ci>100,as.double(NA),high_ci))) %>%
  select(-low_ci,-high_ci) %>% 
  pivot_wider(names_from = param, values_from = value) %>% select(
  -model,-bic,-aic,age=ageCat, N = n, `log(likelihood)`=loglik, `scale [95% CrI]` = scale, `shape [95% CrI]` = shape
  ) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/Table2_weibullParameterisation")
```

# generate exp curves

```{r}
survSurf = out %>% createSurvivalSurfaces(days=10)
surf = plotSurvivalSurface(survSurf, "weibull")
surf %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/Fig4_durationSymptomsToAdmissionDensity")
        
```

# incidence surves more important for other transitions

```{r}
surf2 = plotIncidenceSurface(survSurf, "weibull")
surf2
```
# median distribution

```{r}

(ggplot(survSurf %>% filter(model=="gamma") %>% mutate(median_lo = ifelse(is.na(median_lo),0,median_lo)),aes(x=ageCat,y=median,ymin=median_lo,ymax=median_hi, group=1))+geom_line(stat="smooth")+geom_ribbon(alpha=0.2)+xlab("age")+ylab("median duration symptoms before admission")) %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/Fig5_durationSymptomsToAdmissionMedianByAge")

```

# positive cases (i.e. symptomatic) to admission probability

To estimate this we need to know how likely it is that someone who is symptomatic is admitted.
There is no good definition of symptomatic.
We can use a range of proxy measures for symptomatic or infected

* People testing positive in South Korea => we assume that south korea is catching the majority of patients who are infected who are known positive versus hospital admissions
* People testing positive in Spain => we assume that Spain (like the UK) is testing only those with high probability of disease with severe symptoms
* 111 calls. We assume that those calling 111 are symptomatic, and those being referred for urgent investigation are severe. - no age breakdown though.

# adjust for probability of ITU admission

```{r}
survSurfcsv = out %>% createSurvivalSurfaces(timepoints = 1:10)
write.csv(
  survSurfcsv %>% ungroup() %>% mutate(surv = 1-cdf),
  file="~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/symptomatic-to-admission.csv")        
plotProbabilityMatrix(survSurfcsv,"weibull", p) %>% saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/daily_probability_by_age")

```

# validate against Bristol data

Do a set of Kolmogorov-Smirnov tests on different combinations of parameters and log normal distributions

```{r}
# bri = getBristolData(paths$bristol)
# bri = bri %>% bristolSurvivalSubset() 
# bri2 = bri %>% generateSurvivalData(
#     admission_date,
#     discharge_date
# )
# 
# doKSTest = function(d,g,paramNames,params,n) {
#   sample = d$time
#   params = as.list(params)
#   names(params) = paramNames
#   params$x = jitter(sample)
#   params$y = "plnorm"
#   #browser()
#   ks = do.call("ks.test",params)
#   return(tibble(
#     testN = length(sample),
#     referenceN = max(n),
#     pValue = ks$p.value,
#     dStatistic = ks$statistic
#   ))
# }
# 
# gofBri = bri2 %>% filter(status == 1) %>% group_by(ageCat) %>% group_modify(function(d,g,...) {
#   reference = out %>% filter(model == "lnorm" & ageCat == g$ageCat)
#   out = bind_rows(
#     doKSTest(d,g,reference$param, reference$value, reference$n) %>% mutate(estimate = "central"),
#     doKSTest(d,g,reference$param, reference$low_ci, reference$n) %>% mutate(estimate = "low credible"),
#     doKSTest(d,g,reference$param, reference$high_ci, reference$n) %>% mutate(estimate = "high credible")
#   )
# })
# 
# numbers = gofBri %>% select(ageCat,estimate,test=testN,fit=referenceN) %>% pivot_longer(cols=c("test","fit"),names_to = "source", values_to = "N")
# 
# top = ggplot(gofBri, aes(x=ageCat,y=pValue,fill=estimate))+geom_bar(stat="identity", position="dodge", width=0.7, colour="black")+standardPrintOutput::narrowAndTall()+theme(axis.text.x = element_blank(), axis.title.x = element_blank())+ylab("P value")
# bottom = ggplot(numbers, aes(x=ageCat,y=source,label=N, colour = source))+geom_text()+guides(colour="none")+theme(panel.grid = element_blank(),axis.title.x = element_blank())+ylab("N")
# 
# (top+bottom+patchwork::guide_area()+patchwork::plot_layout(ncol=1, heights = c(9,2,1), guides = "collect")) %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/Fig5_validateBristol")
```




# fit all models

```{r}
# multistatesymptomaticToAdmission = CHESSClean %>% filter(age>15) %>% group_by(caseid, sex) %>% generateMultistateSurvivalData(
#   idVar = caseid,
#   transMatrix = transitionMatrix(
#     admitted=`ITU admission`,
#     admitted=`discharge death`
#     #`ITU admission`=`discharge death`
#   ),
#   ageBreaks = c(-Inf,50,70,80,Inf), 
#   ageLabels = c("10-50","50-70","70-80","80+"),
#   censoringDateExpr = censorDate,
#   startDateCol = "admitted",
#   admitted = hospitaladmissiondate, 
#   `ITU admission` = dateadmittedicu,
#   `discharge death` = if_else(is.na(dateadmittedicu),finaloutcomedate,as.Date(NA))
# )
```

```{r}
# 
# crweib <- flexsurv::flexsurvreg(Surv(time, status) ~ ageCat + trans + shape(trans), data = multistatesymptomaticToAdmission, dist = "weibull")
# crgamma <- flexsurv::flexsurvreg(Surv(time, status) ~ ageCat + trans + shape(trans), data = multistatesymptomaticToAdmission, dist = "gamma")
# # crnorm <- flexsurv::flexsurvreg(Surv(time, status) ~ ageCat + trans + sdlog(trans), data = multistatesymptomaticToAdmission, dist = "lnorm")
# crlnorm <- flexsurv::flexsurvreg(Surv(time, status) ~ ageCat + trans + sdlog(trans), data = multistatesymptomaticToAdmission, dist = "lnorm")
# 
# plot(crweib)
# plot(crgamma)
# plot(crlnorm)
# 
# # https://www.rdocumentation.org/packages/flexsurv/versions/1.1.1/topics/msfit.flexsurvreg
# 
# survcheck(Surv(time, status, type = "mstate") ~ sex + ageCat, multistatesymptomaticToAdmission, id=id)
# survModel = coxph(Surv(time, status, type = "mstate") ~ sex + ageCat, multistatesymptomaticToAdmission, id=id)
# survminer::ggforest(survModel)
# 
#crcox = coxph(Surv(time, status) ~ strata(trans), data = multistatesymptomaticToAdmission)
# out = censoredsymptomaticToAdmission %>% fitModelsByAge(c("lnorm"))
# 
# out %>% mutate(value = sprintf("%1.4f [%1.4f; %1.4f]", value, low_ci, high_ci)) %>% select(ageCat,n,model,aic,bic,loglik,param, value) %>% 
#   group_by(ageCat,n,model,aic,bic,loglik) %>% standardPrintOutput::saveTableLandscape("~/Dropbox/covid19/ventilator-demand/parameterisation/symptomatic-to-admission/SuppTable1_fullBreakdownFittedModels")
```
