---
title: "A spatiotemporal model of ventilator demand in the NHS due to COVID-19"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A spatiotemporal model of ventilator demand in the NHS due to COVID-19}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}

library(tidyverse)
# devtools::install_github("terminological/uk-covid-datatools")
# library(ukcovidtools)
devtools::load_all("~/Git/uk-covid-datatools/")
# devtools::install_github("terminological/standard-print-output")
# library(standardPrintOutput)
devtools::load_all("~/Git/standard-print-output/")
# devtools::install_github("terminological/tidy-info-stats")
# library(tidyinfostats)
devtools::load_all("~/Git/tidy-info-stats/")

library(rgdal)
library(ggplot2)
library(rgeos)
library(maptools)
library(lubridate)
library(ggspatial)
library(openxlsx)

library(ensurer)

ggplot2::theme_set(standardPrintOutput::defaultFigureLayout())

config = config::get(file = "~/Dropbox/ventilator_demand.yaml")

###################
# # ~/Dropbox/ventilator_demand.yaml example
# default:
#  modelFile: /home/terminological/Dropbox/covid19/ventilator-demand/RunsForRob/Rt=2.7/ForMattData.dat
#  dateOrigin: 2020-03-07
###################
```

Load and test the MetaWard epidemiological model


```{r}

ageCatToFactor = function(ageCat) {
  factor(
    ageCat,
    levels = c('0-4','5-9','10-14','15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64','65-69','70-74','75-79','80+'),
    ordered = TRUE
  )
}

ageToAgeCat = function(age) {
  return(cut(age,
    breaks = c(-Inf,5,10,15,20,25,30,35,40,45,50,44,60,65,70,75,80,Inf),
    labels = c('0-4','5-9','10-14','15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64','65-69','70-74','75-79','80+'),
    include.lowest = TRUE, ordered_result = TRUE
  ))
}

```


## Transform infected to symptomatic
N.B. This filters out a large fraction of the young, as they have mild symptoms.
It also introduces a lag because of the incubation period


```{r}
matrix = loadDefaultAgeTimeMatrices()
```

```{r}
prepareModelResult = function(scenario) {
  print(scenario$modelFile)
  #### metawards model output ----
  sim = readSpatialFile(scenario$modelFile, ignoreErrors=TRUE)
  # check various parameters correct:
  # TODO: use ensurer for this
  # length(unique(sim$ward))
  # length(unique(sim$time))
  # sim %>% group_by(time) %>% summarise(ukwideDaily = sum(value,na.rm=TRUE)) %>% pull(ukwideDaily) %>% max()
  # sim %>% group_by(time) %>% summarise(ukwideDaily = sum(value,na.rm=TRUE)) %>% pull(ukwideDaily) %>% max()
  # sim %>% ungroup() %>% summarise(ukwide = sum(value,na.rm=TRUE)) %>% pull(ukwide)
  # File is unique for ward and time point
  sim %>% group_by(time,ward) %>% count() %>% filter(n > 1) %>% ensure(nrow(.) == 0) %>% invisible()
  # Aggregate model to LAD level
  dateOrigin = as.Date(config$dateOrigin)
  simByLAD = sim %>% group_by(time,LAD19CD,LAD19NM) %>% 
    summarise(value = sum(value)) %>% 
    mutate(date = dateOrigin+time) %>% ungroup() %>% select(-time)
  aggDemographics = UKDemographics2019 %>% mutate(ageGroup = ageToAgeCat(age)) %>% group_by(ageGroup,code,name) %>% summarise(
    count = sum(count),
    total = first(total)
  )
  # check sums equal totals
  aggDemographics %>% group_by(code) %>% summarise(total2 = sum(count), total=first(total)) %>% 
    ensure(all.equal(.$total,.$total2)) %>% invisible()
  # Combine Sim and demographics
  simByLADandDemography = 
    simByLAD %>% 
    left_join(aggDemographics, by=c("LAD19CD"="code")) %>% 
    ensure_that(!all(is.na(.$ageGroup)))
  simByLADandDemography = simByLADandDemography %>% mutate(predictedInfected = value*count/total)
  # TODO: at this point we are breaking prediction down by proportion of the population in that age category
  # however we never check if the integral of the prediction is smaller than the population for each category.
  # If may be that the prediction is made on 2011 census numbers and actually we should scale this for up 2019 population estimates on a region by region basis
  # Aggregate to hospital catchment areas
  # OK data missing for some local authority districts. Probably in the calculation for ICU beds. Need to cvalidate / visualise these are working
  # sidestepped this using inner_join but problem remains #TODO: fix
  # LAD_unmatched = simByLADandDemography %>% anti_join(NHSCatchmentAreas$icuBeds, by=c("LAD19CD"="ladcd")) %>% pull(LAD19CD) %>% unique()
  # Lets push on regardless
  # TODO: need to have paralell data processing stream here involing NHSCatchmentAreas for hospital beds
  simByTrustAndDemography = simByLADandDemography %>% inner_join(NHSCatchmentAreas$icuBeds, by=c("LAD19CD"="ladcd")) %>%
    ensure_that(!any(is.na(.$trustId)))
  # aggregate to trust level with fractionOfLADpopulation multiplier
  simByTrustAndDemography = simByTrustAndDemography %>% group_by(trustId, ageGroup, date) %>% summarise(
    predictedInfected = sum(predictedInfected*fractionOfLADPopulation),
    ageSpecificPopulation = sum(count*fractionOfLADPopulation)
  ) %>% left_join(NHSCapacity2019$trusts, by="trustId")
  
  infected = simByTrustAndDemography %>% rename(infectedDate = date) %>% ungroup() %>%
    mutate(ageGroup = as.character(ageGroup))
  return(infected)
}

# TODO: replicate tests from above

extractDemographics = function(modelResult, trustList=c("RBA","RBZ","RH8","RRK")) {
  tmp = modelResult %>% filter(trustId %in% trustList) %>% 
    group_by(trustId,trustName,ageGroup) %>% 
    summarise(population = first(ageSpecificPopulation))  
  tmp = tmp %>% mutate(ageGroup = ageCatToFactor(ageGroup))
  return(tmp)
}

processModelResult = function(infected, trustList=c("RBA","RBZ","RH8","RRK")) {
  infected = infected %>% filter(trustId %in% trustList) %>% group_by(trustId, trustName) #,ageSpecificPopulation,regionId,trustName,acuteBeds,dayBeds,icuBeds,approxLat,approxLong)
  simEnd = max(infected$infectedDate)
  # Transform infected to symptomatic
  # N.B. This filters out a large fraction of the young, as they have mild symptoms.
  # It also introduces a lag because of the incubation period
  
  # TODO: probably could combine all the matrices first and do the larger processing in one step.
  # This woudl be quicker and could be made configurable
  # So we can implement the the model for different catchment areas. (hospital versus ITU)
  
  symptomatic = infected %>% applyAgeTimeMatrix(
    matrix$infected2Symptomatic, 
    predictedInfected, predictedSymptomatic,
    infectedDate, symptomaticDate
  )
  
  # Transform symptomatic to hospitalised
  
  hospitalised = symptomatic %>% applyAgeTimeMatrix(
    matrix$symptomaticToHospitalisation, 
    predictedSymptomatic, predictedHospitalised,
    symptomaticDate, hospitalisedDate
  )
  
 
  inpatient = hospitalised %>%  applyAgeTimeMatrix(
    matrix$hospitalisedToInpatient, 
    predictedHospitalised, predictedInpatient,
    hospitalisedDate, inpatientDate
  )
  
  # Transform symptomatic to itu admission
  
  ituAdmit = symptomatic %>% applyAgeTimeMatrix(
    matrix$symptomaticToItu, 
    predictedSymptomatic, predictedAdmitItu,
    symptomaticDate, admitItuDate
  )
  
  
  ituInpatient = ituAdmit %>% applyAgeTimeMatrix(
    matrix$ituAdmitToItuInpatient, 
    predictedAdmitItu, predictedItuInpatient,
    admitItuDate, ituInpatientDate
  )
  
  
  predictions = infected %>% ungroup() %>%
      select(trustId,trustName,ageGroup,date=infectedDate,newInfections = predictedInfected) %>%
      left_join(
        symptomatic %>%  ungroup() %>%
                  select(trustId,ageGroup,date=symptomaticDate,newSymptomatic = predictedSymptomatic), 
                  by=c("trustId","ageGroup","date")) %>%
      left_join(
        hospitalised %>%  ungroup() %>%
                  select(trustId,ageGroup,date=hospitalisedDate,newHospitalised = predictedHospitalised), 
                  by=c("trustId","ageGroup","date")) %>%
      left_join(
        inpatient %>%  ungroup() %>%
                  select(trustId,ageGroup,date=inpatientDate,inpatient = predictedInpatient), 
                  by=c("trustId","ageGroup","date")) %>%
      left_join(
        ituAdmit %>%  ungroup() %>%
                  select(trustId,ageGroup,date=admitItuDate,newItuAdmission = predictedAdmitItu), 
                  by=c("trustId","ageGroup","date")) %>%
      left_join(
        ituInpatient %>%  ungroup() %>%
                  select(trustId,ageGroup,date=ituInpatientDate,ituInpatient = predictedItuInpatient), 
                  by=c("trustId","ageGroup","date"))
  return(predictions)
}
```

```{r}
#TODO: make the scenario processing steps one function
#scenario = config$scenarios$partialControl
# modelResult = scenario %>% prepareModelResult()
trustList = c("RBA","RBZ","RH8","RRK")
scenarioResult = lapply(config$scenarios, function(s) {
  s$modelResult = s %>% prepareModelResult()
  return(s)
})

```



```{r}
#TODO: create one function here - save result in a Rdata file
#TODO: need to extend this with demographics for hospital admissions and demographics for ITU admissions
demographics = scenarioResult[[1]]$modelResult %>% extractDemographics(trustList)

scenarioResult = lapply(scenarioResult, function(s) {
  s$predictions = s$modelResult %>% processModelResult(trustList)
  s$summaryPredictions = s$predictions %>% group_by(trustName, trustId, date) %>%
    summarise(
      newInfections = sum(newInfections),
      newSymptomatic = sum(newSymptomatic),
      newHospitalised = sum(newHospitalised),
      newItuAdmission = sum(newItuAdmission),
      inpatient = sum(inpatient),
      ituInpatient = sum(ituInpatient)
    ) %>% filter(date < as.Date("2020-09-01"))
  return(s)
})

# Create a blank workbook
for (trust in trustList) {
  
  excelOutput <- openxlsx::createWorkbook()
  # Add some sheets to the workbook
  openxlsx::addWorksheet(excelOutput, "Demographics")
  openxlsx::writeData(excelOutput, sheet = "Demographics", x = demographics %>% 
                        filter(trustId == trust) %>% arrange(ageGroup))
  
  for (i in 1:length(scenarioResult)) {
    scenarioName = names(scenarioResult)[[i]]
    tmpScenario = scenarioResult[[i]]
    summaryTab = paste0("Summary ",tmpScenario$R0) #scenarioName," ",tmpScenario$R0)
    print(summaryTab)
    openxlsx::addWorksheet(excelOutput, summaryTab)
    openxlsx::writeData(excelOutput, sheet = summaryTab, x = tmpScenario$summaryPredictions %>% filter(trustId == trust))
  
    breakdownTab = paste0("Age breakdown ",tmpScenario$R0) #,scenarioName," ",tmpScenario$R0)
    print(breakdownTab)
    openxlsx::addWorksheet(excelOutput, breakdownTab)
    openxlsx::writeData(excelOutput, sheet = breakdownTab, x = tmpScenario$predictions %>% filter(trustId == trust))
  }
  # Export the file
  openxlsx::saveWorkbook(excelOutput, paste0("~/Dropbox/covid19/ventilator-demand/byTrust/",trust,".xlsx"), overwrite = TRUE)
}
```

```{r}

```

```{r}
#### Graveyard ----
# 
# agestats=data.frame(
#   agegroup=c('0-4','5-9','10-14','15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64','65-69','70-74','75-79','80+'),
#   #pyramid=c(0.058,0.063,0.058,0.055,0.063,0.068,0.068,0.066,0.061,0.068,0.070,0.064,0.054,0.050,0.049,0.033,0.049), # this must be demographic breakdown
#   IFR=c(0.0,0.0,0.0,0.0001,0.0002,0.0004,0.0006,0.001,0.001,0.002,0.005,0.008,0.017,0.027,0.043,0.062,0.096),
#   Hosp=c(0.001,0.001,0.001,0.002,0.005,0.01,0.016,0.023,0.029,0.039,0.058,0.072,0.102,0.117,0.146,0.177,0.18),
#   HFR=c(0.038,0.038,0.038,0.038,0.038,0.038,0.038,0.04,0.045,0.056,0.078,0.113,0.169,0.232,0.291,0.348,0.535)
# )
# 
# 
# 
# #### ICU beds by NHS trust (maybe spread between hospitals) ----
# 
# glimpse(NHSCatchmentAreas$acuteBeds)
# glimpse(NHSCatchmentAreas$icuBeds)
# 
# # icu provision per LAD:
# NHSCatchmentAreas$icuBeds %>% left_join(NHSCapacity2019$trusts, by="trustId")
# 
# icuTrusts = NHSCapacity2019$trusts %>% filter(icuBeds>0)
# 
# icuToIcuNetwork = icuTrusts %>% mutate(sourceTrustId = trustId) %>% 
#   tidyinfostats::findKNN(icuTrusts %>% mutate(targetTrustId = trustId), sourceTrustId, targetTrustId, matchVars = vars(approxLat,approxLong), k=5)
# 
# closestTrustToLAD = LADlocation %>% tidyinfostats::findKNN(icuBedsByTrustLocation %>% ungroup() %>% 
#   mutate(lat_mean = Latitude, long_mean = Longitude), ladcd, trustId, matchVars = vars(lat_mean,long_mean), k=2)
# 
# write_csv(icuTrusts, "icuBedsByTrustLocation.csv")
# write_csv(icuToIcuNetwork, "icuToIcuNetwork.csv")
# #write_csv(closestTrustToLAD, "closestIcuToLAD.csv")
# 
# #### Integrate demographics ----
# # 
# tmp = LADlocation %>% inner_join(UKDemographics2019, by=c("ladcd"="code"))
# # 
# 
# # summarise all postcodes in a LAD location to lat long
# ```
# 
# ```{r}
# library(tidyverse)
# library(gghighlight)
# library(purrr)
# library(plotly)
# library(cowplot)
# library(scales)
# library(lubridate)
# library(kableExtra)
# source('~/GitHub/MetaWards/processfunctions.R')
# pathname='~/GitHub/MetaWards/Runs/24_03/'
#   dir(pathname,pattern='PlayInfections.dat',recursive=TRUE)->pfilenames
#   dir(pathname,pattern='WorkInfections.dat',recursive=TRUE)->wfilenames
#   #PLAY infecteds
#   dataIP <- tibble(filename = pfilenames) %>% # create a data frame
#   # holding the file names
#     mutate(file_contents = map(filename,            # read files into a new data column
#                              ~readIncidenceRecovered(., pathname))) %>% # use the function written above to process files
#     mutate(Type=str_split(filename,'/',simplify = TRUE)[,1]) %>% 
# #    mutate(Run=str_split(filename,'/',simplify = TRUE)[,2]) %>% 
#     select(-filename) %>% 
#     unnest(cols=file_contents)
# # WORK infected
#   dataIW <- tibble(filename = wfilenames) %>% # create a data frame
#   # holding the file names
#     mutate(file_contents = map(filename,            # read files into a new data column
#                                ~readIncidenceRecovered(., pathname))) %>% # use the function written above to process files
#     mutate(Type=str_split(filename,'/',simplify = TRUE)[,1]) %>% 
# #    mutate(Run=str_split(filename,'/',simplify = TRUE)[,2]) %>% 
#     select(-filename) %>% 
#     unnest(cols=file_contents)
#   bind_cols(dataIW,dataIP) %>% 
#     mutate(Incidence=Incidence+Incidence1) %>% 
#     mutate(Recovered=Recovered+Recovered1) %>% 
# #    mutate(Prevalence = Prevalence + Prevalence1) %>% 
#     select(Time,Incidence,Recovered,Type)->dataIR
# dataIR %>%
# #  mutate(Date=as.Date(Time,origin='2020-01-13')) %>% 
#   mutate(Date=as.Date(Time,origin='2020-01-20')) %>% 
#   #mutate(Incidence=Incidence/10^6) %>% 
#   #mutate(Recovered=Recovered/10^6) %>% 
#   filter(Time<550) %>% 
#   ggplot(aes(x=Date,group=Type,colour=Type))+
#   geom_line(aes(y=Incidence*0.05),alpha=0.8,size=1)+
# #  geom_line(aes(y=Recovered),alpha=0.1)+
# #  facet_wrap(~Location)+
#   theme_minimal_grid()+
#  theme(legend.position = "top",legend.title = element_blank())+
#  #scale_y_continuous(labels = unit_format(unit = "M"))+
#   xlab('Time')+
#   scale_x_date(breaks=date_breaks("3 months"),
#                labels = date_format("%b"))+
#   ylab('Daily Hospitalisations')
# dataIR %>% 
#   group_by(Type) %>%
#   summarise(sum(Incidence))
# ```
# ```{r}
# N=56082077
# 
# agestats=data.frame(
#   agegroup=c('0-4','5-9','10-14','15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64','65-69','70-74','75-79','80+'),
#   pyramid=c(0.058,0.063,0.058,0.055,0.063,0.068,0.068,0.066,0.061,0.068,0.070,0.064,0.054,0.050,0.049,0.033,0.049),
#   IFR=c(0.0,0.0,0.0,0.0001,0.0002,0.0004,0.0006,0.001,0.001,0.002,0.005,0.008,0.017,0.027,0.043,0.062,0.096),
#   Hosp=c(0.001,0.001,0.001,0.002,0.005,0.01,0.016,0.023,0.029,0.039,0.058,0.072,0.102,0.117,0.146,0.177,0.18),
#   HFR=c(0.038,0.038,0.038,0.038,0.038,0.038,0.038,0.04,0.045,0.056,0.078,0.113,0.169,0.232,0.291,0.348,0.535)
# )
# 
# dataIR %>%
# #  mutate(Date=as.Date(Time,origin='2020-01-13')) %>% 
#   mutate(Date=as.Date(Time,origin='2020-01-20')) %>% 
#   mutate(Week = week(Date),Year = year(Date)) %>% 
#   group_by(Week,Year,Type) %>% 
#   summarise(NewCases=sum(Incidence),
#             Hospitalisations=round(sum(NewCases*agestats$pyramid*agestats$Hosp)),
#             Deaths=round(sum(NewCases*agestats$pyramid*agestats$IFR))) %>% 
#   arrange(Type,Year,Week) %>% 
#   group_by(Type) %>% 
#   summarise(`Attack Rate (%)`=round(100*sum(NewCases)/N),`Total Hospital` = sum(Hospitalisations),`Total Deaths` = sum(Deaths)) %>% 
#   kable
#   
# ```
# ```{r}
# dataIR %>%
# #  mutate(Date=as.Date(Time,origin='2020-01-13')) %>% 
#   mutate(Date=as.Date(Time,origin='2020-01-20')) %>% 
#   mutate(Week = week(Date),Year = year(Date)) %>% 
#   group_by(Week,Year,Type) %>% 
#   summarise(NewCases=sum(Incidence),
#             Hospitalisations=round(sum(NewCases*agestats$pyramid*agestats$Hosp)),
#             Deaths=round(sum(NewCases*agestats$pyramid*agestats$IFR))) %>% 
#   arrange(Type,Year,Week) %>% 
#   group_by(Type,Year) %>% 
#   mutate(Hospitalisations = lag(Hospitalisations,1)) %>% 
#   mutate(Deaths = lag(Deaths,3)) %>% 
#   write_csv('~/GitHub/MetaWards/Runs/24_03/EBWModelProjections.csv')
# dataIR %>%
# #  mutate(Date=as.Date(Time,origin='2020-01-13')) %>% 
#   mutate(Date=as.Date(Time,origin='2020-01-20')) %>% 
#   mutate(Week = week(Date),Year = year(Date)) %>% 
#   group_by(Week,Year,Type) %>% 
#   summarise(NewCases=sum(Incidence),
#             Hospitalisations=round(sum(NewCases*agestats$pyramid*agestats$Hosp)),
#             Deaths=round(sum(NewCases*agestats$pyramid*agestats$IFR))) %>% 
#   arrange(Type,Year,Week) %>% 
#   group_by(Type,Year) %>% 
#   mutate(Hospitalisations = lag(Hospitalisations,1)) %>% 
#   mutate(Deaths = lag(Deaths,3))
#   
# dataIR %>%
# #  mutate(Date=as.Date(Time,origin='2020-01-13')) %>% 
#   mutate(Date=as.Date(Time,origin='2020-01-20')) %>% 
#   mutate(Week = week(Date),Year = year(Date)) %>% 
#   group_by(Week,Year,Type) %>% 
#   summarise(NewCases=sum(Incidence),
#             Hospitalisations=round(sum(NewCases*agestats$pyramid*agestats$Hosp)),
#             Deaths=round(sum(NewCases*agestats$pyramid*agestats$IFR))) %>% 
#   group_by(Type) %>% 
#   summarise(max(Hospitalisations))
#   
# dataIR %>%
# #  mutate(Date=as.Date(Time,origin='2020-01-13')) %>% 
#   mutate(Date=as.Date(Time,origin='2020-01-20')) %>% 
#   mutate(Week = week(Date),Year = year(Date)) %>% 
#   group_by(Week,Year,Type) %>% 
#   summarise(NewCases=sum(Incidence),
#             Hospitalisations=round(sum(NewCases*agestats$pyramid*agestats$Hosp)),
#             Deaths=round(sum(NewCases*agestats$pyramid*agestats$IFR))) %>% 
#   arrange(Type,Year,Week) %>% 
#   ggplot(aes(x=Week,group=Type,colour=Type))+
#     geom_line(aes(y=Hospitalisations),alpha=0.8,size=1)+
# #  geom_line(aes(y=Recovered),alpha=0.1)+
# #  facet_wrap(~Location)+
#     facet_wrap(~Year)+
#     theme_minimal_grid()+
#     theme(legend.position = "top",legend.title = element_blank())+
#  #scale_y_continuous(labels = unit_format(unit = "M"))+
#     xlab('Week')+
#     ylab('Weekly Hospitalisations')
```

```{r eval = FALSE, echo = FALSE}
# print(config$scenarios$uncontrolled$modelFile)
# #### metawards model output ----
# sim = readSpatialFile(config$scenarios$uncontrolled$modelFile, ignoreErrors=TRUE)
# 
# # glimpse(UKCovidMaps$ward)
# # unique(sim$time)
# tmpSim = UKCovidMaps$ward %>% left_join(sim %>% filter(time==100), by=c("wd11cd" = "WD11CD"))
# 
# ggplot(tmpSim)+geom_sf(aes(fill=value),lwd = 0)+scale_fill_gradient(low="white",high="red",trans="sqrt",na.value = "grey90")+standardPrintOutput::mapTheme()
# 
# # check various parameters correct:
# # TODO: use ensurer for this
# # length(unique(sim$ward))
# # length(unique(sim$time))
# # sim %>% group_by(time) %>% summarise(ukwideDaily = sum(value,na.rm=TRUE)) %>% pull(ukwideDaily) %>% max()
# # sim %>% group_by(time) %>% summarise(ukwideDaily = sum(value,na.rm=TRUE)) %>% pull(ukwideDaily) %>% max()
# # sim %>% ungroup() %>% summarise(ukwide = sum(value,na.rm=TRUE)) %>% pull(ukwide)
# # File is unique for ward and time point
# sim %>% group_by(time,ward) %>% count() %>% filter(n > 1) %>% ensure(nrow(.) == 0) %>% invisible()
# 
# ggplot(sim %>% group_by(time) %>% summarise(ukwide = sum(value,na.rm=TRUE)), aes(x=time,y=ukwide))+geom_line()


```
# Aggregate model to LAD level

```{r eval = FALSE, echo = FALSE}
# dateOrigin = as.Date(config$dateOrigin)
# simByLAD = sim %>% group_by(time,LAD19CD,LAD19NM) %>% summarise(value = sum(value)) %>% mutate(date = dateOrigin+time) %>% ungroup() %>% select(-time)

```
# Combine Sim and demographics

```{r eval = FALSE, echo = FALSE}
# 
# # simByLAD %>% anti_join(aggDemographics, by=c("LAD19CD"="code")) %>% select(LAD11CD) %>% distinct()
# 
# simByLADandDemography = 
#   simByLAD %>% 
#   left_join(aggDemographics, by=c("LAD19CD"="code")) %>% 
#   ensure_that(!all(is.na(.$ageGroup)))
# 
# simByLADandDemography = simByLADandDemography %>% mutate(predictedInfected = value*count/total)
# # TODO: at this point we are breaking prediction down by proportion of the population in that age category
# # however we never check if the integral of the prediction is smaller than the population for each category.
# # If may be that the prediction is made on 2011 census numbers and actually we should scale this for up 2019 population estimates on a region by region basis

```

# Aggregate to hospital catchment areas

```{r eval = FALSE, echo = FALSE}
# #### get data about beds etc ----
# 
# # glimpse(NHSCapacity2019$hospitals)
# # glimpse(NHSCapacity2019$trusts) 
# # glimpse(NHSCatchmentAreas$icuBeds)
# 
# # OK data missing for some local authority districts. Probably in the calculation for ICU beds. Need to cvalidate / visualise these are working
# # sidestepped this using inner_join but problem remains #TODO: fix
# # LAD_unmatched = simByLADandDemography %>% anti_join(NHSCatchmentAreas$icuBeds, by=c("LAD19CD"="ladcd")) #%>% pull(LAD19CD) %>% unique()
# # Lets push on regardless
# simByTrustAndDemography = simByLADandDemography %>% inner_join(NHSCatchmentAreas$icuBeds, by=c("LAD19CD"="ladcd")) %>% ensure_that(!any(is.na(.$trustId)))
# 
# # aggregate to trust level with fractionOfLADpopulation multiplier
# simByTrustAndDemography = simByTrustAndDemography %>% group_by(trustId, ageGroup, date) %>% summarise(
#   predictedInfected = sum(predictedInfected*fractionOfLADPopulation),
#   ageSpecificPopulation = sum(count*fractionOfLADPopulation)
# ) %>% left_join(NHSCapacity2019$trusts, by="trustId")
# 
# # set of curves of same shape with height reflecting deographics of area
# ggplot(simByTrustAndDemography %>% filter(trustId=="R0A"), aes(x=date, y=predictedInfected, colour=ageGroup)) +geom_line()
# 
# # check population size is static and appropriate over the model run
# # simByTrustAndDemography %>% group_by(date) %>% summarise(total = sum(ageSpecificPopulation)) %>% pull(total)
# # this is  very close to ENGLAND AND WALES 59115809
# # which is reassuring. Mismatch could be due to some patients being assigned to more than one trust...?
# # NHSCatchmentAreas$icuBeds %>% left_join(NHSCapacity2019$trusts, by="trustId")

```

# Assumptions

```{r eval = FALSE, echo = FALSE}
# 
# infected = simByTrustAndDemography %>% rename(infectedDate = date) %>% ungroup() %>% mutate(ageGroup = as.character(ageGroup))
# 
# simEnd = max(infected$infectedDate)
```
# Demographics by LAD

```{r eval = FALSE, echo = FALSE}
# 
# aggDemographics = UKDemographics2019 %>% mutate(ageGroup = cut(age,
#       breaks = c(-Inf,5,10,15,20,25,30,35,40,45,50,44,60,65,70,75,80,Inf),
#       labels = c('0-4','5-9','10-14','15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64','65-69','70-74','75-79','80+'),
#       include.lowest = TRUE, ordered_result = TRUE
# )) %>% group_by(ageGroup,code,name) %>% summarise(
#   count = sum(count),
#   total = first(total)
# )
# 
# 
# aggDemographics %>% group_by(code) %>% summarise(total2 = sum(count), total=first(total)) %>% 
#   ensure(all.equal(.$total,.$total2)) %>% invisible()

```{r}

# demographics = scenarioResult[[1]]$modelResult %>% extractDemographics(trustList)
# ggplot(demographics,aes(x=ageGroup,y=population))+geom_bar(stat="identity")+facet_wrap(vars(trustName))+narrowAndTall()
```

```{r}
# predictions = modelResult %>% processModelResult(trustList)
# summaryPredictions = predictions %>% group_by(trustName, trustId, date) %>%
#   summarise(
#     newInfections = sum(newInfections),
#     newSymptomatic = sum(newSymptomatic),
#     newHospitalised = sum(newHospitalised),
#     newItuAdmission = sum(newItuAdmission),
#     inpatient = sum(inpatient),
#     ituInpatient = sum(ituInpatient)
#   ) %>% filter(date < as.Date("2020-09-01"))
# 
# infectionsPlot = ggplot(summaryPredictions,aes(x=date,y=newInfections,colour=trustName))+geom_line()+scale_x_date(date_breaks = "1 month")
# symptomaticPlot = ggplot(summaryPredictions,aes(x=date,y=newSymptomatic,colour=trustName))+geom_line()+scale_x_date(date_breaks = "1 month")
# hospitalAdmissionsPlot = ggplot(summaryPredictions,aes(x=date,y=newHospitalised,colour=trustName))+geom_line()+scale_x_date(date_breaks = "1 month")
# ituAdmissionsPlot = ggplot(summaryPredictions,aes(x=date,y=newItuAdmission,colour=trustName))+geom_line()+scale_x_date(date_breaks = "1 month")
# hospitalInpatientsPlot = ggplot(summaryPredictions,aes(x=date,y=inpatient,colour=trustName))+geom_line()+scale_x_date(date_breaks = "1 month")
# ituInpatientsPlot = ggplot(summaryPredictions,aes(x=date,y=ituInpatient,colour=trustName))+geom_line()+scale_x_date(date_breaks = "1 month")
# 
# ((infectionsPlot | symptomaticPlot) / (hospitalAdmissionsPlot | ituAdmissionsPlot) / (hospitalInpatientsPlot | ituInpatientsPlot) / patchwork::guide_area() + patchwork::plot_layout(guides="collect")) %>% standardPrintOutput::saveFullPageFigure("~/Dropbox/covid19/ventilator-demand/Scenario2")

```


```{r eval = FALSE, echo = FALSE}
# infected2SymptomaticUrl = "https://docs.google.com/spreadsheets/d/e/2PACX-1vSl_h_Vjag3N1bkfBlqCotUmB_Mg5BPfugGzhsmK-ByycxKGvODIGBD-IRtEwjOclu4UUchiZ3j45xU/pub?gid=2044085188&single=true&output=csv"
# 
# #loadAgeTimeMatrix(infected2SymptomaticUrl)
# 
# symptomatic = infected %>% applyAgeTimeMatrix(
#   loadAgeTimeMatrix(infected2SymptomaticUrl), 
#   predictedInfected, predictedSymptomatic,
#   infectedDate, symptomaticDate
# )
# 
# ggplot(symptomatic %>% filter(trustId=="R0A"), aes(x=symptomaticDate, y=predictedSymptomatic, colour=ageGroup))+geom_line()

```

```{r eval = FALSE, echo = FALSE}
# symptomaticToHospitalisationUrl = "https://docs.google.com/spreadsheets/d/e/2PACX-1vSl_h_Vjag3N1bkfBlqCotUmB_Mg5BPfugGzhsmK-ByycxKGvODIGBD-IRtEwjOclu4UUchiZ3j45xU/pub?gid=1708683854&single=true&output=csv"
# 
# hospitalised = symptomatic %>% applyAgeTimeMatrix(
#   loadAgeTimeMatrix(symptomaticToHospitalisationUrl), 
#   predictedSymptomatic, predictedHospitalised,
#   symptomaticDate, hospitalisedDate
# )
# 
# ggplot(hospitalised %>% filter(trustId=="R0A"), 
#        aes(x=hospitalisedDate, y=predictedHospitalised, colour=ageGroup))+geom_line()
```

```{r eval = FALSE, echo = FALSE}
# symptomaticToItuUrl =
# "https://docs.google.com/spreadsheets/d/e/2PACX-1vSl_h_Vjag3N1bkfBlqCotUmB_Mg5BPfugGzhsmK-ByycxKGvODIGBD-IRtEwjOclu4UUchiZ3j45xU/pub?gid=895816136&single=true&output=csv"
# 
# itu = symptomatic %>% applyAgeTimeMatrix(
#   loadAgeTimeMatrix(symptomaticToItuUrl), 
#   predictedSymptomatic, predictedAdmitItu,
#   symptomaticDate, admitItuDate
# )
# 
# ggplot(itu %>% mutate(ageGroup = ageCatToFactor(ageGroup)) %>% filter(trustId=="RBZ"), 
#   aes(x=admitItuDate, y=ageGroup, fill=predictedAdmitItu))+geom_tile()+
#   scale_fill_gradient(low="white",high="red",trans="sqrt")
```