---
title: "length of stay in hospital icu parameterisation"
author: "Rob Challen"
date: '`r format(Sys.Date(), "%d-%m-%Y")`'
output: 
  pdf_document :
    fig_caption: yes
header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}
knit: (function(inputFile, encoding,...) {
  rmarkdown::render(inputFile, encoding = encoding, output_dir = "~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge" )})
fig_width: 7
fig_height: 5
out.width: "100%"
bibliography: current-rt.bib
csl: current-rt.csl
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lubridate)
devtools::load_all("~/Git/standard-print-output/")
library(survival)

devtools::load_all("~/Git/standard-print-output/")
library(ggplot2)
library(patchwork)

standardPrintOutput::setDefaults()

cap = list(
  fig = captioner::captioner(prefix="Figure"),
  tab = captioner::captioner(prefix="Table"),
  sfig = captioner::captioner(prefix="Supplemental figure"),
  stab = captioner::captioner(prefix="Supplemental table")
)

ref = list(
  fig = function(name) cap$fig(name,display="cite"),
  tab = function(name) cap$tab(name,display="cite"),
  sfig = function(name) cap$sfig(name,display="cite"),
  stab = function(name) cap$stab(name,display="cite")
)

dpc = tsp = chp = srv = mwp = NULL

reload = function() {
  devtools::load_all("~/Git/uk-covid-datatools/")
  dpc <<- DataProviderController$setup("~/Data/maps/", "~/S3/encrypted/")
  tsp <<- dpc$timeseriesProcessor()
  # tsp$printSerialInterval()
  chp <<- dpc$chessProcessor()
  srv <<- dpc$survivalProcessor()
  mwp <<- dpc$metawardProcessor()
}

reload()
```

Robert Challen^1,2^; Krasimira Tsaneva-Atanasova^1,3^; Leon Danon^3,4^

1)  EPSRC Centre for Predictive Modelling in Healthcare, University of Exeter, Exeter, Devon, UK.
2)  Taunton and Somerset NHS Foundation Trust, Taunton, Somerset, UK.
3)  The Alan Turing Institute, British Library, 96 Euston Rd, London NW1 2DB, UK.
4)  Data Science Institute, College of Engineering, Mathematics and Physical Sciences, University of Exeter, Exeter, UK.

# Abstract


```{r}

CHESS = dpc$spim$getSARI()
cq = CHESS %>% chp$chessQuality()
# ggplot(cq, aes(x=trustcode, y=records))+geom_bar(stat="identity")+standardPrintOutput::narrowAndTall()
p = ggplot(cq, aes(x=knownOutcomePercent, y=knownAdmittedItuPercent, label=trustcode, colour=updatedRecords/records))+geom_point()+geom_vline(xintercept=0.1)+geom_hline(yintercept=0.5)+scale_color_gradient(low="red",high="green")+
  standardPrintOutput::narrowAndTall()
p2 = ggExtra::ggMarginal(p, type="histogram")
p2 %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/chessDataQuality")
```

```{r}
CHESSClean = CHESS %>% chp$chessItuSubset(updatedWithin = 60) %>% 
  mutate(
    withDischdate = !is.na(dateleavingicu), 
    withOutcomedate = !is.na(finaloutcomedate) )%>%
  mutate(patientstillonicu = case_when(
    finaloutcome == "Transferred" ~ "Unknown",
    withDischdate ~ "No",
    withOutcomedate ~ "No",
    finaloutcome %in% c("Death","Discharged") ~ "No",
    TRUE ~ "Unknown"
  ))

```

# Survival model - Admission to outcome

* Excludes hospitals which:
* A) haven;t updated data from > 3 days
* B) Have censored outcomes mode than one sd from the mean.
* N.b. This is over half of the hospitals.

* furthermore some other data quality issues e.g. outcome date before admission date
* multiple admissions: just use first admission

```{r}
ituAdmissionToDischargeOrDeath = CHESSClean %>% filter(age>15) %>% 
  mutate(
    startDate = dateadmittedicu, 
    endDate = pmin(dateleavingicu,finaloutcomedate,na.rm = TRUE)
  )  %>%
  srv$generateSurvivalData(
  idVar = caseid,
  startDateVar = startDate, 
  endDateExpr = endDate, 
  censoredDateExpr = pmin(dateupdated,censorDate,na.rm = TRUE), 
  statusExpr = if_else(patientstillonicu=="No",1,0),
  ageBreaks = c(-Inf,35,45,55,65,75,Inf),
  ageLabels = c("<35","35-45","45-55","55-65","65-75","75+"),
  ageReferenceCat = "<35",
  #ageBreaks = c(-Inf,30,50,70,Inf),
  #ageLabels = c("10-30","30-50","50-70","70+"),
  #ageReferenceCat = "30-50",
  statusLabels = c("unknown","discharged/died")
) %>%
  filter(!((status=="unknown") & is.na(dateupdated))) %>% 
  filter(time>=0) %>%
  select(
  caseid, status, trustcode, finaloutcome, time, ageCat, sex, startDate, endDate, withDischdate, withOutcomedate, dateupdated
)
```

## summarise data

```{r}
summary = bind_rows(
  ituAdmissionToDischargeOrDeath %>% group_by(value = finaloutcome) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(ituAdmissionToDischargeOrDeath)*100)) %>% mutate(category="Outcome", value = if_else(is.na(value),"Inpatient",as.character(value))),
  ituAdmissionToDischargeOrDeath %>% group_by(value = sex) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(ituAdmissionToDischargeOrDeath)*100)) %>% mutate(category="Gender", value = as.character(value)),
  ituAdmissionToDischargeOrDeath %>% group_by(value = ageCat) %>% summarise(N=n(), `%age`=sprintf("%1.1f %%",n()/nrow(ituAdmissionToDischargeOrDeath)*100)) %>% mutate(category="Age", value = as.character(value)) %>% arrange(value),
  ituAdmissionToDischargeOrDeath %>% summarise(N=n()) %>% mutate(category="Total", value = "", `%age`="" )
)

summary %>% group_by(category) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/Table1_summary")
```


## fit a survival model

```{r}
tmp = ituAdmissionToDischargeOrDeath %>% rename(age = ageCat) %>% mutate(status = ifelse(status=="unknown",0,1))
```


```{r}
fitByAgeCat = survival::survfit(survival::Surv(time,status) ~ age, tmp)
```


```{r}
plot = survminer::ggsurvplot(
  fitByAgeCat, data=tmp, conf.int = TRUE
)+ylab("P(itu patient)")
```


```{r}
(plot$plot+standardPrintOutput::defaultFigureLayout()) %>% 
  standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/Fig1_kaplanMeierByAge")

# summary(admissionToDischargeOrDeath)
```


```{r}
survModel = survival::coxph(survival::Surv(time,status) ~ sex + age, tmp)
# survModel %>% gtsummary::tbl_regression(exp = TRUE) 
```


```{r}
survminer::ggforest(survModel,data = tmp) %>% 
  standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/Fig2_coxModel")
```

Interpretation is that a high HR => discharge or death is more likely - i.e. an inpatient admission is likely to be shorter.

# Distribution fitting

Uses different age bands

```{r}
censoredItuAdmissionToDischargeOrDeath = CHESSClean %>% filter(age>15) %>% filter(!is.na(finaloutcome)) %>% 
  mutate(
    ageCat = tsp$cutByAge(age, ageBreaks = seq(15,85,5)),
    ageCatMid = tsp$cutByAge(age, ageBreaks = seq(15,85,10)),
    ageCatBig = tsp$cutByAge(age, ageBreaks = seq(15,75,20))
  ) %>% #tsp$ageToAgeCat(age)) %>%
  mutate(
    startDate = dateadmittedicu, 
    endDate = pmin(dateleavingicu,finaloutcomedate,na.rm = TRUE),
    censorDate = pmin(dateleavingicu,finaloutcomedate,dateupdated,censorDate,na.rm = TRUE)
  ) %>% 
  mutate(
    era = ifelse(startDate < "2020-06-16", "pre-dexa", "post-dexa"),
    lowerTime = if_else(patientstillonicu=="No",as.numeric(endDate-startDate)+0.01,as.numeric(censorDate-startDate))+0.01,
    upperTime = if_else(patientstillonicu=="No",as.numeric(endDate-startDate)+1.01,NA_real_)
  ) %>% 
  filter(lowerTime>0) %>%
  select(ageCat,ageCatMid,ageCatBig,era,startDate,endDate,censorDate,lowerTime,upperTime,patientstillonicu,finaloutcome)
  

```


```{r}

fit = DistributionFit$new(distributions = c("lnorm","gamma","weibull","nbinom"))

fitData = censoredItuAdmissionToDischargeOrDeath %>% group_by(ageCat) 
fit$fromUncensoredData(fitData%>% filter(patientstillonicu=="No"), valueExpr = (lowerTime+upperTime)/2,bootstraps = 100)
# fit$fromCensoredData(fitData%>% filter(patientstillonicu=="No"), lowerValueExpr = lowerTime, upperValueExpr = upperTime ,bootstraps = 100)

fit$plot(xlim=c(0,60)) %>% 
  standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/ITUlosByAgeFits")
```


## Fit with min AIC

```{r}
fitMinAIC = fit$clone()$filterModels(aic == min(aic))

bestFitTable = fitMinAIC$printDistributionDetail() %>% ungroup() %>% 
  filter(param %in% c("mean","sd")) %>% 
  mutate(ci = sprintf("%1.2f (%1.2f; %1.2f)", mean, lower, upper)) %>%
  select(-`Mean ± SD (95% CI)`, -dist,-aic,-bic,-loglik,-mean,-sd,-lower,-upper,-shift) %>% 
  pivot_wider(names_from = param, values_from = ci) %>%
  rename(`Mean (95% CI)`=mean,`SD (95% CI)`=sd) %>%
  rename(`Age group`=ageCat, N=n)
bestFitTable %>% standardPrintOutput::saveTable("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/MinAICFittedModels",defaultFontSize = 8)
```

## Survival by age and time

```{r}
surfacesDf = fitMinAIC$discreteSurvival(q = 0:60, summarise = TRUE) %>% ungroup() 
surfacesDf %>% select(-dist) %>% rename(daysPostAdmission = value) %>% write.csv(file="~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/ProbabilityInICUBedSurface.csv")
```



```{r}
surfacesDf = surfacesDf %>% mutate(days = value, cdf = 1-Quantile.0.5.survival, cdf_lo=1-Quantile.0.025.survival, cdf_hi=1-Quantile.0.975.survival) %>% filter(days<40)
                      
surf1 = ggplot(surfacesDf,
               aes(x=days,y=as.integer(ageCat),z=1-cdf, fill=1-cdf))+geom_tile()+scale_fill_gradient2(high="red",mid="yellow",low="green", midpoint=0.5, guide="none", limits=c(0,1))+xlab("days")+ylab("age")+
  metR::geom_contour2(colour="black", breaks=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9))+scale_y_continuous(
    breaks = 4:17,
    labels = c('15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64','65-69','70-74','75-79','80+'))+
  metR::geom_contour2(colour="blue", breaks=c(0.5), size=1.5)+
  metR::geom_text_contour(breaks=c(0.1,0.3,0.5,0.7,0.9),stroke=0.2)+guides(fill="none")
  # theme(axis.title=element_blank(),
  #       axis.text.y=element_blank())+guides(fill="none")


surf_lo = ggplot(surfacesDf,
                 aes(x=days,y=as.integer(ageCat),z=1-cdf_hi, fill=1-cdf_hi))+geom_tile()+scale_fill_gradient2(high="red",mid="yellow",low="green", midpoint=0.5, guide="none", limits=c(0,1))+xlab("days")+
  metR::geom_contour2(colour="black", breaks=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9))+scale_y_continuous(
    breaks = 4:17,
    labels = c('15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64','65-69','70-74','75-79','80+'))+
  metR::geom_contour2(colour="blue", breaks=c(0.5), size=1.5)+
  theme(axis.title=element_blank(),
        axis.text.y=element_blank(),
        axis.text.x=element_blank())+guides(fill="none")

surf_hi = ggplot(surfacesDf,
                 aes(x=days,y=as.integer(ageCat),z=1-cdf_lo, fill=1-cdf_lo))+geom_tile()+scale_fill_gradient2(high="red",mid="yellow",low="green", midpoint=0.5, guide="none", limits=c(0,1))+xlab("days")+
  metR::geom_contour2(colour="black", breaks=c(0.1,0.2,0.3,0.4,0.5,0.6,0.7,0.8,0.9))+scale_y_continuous(
    breaks = 4:17,
    labels = c('15-19','20-24','25-29','30-34','35-39','40-44','45-49','50-54','55-59','60-64','65-69','70-74','75-79','80+'))+
  metR::geom_contour2(colour="blue", breaks=c(0.5), size=1.5)+
  theme(axis.title=element_blank(),
        axis.text.y=element_blank(),
        axis.text.x=element_blank()
  )+guides(fill="none")

#surf = Nplot + surf1 + (surf_lo / surf_hi) + patchwork::plot_annotation(tag_levels = 'A') + patchwork::plot_layout(ncol=3,widths = c(0.3,2,1))
surf = surf1 + (surf_lo / surf_hi) + patchwork::plot_annotation(tag_levels = 'A') + patchwork::plot_layout(ncol=2,widths = c(2,1))
surf %>% 
  standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/ITUlosByAgeSurfaces") 

```


## Pre and post dexa

```{r}

dfit2 = DistributionFit$new(distributions = c("lnorm","gamma","weibull","nbinom"))
fits4 = censoredItuAdmissionToDischargeOrDeath %>% filter(patientstillonicu=="No") %>% group_by(era,ageCatBig) %>% 
   dfit2$fromUncensoredData(valueExpr = (lowerTime+upperTime)/2,bootstraps = 100)

# fits4 = censoredItuAdmissionToDischargeOrDeath %>% group_by(era,ageCatBig) %>% 
#    dfit2$fromCensoredData(lowerValueExpr = lowerTime, upperValueExpr = upperTime,bootstraps = 100)

fits4$plot(xlim=c(0,60)) %>% 
  standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/ITUlosByPrePostDex")


```

## Statistical comparison

<!-- # Post dexa -->

<!-- ```{r} -->

<!-- postDexa = DistributionFit$new(distributions = c("lnorm","gamma","weibull","nbinom")) -->

<!-- postDexDf = censoredItuAdmissionToDischargeOrDeath %>% filter(era=="post-dexa") %>% group_by(ageCatMid)  -->

<!-- postDexa$fromCensoredData(postDexDf, lowerValueExpr = lowerTime, upperValueExpr = upperTime,bootstraps = 100) -->
<!-- # postDexa$fromUncensoredData(postDexDf %>% filter(patientstillonicu=="No"), valueExpr = (lowerTime+upperTime)/2,bootstraps = 100) -->

<!-- postDexa$plot(xlim=c(0,60)) -->


<!-- ``` -->

<!-- ```{r} -->
<!-- #postDexa$filterModels(aic == min(aic)) -->
<!-- postDexa2 = postDexa$clone() -->
<!-- postDexa2$filterModels(dist == "lnorm") -->

<!-- postDexaBestFitTable = postDexa2$printDistributionDetail() %>% ungroup() %>%  -->
<!--   filter(param %in% c("mean","sd")) %>%  -->
<!--   mutate(ci = sprintf("%1.2f (%1.2f; %1.2f)", mean, lower, upper)) %>% -->
<!--   select(-`Mean ± SD (95% CI)`, -dist,-aic,-bic,-loglik,-mean,-sd,-lower,-upper,-shift) %>%  -->
<!--   pivot_wider(names_from = param, values_from = ci) %>% -->
<!--   rename(`Mean (95% CI)`=mean,`SD (95% CI)`=sd) %>% -->
<!--   rename(`Age group`=ageCatMid, N=n) -->

<!-- postDexaBestFitTable %>% standardPrintOutput::saveTable("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/LogNormPostDexaFittedModels",defaultFontSize = 8) -->
<!-- ``` -->

<!-- ```{r} -->
<!-- survPostDex = postDexa2$discreteSurvival(q = 0:60, summarise = TRUE) %>% ungroup() -->
<!-- survPostDex %>% select(-dist) %>% rename(daysPostAdmission = value) %>% write.csv(file="~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/PostDexaProbabilityInICUBedSurface.csv") -->
<!-- ``` -->

# References

<div id="refs"></div>

# Supplementary material

```{r}
suppTable = fit$printDistributionDetail() %>% ungroup() %>% 
  select(-dist,-bic,-loglik,-mean,-sd,-lower,-upper,-shift) %>% 
  rename(`Age group`=ageCat, Parameter= param, AIC= aic, N=n)  %>% group_by(`Age group`, N, AIC, Distribution) %>% arrange(AIC)
suppTable %>% standardPrintOutput::saveTableLandscape("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/SuppTable1_fullBreakdownFittedModels",defaultFontSize = 7)
```

<!-- # generate log normal curves for each age group -->

<!-- * A - "best fit" estimate -->
<!-- * B - low CrI estimate -->
<!-- * C - high CrI estimate -->

<!-- ```{r} -->
<!-- survSurf = out %>% createSurvivalSurfaces(days=40) -->
<!-- surf = plotSurvivalSurface(survSurf, "gamma") -->
<!-- surf %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/Fig4_losDensity") -->



<!-- ``` -->
<!-- # median distribution -->

<!-- ```{r} -->

<!-- largerOf = function(x,y) {return(ifelse(x>y,x,y))} -->
<!-- smallerOf = function(x,y) {return(ifelse(x<y,x,y))} -->

<!-- (ggplot(survSurf %>% filter(model=="gamma") %>% mutate(median_lo = ifelse(is.na(median_lo),0,median_lo)),aes(x=ageCat,y=median,ymin=smallerOf(median_lo,median_hi),ymax=largerOf(median_lo,median_hi), group=1))+geom_line()+geom_ribbon(alpha=0.2)+xlab("age")+ylab("median length of stay")) %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/Fig5_losMedianByAge") -->

<!-- ``` -->


<!-- # incidence surfaces  -->

<!-- more important for other transitions where we are looking at  -->

<!-- ```{r} -->
<!-- surf2 = plotIncidenceSurface(survSurf, "gamma") -->
<!-- surf2 -->
<!-- ``` -->

<!-- ```{r} -->
<!-- survSurfcsv = out %>% createSurvivalSurfaces(timepoints = 0:45) -->
<!-- write.csv( -->
<!--   survSurfcsv %>% ungroup() %>% mutate(surv = 1-cdf), -->
<!--   file="~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/itu-admission-to-itu-discharge.csv")         -->
<!-- plotProbabilityMatrix(survSurfcsv,"gamma",pExpr=1-cdf) %>% saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/daily_probability_by_age") -->

<!-- ``` -->


<!-- # validate against Bristol data -->

<!-- Do a set of Kolmogorov-Smirnov tests on different combinations of parameters and log normal distributions -->

<!-- ```{r} -->
<!-- bri = getBristolData(paths$bristol) -->
<!-- bri2 = bri %>% bristolSurvivalSubset()  -->
<!-- bri3 = bri2 %>% generateSurvivalData( -->
<!--   idVar = record_number, -->
<!--   startDateVar = admission_date, -->
<!--   endDateExpr = discharge_date, -->
<!--   statusLabels = c("censored","disch or died") -->
<!-- ) -->

<!-- doKSTest = function(d,g,paramNames,params,n) { -->
<!--   sample = d$time -->
<!--   params = as.list(params) -->
<!--   names(params) = paramNames -->
<!--   params$x = jitter(sample) -->
<!--   params$y = "plnorm" -->
<!--   # browser() -->
<!--   ks = do.call("ks.test",params) -->
<!--   return(tibble( -->
<!--     testN = length(sample), -->
<!--     referenceN = max(n), -->
<!--     pValue = ks$p.value, -->
<!--     dStatistic = ks$statistic -->
<!--   )) -->
<!-- } -->

<!-- gofBri = bri3 %>% filter(status == "disch or died") %>% group_by(ageCat) %>% group_modify(function(d,g,...) { -->
<!--   reference = out %>% filter(model == "lnorm" & ageCat == g$ageCat) -->
<!--   out = bind_rows( -->
<!--     doKSTest(d,g,reference$param, reference$value, reference$n) %>% mutate(estimate = "central"), -->
<!--     doKSTest(d,g,reference$param, reference$low_ci, reference$n) %>% mutate(estimate = "low credible"), -->
<!--     doKSTest(d,g,reference$param, reference$high_ci, reference$n) %>% mutate(estimate = "high credible") -->
<!--   ) -->
<!-- }) -->

<!-- numbers = gofBri %>% select(ageCat,estimate,test=testN,fit=referenceN) %>% pivot_longer(cols=c("test","fit"),names_to = "source", values_to = "N") -->

<!-- top = ggplot(gofBri, aes(x=ageCat,y=pValue,fill=estimate))+geom_bar(stat="identity", position="dodge", width=0.7, colour="black")+standardPrintOutput::narrowAndTall()+theme(axis.text.x = element_blank(), axis.title.x = element_blank())+ylab("P value") -->
<!-- bottom = ggplot(numbers, aes(x=ageCat,y=source,label=N, colour = source))+geom_text()+guides(colour="none")+theme(panel.grid = element_blank(),axis.title.x = element_blank())+ylab("N") -->

<!-- (top+bottom+patchwork::guide_area()+patchwork::plot_layout(ncol=1, heights = c(9,2,1), guides = "collect")) %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/ventilator-demand/parameterisation/itu-admission-to-itu-discharge/Fig6_validateBristol") -->
<!-- ``` -->

<!-- Suggests the fitted distributions to the UK wide data reasonably describes the Bristol data from 30-70 year olds. The high credible estimates are closer to the bristol data in the older age group.  -->

<!-- Possible that my exclusion of hospitals with high numbers of censoring was too enthusiastic, and this would tend to drive down the LOS. Alternatively Bristol OAPs stay longer in hospital as they don't die as quickly as the rest of the country. -->

<!-- # unsing surv -->

<!-- ```{r} -->


<!-- crlnorm <- flexsurv::flexsurvreg(Surv(time, status != "censored") ~ ageCat, data = admissionToDischargeOrDeath, dist = "gamma") -->
<!-- crlnorm -->

<!-- ``` -->