---
title: "New Variant"
author: "Rob Challen"
date: "27/11/2020"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
knitr::opts_chunk$set(echo = TRUE)
devtools::load_all("~/Git/uk-covid-datatools/")
# ukcovidtools::setup()
ukcovidtools::reload()

devtools::load_all("~/Git/standard-print-output/")
standardPrintOutput::setDefaults()
```


```{r}
sgll = dpc$spim$getSGeneLineList()
ll = dpc$spim$getLineList()
dll = dpc$spim$getDeathsLineList()
sgFrac = dpc$spim$getSDropoutFreqency(codeTypes = "LAD")
```

## What does reporting of S gene level look like across regions?

```{r}

sGeneTested = bind_rows(
  ll %>% semi_join(sgll,by="FINALID") %>% mutate(SGeneResult = "Known"),
  ll %>% anti_join(sgll,by="FINALID") %>% mutate(SGeneResult = "Unknown")
)

#summary = sGeneTested %>% group_by(pillar,NHSER_name,SGeneResult) %>% summarise(n=n(),maxDate=max(specimen_date))
#summary %>% View()
#truncate = as.Date(min(summary$maxDate),"1970-01-01")
#tmp4 = sGeneTested %>% filter(specimen_date < truncate)

testedGeography = sGeneTested %>% filter(specimen_date > "2020-10-01") %>% group_by(LTLA_code, LTLA_name, pillar, SGeneResult) %>% summarise(count = n()) %>% mutate(percent = count/sum(count))
ladMap = dpc$geog$getMap("LAD19") %>% inner_join(testedGeography, by=c("code"="LTLA_code"))
ggplot(ladMap %>% filter(SGeneResult=="Unknown"),aes(fill=(1-percent)*100))+geom_sf()+facet_wrap(vars(pillar))
```

```{r}
mutationIncidence = tsp$spim$getLineListIncidence(ll=sGeneTested %>% filter(pillar=="Pillar 2"), ageBreaks = c(5,17,23,60,75), subgroup = SGeneResult) %>% tsp$aggregateGender()

#if(mutationIncidence %>% filter(codeType=="NHSER") %>% pull(value) %>% sum() != nrow(tmp4)) warning("incidence error")

mutationIncidence2 = mutationIncidence %>% pivot_wider(names_from = subgroup,values_from=value)  %>%
  mutate(
    Known = ifelse(is.na(Known),0,Known),
    Unknown = ifelse(is.na(Unknown),0,Unknown)
  )

mutationIncidence2 = mutationIncidence2 %>% covidStandardGrouping() %>% arrange(date) %>% mutate(
  Roll.Known = stats::filter(Known,filter=rep(1,7),sides=1),
  Roll.Unknown = stats::filter(Unknown,filter=rep(1,7),sides=1)
) %>% filter(!is.na(Roll.Known) & !is.na(Roll.Unknown)) %>% ungroup()

mutationIncidence2 = mutationIncidence2 %>% mutate(
  binom::binom.confint(Roll.Known,Roll.Unknown+Roll.Known,methods = "wilson")
)

ggplot(mutationIncidence2 %>% filter(name !="Unknown (England)" & ageCat != "unknown" & date > "2020-09-01"), aes(x=date,y=mean,ymin=lower,ymax=upper,fill=ageCat,colour=ageCat))+geom_ribbon(alpha=0.4,colour=NA)+geom_line()+facet_wrap(vars(name))+ylab("P(S gene status known)")

```

## S gene detection frequency over time as function of ORF1ab and N gene cutoff

```{r}

tmp2 = bind_rows(
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 15,N_CT = 15) %>% mutate(cutoff=15),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 20,N_CT = 20) %>% mutate(cutoff=20),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 25,N_CT = 25) %>% mutate(cutoff=25),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 30,N_CT = 30) %>% mutate(cutoff=30),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 35,N_CT = 35) %>% mutate(cutoff=35),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 40,N_CT = 40) %>% mutate(cutoff=40)
)


ggplot(tmp2,aes(x=date,y=mean*100,ymin=lower*100,ymax=upper*100))+
  #geom_ribbon(alpha=0.4,mapping=aes(fill=as.factor(cutoff)))+
  geom_line(mapping=aes(colour=as.factor(cutoff)))+
  facet_grid(rows=vars(name),cols=vars(sGene))

# Beta distribution with parameters (x + 1/2, n – x + 1/2).

# a = ggplot(tmp2,aes(x=P2CH1CQ,y=P2CH3CQ,colour=sgtf))+geom_point()
# b = ggplot(tmp2,aes(x=P2CH2CQ,y=P2CH3CQ,colour=sgtf))+geom_point()
# c = ggplot(tmp2,aes(x=P2CH1CQ,y=P2CH2CQ,colour=sgtf))+geom_point()
# a+b+c+patchwork::plot_layout(ncol=3)

```

```{r}

tmp2 = bind_rows(
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 15,N_CT = 15,equivocal.rm = FALSE) %>% mutate(cutoff=15),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 20,N_CT = 20,equivocal.rm = FALSE) %>% mutate(cutoff=20),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 25,N_CT = 25,equivocal.rm = FALSE) %>% mutate(cutoff=25),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 30,N_CT = 30,equivocal.rm = FALSE) %>% mutate(cutoff=30),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 35,N_CT = 35,equivocal.rm = FALSE) %>% mutate(cutoff=35),
  dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 40,N_CT = 40,equivocal.rm = FALSE) %>% mutate(cutoff=40)
)


ggplot(tmp2,aes(x=date,y=mean*100,ymin=lower*100,ymax=upper*100))+
  #geom_ribbon(alpha=0.4,mapping=aes(fill=as.factor(cutoff)))+
  geom_line(mapping=aes(colour=as.factor(cutoff)))+
  facet_grid(rows=vars(name),cols=vars(sGene))

```

proportion of S neg dropout seems to be plateau-ing (sic?) on most recent data. The level depends on the definition of S negativity - in terms of the CT values of the other genes needed to define a positive - the "standard" definition is S_neg_30. Which gives the dropout at around 70% of the total. This could be successful prevention of spread... or maybe indication that 

## SGene versus COG Data

```{r}

cogData = dpc$datasets$getCOGUK()

cogSf = cogData %>% filter(adm1=="UK-ENG" & !is.na(latitude) & !is.na(longitude) & latitude>0) %>% group_by(latitude,longitude) %>% sf::st_as_sf(coords=c("longitude","latitude"),crs=4326) %>% mutate(name=id, code=id)
tmp = dpc$geog$getContainedIn(inputSf = cogSf,outputMapId = "NHSER20", outputIdVar = name)

# missing NHSER values are mostly in wales and scotland
# missingSf = cogData %>% filter(adm1=="UK-ENG" & !is.na(latitude) & !is.na(longitude)) %>% filter(is.na(NHSER)) %>% sf::st_as_sf(coords=c("longitude","latitude"),crs=4326) %>% mutate(name=id, code=id)
# dpc$geog$preview(mapId="NHSER20",poi = missingSf)

cogData = cogData %>% mutate(voc = lineage %>% stringr::str_detect("^B\\.1\\.1\\.7(\\..+)?$")) %>% left_join(tmp %>% rename(NHSER = to), by=c("id"="from"))
```

```{r}
# ggplot(cogData %>% filter(voc),aes(x=sample_date))+geom_histogram()
# cogData %>% filter(voc & adm1=="UK-ENG") %>% View()

vocProp = cogData %>% filter(adm1=="UK-ENG") %>% group_by(sample_date,voc) %>% count() %>% ungroup() %>% tidyr::complete(sample_date = as.Date(min(cogData$sample_date):max(cogData$sample_date),"1970-01-01"),voc, fill=list(n=0)) %>% group_by(voc) %>% arrange(sample_date) %>% mutate(Roll.n = stats::filter(n,filter=rep(1,7),sides = 1)) %>% filter(!is.na(Roll.n)) %>% group_by(sample_date) %>% mutate(binom::binom.confint(Roll.n,sum(Roll.n),methods = "wilson"))

sgene = dpc$spim$getSDropoutFreqency(codeTypes = "CTRY",ORF1ab_CT = 30,N_CT = 30) %>% select(date,sGene,mean,lower,upper)
b117 = vocProp %>% rename(date = sample_date) %>% mutate(sGene = ifelse(voc,"B.1.1.7","Non B.1.1.7")) %>% select(date,sGene,mean,lower,upper)

ggplot(bind_rows(sgene,b117) %>% filter(sGene %in% c("B.1.1.7","Negative")),aes(x=date,y=mean*100,ymin=lower*100,ymax=upper*100))+
  geom_ribbon(alpha=0.4,mapping=aes(fill=sGene),alpha=0.5)+
  geom_line(mapping=aes(colour=sGene))+coord_cartesian(xlim = as.Date(c("2020-10-01",NA)))

```

```{r}
# ggplot(cogData %>% filter(voc),aes(x=sample_date))+geom_histogram()
# cogData %>% filter(voc & adm1=="UK-ENG") %>% View()

vocPropNHSER = cogData %>% filter(adm1=="UK-ENG") %>% group_by(sample_date,voc,NHSER) %>% count() %>% ungroup() %>% tidyr::complete(sample_date = as.Date(min(cogData$sample_date):max(cogData$sample_date),"1970-01-01"),voc,NHSER, fill=list(n=0)) %>% group_by(voc,NHSER) %>% arrange(sample_date) %>% mutate(Roll.n = stats::filter(n,filter=rep(1,7),sides = 1)) %>% filter(!is.na(Roll.n)) %>% group_by(sample_date,NHSER) %>% mutate(binom::binom.confint(Roll.n,sum(Roll.n),methods = "wilson"))

sgeneNHSER = dpc$spim$getSDropoutFreqency(codeTypes = "NHSER",ORF1ab_CT = 30,N_CT = 30) %>% select(date,sGene,mean,lower,upper,NHSER=name)
b117 = vocPropNHSER %>% rename(date = sample_date) %>% mutate(sGene = ifelse(voc,"B.1.1.7","Non B.1.1.7")) %>% select(date,sGene,mean,lower,upper,NHSER) %>% filter(!is.na(NHSER))




ggplot(bind_rows(sgeneNHSER,b117) %>% filter(sGene %in% c("B.1.1.7","Negative")),aes(x=date,y=mean*100,ymin=lower*100,ymax=upper*100))+
  geom_ribbon(alpha=0.4,mapping=aes(fill=sGene),alpha=0.5)+
  geom_line(mapping=aes(colour=sGene))+coord_cartesian(xlim = as.Date(c("2020-10-01",NA)))+facet_wrap(vars(NHSER))

```



```{r}

currentLineages = bind_rows(
  cogData %>% filter(adm1=="UK-ENG" & sample_date > "2020-11-01") %>% group_by(lineage) %>% count() %>% arrange(desc(n)) %>% head(7),
  cogData %>% filter(adm1=="UK-ENG" & sample_date < "2020-03-01") %>% filter(sample_date == min(sample_date)) %>% select(lineage)
)

lineageFreq = cogData %>% filter(adm1=="UK-ENG") %>% 
  mutate(
    lineage = ifelse(lineage %in% currentLineages$lineage, lineage, "other"),
    lineage__colour= ifelse(lineage %in% currentLineages$lineage, lineage__colour, "#c0c0c0"),
  ) %>%
  group_by(sample_date,lineage,lineage__colour) %>% count() %>% ungroup() %>%
  tidyr::complete(sample_date = as.Date(min(cogData$sample_date):max(cogData$sample_date),"1970-01-01"),nesting(lineage,lineage__colour), fill=list(n=0)) %>%
  group_by(lineage,lineage__colour) %>% arrange(sample_date) %>% mutate(Roll.n = stats::filter(n, rep(1,7))) %>% filter(!is.na(Roll.n)) %>%
  group_by(sample_date) %>% mutate(binom::binom.confint(Roll.n, sum(Roll.n),methods = "wilson"))

colours = lineageFreq %>% ungroup() %>% select(lineage,lineage__colour) %>% distinct() %>% deframe()

ggplot(lineageFreq,aes(x=sample_date,fill=lineage,y=mean))+geom_area()+
  #scale_fill_manual(values = colours)
  scale_fill_brewer(palette = "Set1")+
  coord_cartesian(xlim=as.Date(c("2020-04-01",NA)))


```

```{r}

lineageFreqNHSER = cogData %>% filter(adm1=="UK-ENG") %>% 
  mutate(
    lineage = ifelse(lineage %in% currentLineages$lineage, lineage, "other"),
    lineage__colour= ifelse(lineage %in% currentLineages$lineage, lineage__colour, "#c0c0c0")
  ) %>%
  group_by(sample_date,lineage,lineage__colour,NHSER) %>% count() %>% ungroup() %>%
  tidyr::complete(sample_date = as.Date(min(cogData$sample_date):max(cogData$sample_date),"1970-01-01"),nesting(lineage,lineage__colour),NHSER, fill=list(n=0)) %>%
  group_by(lineage,lineage__colour,NHSER) %>% arrange(sample_date) %>% mutate(Roll.n = stats::filter(n, rep(1,7))) %>% filter(!is.na(Roll.n)) %>%
  group_by(sample_date,NHSER) %>% mutate(binom::binom.confint(Roll.n, sum(Roll.n),methods = "wilson"))

colours = lineageFreq %>% ungroup() %>% select(lineage,lineage__colour) %>% distinct() %>% deframe()

ggplot(lineageFreqNHSER %>% filter(!is.na(NHSER)),aes(x=sample_date,fill=lineage,y=mean))+geom_area()+
  #scale_fill_manual(values = colours)
  scale_fill_brewer(palette = "Set1")+
  coord_cartesian(xlim=as.Date(c("2020-07-01",NA)))+
  facet_wrap(vars(NHSER))


```


```{r}
cogSf = cogData %>% filter(adm1=="UK-ENG" & !is.na(latitude) & !is.na(longitude) & latitude>0) %>% group_by(latitude,longitude) %>% sf::st_as_sf(coords=c("longitude","latitude"),crs=4326) %>% mutate(name=id, code=id)
tmp = dpc$geog$getContainedIn(inputSf = cogSf,outputMapId = "NHSER20", outputIdVar = name)

# names(dpc$geog$sources$maps)
# "WD11"      "LSOA11"    "MSOA11"    "DZ11"      "HB19"      "LHB19"     "CTYUA19"   "LAD19"     "LAD20"     "CCG20"     "NHSER20"   "PHEC16"    "CTRY19"    "LGD12"     "OUTCODE"   "ISO3166_2" "ISO3166_3"
# phec16 = dpc$geog$getMap("ISO3166_2",nocache=TRUE)
# ggplot()+geom_sf(data = phec16)+geom_sf(data = cogSf)
# dpc$geog$preview(mapId="NHSER20",poi = cogSf)
```

# QA of S Gene dropout

```{r}
tmp3 = ll %>% left_join(sgll, by="FINALID", suffix=c("",".sgene"))
tmp3 = tmp3 %>% group_by(FINALID) %>% arrange(specimen_date.sgene) %>% mutate(est.testNumber = row_number()) %>%
  ungroup()

# How many do we have CT values for? & how many of those were repeat tests?
sGeneTestStatus = tmp3 %>% group_by(specimen_date,age,FINALID,NHSER_code,NHSER_name,LTLA_code,LTLA_name) %>% summarise(
  sGeneTested = all(!is.na(sgtf)),
  sGeneRepeats = n()
) %>% mutate(sGeneRepeats = sGeneRepeats-ifelse(sGeneTested,0,1))
```


```{r}
sGeneTestStatus %>% group_by(NHSER_name,sGeneTested) %>% summarise(count = n()) %>% 
  mutate(percent = sprintf("%1.2f%%",count/sum(count)*100), col = ifelse(sGeneTested, "CT known", "CT unknown")) %>% ungroup() %>% 
  select(NHSER_name,col,percent) %>% pivot_wider(names_from=col,values_from=percent)


tmp3 %>% filter(!is.na(sgtf)) %>% group_by(est.testNumber,sgtf) %>% summarise(count=n()) %>% 
  mutate(percent = sprintf("%1.2f%%",count/sum(count)*100), col = ifelse(sgtf==1, "S-", "S+")) %>% ungroup() %>% 
  select(est.testNumber,col,percent) %>% pivot_wider(names_from=col,values_from=percent)

# pillar 2 testing kit changes don't affect
tmp6 = tmp3 %>% ungroup() %>% filter(!is.na(sgtf) & specimen_date > as.Date("2020-11-01"))
tmp6 = tmp6 %>% mutate(pillar_2_testingkit = tolower(pillar_2_testingkit)) 
tmp6 %>% filter(specimen_date > as.Date("2020-12-15") & specimen_date < as.Date("2020-12-21")) %>% group_by(pillar_2_testingkit,sgtf) %>% summarise(count=n()) %>% 
  mutate(percent = sprintf("%1.2f%%",count/sum(count)*100), col = ifelse(sgtf==1, "S-", "S+"), N=sum(count)) %>% ungroup() %>% 
  select(pillar_2_testingkit,col,percent,N) %>% pivot_wider(names_from=col,values_from=percent)

```

## Repeat specimens

```{r}



byPatient = tmp3 %>% filter(!is.na(sgtf)) %>% group_by(FINALID) %>% summarise(sgtfResults = n_distinct(sgtf), sgtf=mean(sgtf), sgtfTests = n()) 
duplicates = byPatient %>% filter(sgtfTests > 1)
# of all patients most have had only one test sequenced. A small number have had a lot.
byPatient %>% group_by(sgtfTests) %>% count() %>% ungroup() %>% mutate(percent = sprintf("%1.2f%%",n/sum(n)*100))
duplicates %>% group_by(sgtfResults) %>% count() %>% ungroup() %>% mutate(percent = sprintf("%1.2f%%",n/sum(n)*100))

# excluding patients with inconsistent results 20% are S-
# in the subgroup that have repeat tests fewer - 9% are S-. If repeat tests is a function of time and S- are more recent then this is explainable by right censoring.
byPatient %>% filter(sgtfResults == 1) %>% group_by(sgtf) %>% count() %>% ungroup() %>% mutate(percent = sprintf("%1.2f%%",n/sum(n)*100))
duplicates %>% filter(sgtfResults == 1) %>% group_by(sgtf) %>% count() %>% ungroup() %>% mutate(percent  = sprintf("%1.2f%%",n/sum(n)*100))


# Find patients that switch from S+ to S-
tmp4 = tmp3 %>% dpc$spim$interpretSGene() %>%
  mutate(result = ifelse(result %in% c("S+N+ORF+","S-N+ORF+"),result,"Other")) %>%
  arrange(FINALID,specimen_date.sgene) %>% 
  mutate(
  lead.FINALID = lead(FINALID),
  lead.sgtf = lead(sgtf),
  lead.result = lead(result),
  lead.specimen_date.sgene = lead(specimen_date.sgene),
  lead.CDR_Specimen_Request_SK = lead(CDR_Specimen_Request_SK)
) %>% filter(lead.FINALID == FINALID)

repeats = tmp4 %>% 
  mutate(interval = as.numeric(lead.specimen_date.sgene-specimen_date.sgene)) %>% # was it > 12 days later
  mutate(transition = paste0(result," > ",lead.result))
  #View()
```

## Repeats and entropy

```{r}
probSgene = dpc$spim$getSDropoutFreqency(codeTypes = "CTRY",equivocal.rm = FALSE) %>%
  mutate(result = case_when(
    sGene == "Positive" ~ "S+N+ORF+",
    sGene == "Negative" ~ "S-N+ORF+",
    TRUE ~ "Other"
  ))

repeats = repeats %>% 
  left_join(probSgene, by=c("specimen_date.sgene"="date","result"="result"),suffix=c("",".prob")) %>% #P(first result observed at time t1|a result observed at time t1)
  left_join(probSgene %>% rename_with(~paste0("lead.",.x)), by=c("lead.specimen_date.sgene"="lead.date","lead.result"="lead.result"),suffix=c("",".lead.prob")) #P(second result observed at time t2|a result observed at time t2)

repeats = repeats %>% mutate(transition.prob = mean*lead.mean)

tmp = repeats %>% filter(interval>14) %>% mutate(transition = ifelse(transition %in% c("S+N+ORF+ > S-N+ORF+","S-N+ORF+ > S+N+ORF+"),"Sneg/Spos","Other")) %>% group_by(lead.specimen_date.sgene,transition) %>% summarise(entropy = -mean(transition.prob*log(transition.prob),na.rm = TRUE))
ggplot(tmp,aes(x=lead.specimen_date.sgene,y=entropy,colour=transition))+geom_line()

# test = repeats %>% group_by(transition, transition.prob, lead.specimen_date.sgene) %>% summarise(transition.count = n()) %>% ungroup() %>%
#   tidyr::complete(lead.specimen_date.sgene = as.Date(min(repeats$lead.specimen_date.sgene):max(repeats$lead.specimen_date.sgene),"1970-01-01"),nesting(transition, transition.prob), fill=list(transition.count=0)) %>%
#   group_by(transition, transition.prob) %>% arrange(lead.specimen_date.sgene) %>% mutate(Roll.transition.count = stats::filter(transition.count, filter=rep(1,7), sides = 1)) %>%
#   filter(!is.na(Roll.transition.count)) %>%
#   group_by(lead.specimen_date.sgene) %>% mutate(binom::binom.confint(Roll.transition.count, sum(Roll.transition.count), methods="wilson")) %>% 
#   rename(transition.obs = mean,transition.obs.lower = lower,transition.obs.upper = upper)
```

## potential reinfection timelines

```{r}

reinfections = repeats %>% filter(interval>14 & transition %in% c("S+N+ORF+ > S-N+ORF+","S-N+ORF+ > S+N+ORF+"))

revReinfectPlot = sgll %>% semi_join(reinfections, by="FINALID")
timelines = revReinfectPlot %>% group_by(FINALID) %>%
  summarise(xmin=as.Date(min(specimen_date),"1970-01-01"),xmax=as.Date(max(specimen_date),"1970-01-01")) %>%
  arrange(desc(xmin))

ggplot()+
  geom_errorbarh(data = timelines,mapping=aes(y=ordered(FINALID,levels=timelines$FINALID),xmin=xmin,xmax=xmax),inherit.aes = FALSE,colour="grey")+
  geom_point(data=revReinfectPlot,mapping=aes(y=ordered(FINALID,levels=timelines$FINALID),x=specimen_date,colour=as.factor(sgtf)))+
  theme(panel.grid = element_blank())


#repeats %>% filter(interval < 10) %>% group_by(transition) %>% count() %>% readr::write_csv("~/tmp/transitions.csv")
ggplot(repeats,aes(x=transition,y=interval))+geom_violin()
  
```

## Repeat tests survival model

Here we look at repeat samples available in S gene line list according to transition status. Need all SGene line list entries by FINALID, with end date as next date or censored if none

```{r}

# reinfectSurv = tmp3 %>% dpc$spim$interpretSGene() %>%
#   mutate(result = ifelse(result %in% c("S+N+ORF+","S-N+ORF+"),result,"Other")) %>%
#   arrange(FINALID,specimen_date.sgene) %>% 
#   mutate(
#   lead.FINALID = lead(FINALID),
#   lead.sgtf = lead(sgtf),
#   lead.result = lead(result),
#   lead.specimen_date.sgene = lead(specimen_date.sgene),
#   lead.CDR_Specimen_Request_SK = lead(CDR_Specimen_Request_SK)
# ) %>% mutate(
#   across(starts_with("lead"),~ifelse(lead.FINALID==FINALID,.x,NA))
# ) %>% mutate(lead.specimen_date.sgene = as.Date(lead.specimen_date.sgene,"1970-01-01"))
# 
# reinfectSurv = reinfectSurv %>% filter(result == "S+N+ORF+") %>% mutate(
#   time = as.numeric(ifelse(is.na(lead.specimen_date.sgene),Sys.Date()-specimen_date.sgene, lead.specimen_date.sgene-specimen_date.sgene)),
#   preNov = ifelse(specimen_date.sgene<"2020-11-01","pre nov","post nov"),
#   status = ifelse(is.na(lead.result),"No reinfection",lead.result) 
# ) %>% mutate(
#   status = ordered(status,levels=c("No reinfection","S-N+ORF+","S+N+ORF+","Other")),
# )

# summary(admissionToDischargeOrDeath)

# TODO: msm - use 
# survModel %>% gtsummary::tbl_regression(exp = TRUE) 


```
# Estimates of R(t)

```{r}

casesAge = dpc$spim$getLineListIncidence(codeTypes = c("CTRY"), ageBreaks = c(6,18,23,60,75), subgroup = ethnicity_final, filterExpr = asymptomatic_indicator!="Y") %>% 
      #filter(subgroup!="Y") %>% 
      tsp$aggregateGender() %>% 
      tsp$aggregateSubgroup() %>%
      dpc$demog$findDemographics()
sgFracAge = dpc$spim$getSDropoutFreqency(codeTypes = c("CTRY"), ageBreaks = c(6,18,23,60,75))

sgCasesAge = casesAge %>% left_join(sgFracAge, by=c("date","ageCat","name","code","codeType")) 

sgCasesAge = sgCasesAge %>% mutate(subgroup = sGene, original.value = value) %>% mutate(value=mean*value) %>% select(-count,-Roll.count) %>% filter(date>"2020-09-06" & ageCat != "unknown")

(sgCasesAge %>% tsp$plotIncidenceQuantiles(colour = subgroup,denominatorExpr = population/1000000))+facet_wrap(vars(ageCat),scales="free_y")+scale_y_continuous(trans="log1p")

sgCasesAge %>% tsp$plotRt(colour = subgroup,quick=TRUE)+facet_wrap(vars(ageCat),scales="free_y")
```

```{r}


tmp2 = dpc$demog$getDemographicsWithEstimatedEthnicity()
tmp3 = tmp2 %>% filter(is.finite(count)) %>% inner_join(dpc$codes$getTransitiveClosure() %>% filter(fromCodeType=="LSOA",toCodeType=="NHSER"),by=c("code"="fromCode")) %>% group_by(toCode,ethnicity_final) %>% summarise(count=sum(count))
popByRegion = tmp3 %>% rename(code = toCode, population = count) #%>% dpc$codes$findNamesByCode()


```

```{r}
casesEthnic = dpc$spim$getLineListIncidence(codeTypes = c("NHSER"), subgroup = ethnicity_final, filterExpr = asymptomatic_indicator!="Y", nocache=TRUE) %>% 
      #filter(subgroup!="Y") %>% 
      tsp$aggregateGender() %>% 
      inner_join(popByRegion,by=c("code","subgroup"="ethnicity_final"))

tmpLL = dpc$spim$getLineList()

casesEthnic %>% tsp$plotIncidenceQuantiles(denominatorExpr = population/1000000, colour=subgroup)+facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,2,5,20,50,200,500,2000,5000))

sgFracEthnic = dpc$spim$getSDropoutFreqency(codeTypes = c("NHSER"), ll=tmpLL %>% group_by(ethnicity_final),nocache=TRUE)

ggplot(sgFracEthnic %>% filter(sGene=="Negative"), aes(x=date,y=mean,ymin=lower,ymax=upper))+geom_ribbon(mapping=aes(fill=ethnicity_final),alpha=0.2)+geom_line(mapping=aes(colour=ethnicity_final))+
  facet_wrap(vars(name))

sgCasesEthnic = casesEthnic %>% left_join(sgFracEthnic %>% ungroup() %>% select(-ageCat), by=c("subgroup"="ethnicity_final","date","name","code","codeType")) 

sgCasesEthnic = sgCasesEthnic %>% mutate(original.value = value) %>% mutate(value=mean*value) %>% select(-count,-Roll.count) %>% filter(date>"2020-09-06") %>% mutate(ageCat = sGene) 

sgCasesEthnic %>% tsp$plotIncidenceQuantiles(colour = sGene,denominatorExpr = population/1000000)+facet_grid(cols=vars(subgroup),rows=vars(name),scales="free_y")+scale_y_continuous(trans="log1p", breaks = c(0,2,5,20,50,200,500,2000,5000))

sgCasesEthnic %>% filter(date>"2020-10-26") %>% tsp$plotWindowedGrowthRate(colour = sGene,quick=TRUE,rlim=c(-0.1,0.2))+facet_grid(rows=vars(subgroup),cols=vars(name),scales="free_y")

```

```{r}

cases = dpc$spim$getLineListIncidence(codeTypes = c("CTRY","NHSER"), filterExpr = asymptomatic_indicator != "Y") %>% 
      tsp$aggregateGender() %>% 
      tsp$aggregateSubgroup() %>%
      tsp$aggregateAge() %>%
      dpc$demog$findDemographics()
sgFrac = dpc$spim$getSDropoutFreqency(codeTypes = c("CTRY","NHSER"))

sgCases = cases %>% left_join(sgFrac, by=c("date","ageCat","name","code","codeType")) 

sgCases = sgCases %>% mutate(subgroup = sGene, original.value = value) %>% mutate(value=mean*value) %>% select(-count,-Roll.count) %>% filter(date>"2020-09-06")

(cases  %>% tsp$aggregateSubgroup() %>% tsp$plotIncidenceQuantiles(denominatorExpr = population/1000000))+facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,2,5,20,50,200,500,2000,5000))

(sgCases  %>% tsp$plotIncidenceQuantiles(colour=subgroup,denominatorExpr = population/1000000))+facet_wrap(vars(name))+scale_y_continuous(trans="log1p", breaks = c(0,2,5,20,50,200,500,2000,5000))

```
```{r}
sgRt = sgCases %>% tsp$logIncidenceStats() %>% tsp$estimateRtQuick() 
sgRt %>% tsp$plotRt(colour = subgroup)+facet_wrap(vars(name),scales="free_y")

sgDelta = sgRt %>% mutate(subgroup=NA) %>% ukcovidtools::covidStandardDateGrouping() %>% group_modify(function(d,g) {
  tmp = d %>% select(sGene,growth=Growth.windowed.value,R=`Mean(R)`) %>% pivot_longer(cols = -sGene) %>% pivot_wider(names_from = sGene)
  tmp = tmp %>% mutate(absolute = Negative-Positive, relative = Negative/Positive, log = log(Negative)/log(Positive)) %>% select(-Negative,-Positive) %>% mutate(name = paste0(name,".delta."))
  tmp = tmp %>% pivot_longer(cols = -name, names_to="subname",values_to="value") %>% mutate(delta = paste0(name,subname)) %>% select(-subname,-name)
  return(tmp)
})

ggplot(sgDelta %>% filter(delta %>% stringr::str_starts("R")),aes(x=date,y=value,colour=delta))+geom_line()+facet_wrap(vars(name))

ggplot(sgDelta %>% filter(delta=="growth.delta.log"),aes(x=date,y=value,colour=delta))+geom_line()+facet_wrap(vars(name))+coord_cartesian(ylim=c(-2,2))

```

# Cox Model

```{r}
ll = dpc$spim$getLineList()
dll = dpc$spim$getDeathsLineList()
#sgll3 = dpc$spim$getSGeneLineList() %>% dpc$spim$interpretSGene() %>% filter(!is.na(FINALID)) %>% group_by(FINALID) %>% arrange(desc(specimen_date)) %>% filter(row_number() == 1) %>% ungroup()
sgll3 = dpc$spim$getSGeneLineList() %>% dpc$spim$interpretSGene() %>% filter(!is.na(FINALID)) %>% group_by(FINALID) %>% filter(n() == 1) %>% ungroup()

tmp2 = dpc$demog$getDemographicsWithEstimatedEthnicity()
tmp3 = tmp2 %>% filter(is.finite(count)) %>% inner_join(dpc$codes$getTransitiveClosure() %>% filter(fromCodeType=="LSOA",toCodeType=="NHSER"),by=c("code"="fromCode")) %>% group_by(toCode,ethnicity_final) %>% summarise(count=sum(count))
popByRegion = tmp3 %>% rename(code = toCode, population = count) #%>% dpc$codes$findNamesByCode()

coxData = ll %>% left_join(sgll3, by="FINALID", suffix=c("",".sgene")) %>% left_join(popByRegion, by=c("NHSER_code"="code","ethnicity_final"))
coxData = coxData %>% left_join(dll %>% mutate(FINALID=as.numeric(finalid)), by=c("FINALID"),suffix=c("",".death"))

censorLength = 28
```

```{r}
coxData2 = coxData %>% select(FINALID, specimen_date, lab_report_date, specimen_date.sgene, dateadmission_NHSE, dod,death_type28,age,sex,imd_decile,sGene,NHSER_name,CT_N = N_L, ethnicity_final,NHSER_population=population,LTLA_name) %>% 
  mutate(
    died = !is.na(dod),
    diedWithin28days = !is.na(death_type28),
    dodAt28days = ifelse(is.na(specimen_date.sgene) | dod > specimen_date.sgene+censorLength | dod < specimen_date.sgene, NA, dod),
    censoredAt28days = ifelse(is.na(dodAt28days),0,1),
    censoredDate = pmin(Sys.Date(),specimen_date.sgene+censorLength,na.rm = TRUE),
    censoredDate2 = pmin(Sys.Date(),pmax(dateadmission_NHSE,specimen_date.sgene,na.rm = TRUE)+censorLength)
  ) %>% 
  mutate(
    reportingDelay = as.numeric(lab_report_date-specimen_date),
    time = ifelse(censoredAt28days==0, censoredDate-as.numeric(specimen_date.sgene), dodAt28days-as.numeric(specimen_date.sgene)), 
    time2 = ifelse(censoredAt28days==0, censoredDate2-as.numeric(pmax(dateadmission_NHSE,specimen_date.sgene,na.rm = TRUE)), dodAt28days-as.numeric(pmax(dateadmission_NHSE,specimen_date.sgene,na.rm = TRUE))), 
    admissionDelay = as.numeric(dateadmission_NHSE-specimen_date),
    status = censoredAt28days
  ) %>%
  mutate(
    ageCat = tsp$cutByAge(age, ageBreaks = c(60,70,80)),
    sGeneEra = case_when(
      is.na(sGene) ~ "Unknown (and P1)",
      sGene == "Negative" & specimen_date.sgene>="2020-10-01" ~ "Neg Post B.1.1.7",
      sGene == "Negative" & specimen_date.sgene<"2020-10-01" ~ "Neg Pre B.1.1.7",
      TRUE ~ as.character(sGene))
  )

#ggplot(coxData2,aes(x=admissionDelay,fill=sGene))+geom_histogram(position="dodge")
```

```{r}
coxData3 = coxData2 %>%
  filter(!is.na(imd_decile)) %>%
  filter(!is.na(sex) & sex != "Unknown") %>% mutate() %>%
  filter(!is.na(ageCat) & !is.na(age)) %>%
  filter(!is.na(ethnicity_final) & !ethnicity_final %in% c("Unknown")) %>%
  filter(is.na(admissionDelay) | admissionDelay >= 0) %>%
  filter(is.na(reportingDelay) | (reportingDelay >= 0 & reportingDelay <10) ) %>%
  mutate(
    ethnicity_final = ethnicity_final %>% forcats::as_factor() %>% forcats::fct_relevel("White"),
    sex = sex %>% forcats::as_factor(),
    imd_decile = forcats::fct_relevel(forcats::as_factor(as.numeric(imd_decile)),"5"),
    imd_decile_2 = forcats::as_factor(as.numeric(imd_decile)),
    sGene = sGene %>% forcats::fct_relevel("Positive"),
    sGeneEra = sGeneEra %>% forcats::as_factor() %>% forcats::fct_relevel("Positive"),
    relativeCopyNumber = 2^(median(CT_N,na.rm=TRUE)-CT_N),
    deathStatus = ifelse(diedWithin28days,"Dead <28 days","Other")
  )

rm(coxData,coxData2)
```

## Summary

```{r}

summaryTable = function(tmp) {
  groupPercent = function(df) df %>% summarise(N=n()) %>% mutate(`%age`=sprintf("%1.1f%%",N/sum(N)*100))
  groupMean = function(df,col) df %>% filter(is.finite({{col}})) %>% summarise(N=n(),  `mean (SD)`=sprintf("%1.1f (\u00B1%1.1f)",mean({{col}},na.rm=TRUE),sd({{col}},na.rm=TRUE)))
   bind_rows(
    tmp %>% group_by(value = NHSER_name) %>% groupPercent() %>% mutate(category="Region"),
    tmp %>% group_by(value = ethnicity_final) %>% groupPercent() %>% mutate(category="Ethnicity"),
    tmp %>% group_by(value = sGeneEra) %>% groupPercent() %>% mutate(category="S gene status"),
    tmp %>% group_by(value = deathStatus) %>% groupPercent() %>% mutate(category="Status"),
    tmp %>% group_by(value = ageCat) %>% groupPercent() %>% mutate(category="Age by category"),
    tmp %>% groupMean(age) %>% mutate(category="Age"),
    tmp %>% groupMean(CT_N) %>% mutate(category="N Gene CT"),
    tmp %>% group_by(value=imd_decile_2) %>% groupPercent() %>% mutate(category="IMD")
  ) %>% ungroup()
}

combinedSummary = summaryTable(coxData3) %>% rename_with(~paste0("All ",.x),.cols=c(N,`%age`,`mean (SD)`)) %>%
  left_join(summaryTable(coxData3 %>% filter(!is.na(sGene))) %>% rename_with(~paste0("Tested ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value")) %>%
  left_join(summaryTable(coxData3 %>% filter(diedWithin28days)) %>% rename_with(~paste0("Died ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value")) %>%
  left_join(summaryTable(coxData3 %>% filter(diedWithin28days & !is.na(sGene))) %>% rename_with(~paste0("Died + Tested ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value"))

combinedSummary %>% group_by(category) %>% standardPrintOutput::saveTableLandscape("~/Dropbox/covid19/new-variant/CombinedSummaryTable",defaultFontSize = 8)
```

## Test reporting delay versus S Gene status


```{r}
ggplot(coxData3,aes(fill=sGeneEra,x=reportingDelay))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")
  # geom_text(aes( label = scales::percent(..prop..,accuracy=1),
  #                  y= ..prop.. ), stat= "count", vjust = 0.5, hjust=-0.1, position = position_dodge(width=0.6),size=3,angle=90)
  
```

## Time to admission versus S Gene status


```{r}
ggplot(coxData3,aes(fill=sGeneEra,x=admissionDelay))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")+
  #geom_density(mapping=aes(colour=sGene),alpha=0.2)+
  #facet_wrap(vars(ethnicity_final),scales="free_y")+
  coord_cartesian(xlim=c(0,15))+facet_wrap(~sGeneEra, scales = "free_y", ncol=1)
```

## Age distributions of cases by sGene status

```{r}

tmp = dpc$demog$getDetailedDemographics()
tmp2 = tmp %>% semi_join(dpc$codes$getTransitiveClosure() %>% filter(toCode=="E92000001"),by=c("code"="fromCode")) %>% group_by(age) %>% summarise(count = sum(count)) %>% mutate(percent = count/sum(count))

ggplot(coxData3,aes(fill=sGeneEra,x=age))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  geom_line(data=tmp2,aes(x=age,y=percent),colour="grey60",inherit.aes = FALSE)+
  ylab("probability")+facet_wrap(~sGeneEra, scales = "free_y", ncol=1)
  # geom_text(aes( label = scales::percent(..prop..,accuracy=1),
  #                  y= ..prop.. ), stat= "count", vjust = 0.5, hjust=-0.1, position = position_dodge(width=0.6),size=3,angle=90)
  
```
## Age distributions of infections by sGene status who died

```{r}
ggplot(coxData3 %>% filter(diedWithin28days),aes(fill=sGeneEra,x=age))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")+facet_wrap(~sGeneEra, scales = "free_y", ncol=1)
  # geom_text(aes( label = scales::percent(..prop..,accuracy=1),
  #                  y= ..prop.. ), stat= "count", vjust = 0.5, hjust=-0.1, position = position_dodge(width=0.6),size=3,angle=90)
  
```

## Death within 28 days versus CT values / relative copy numbers

```{r}
p1 = ggplot(coxData3,aes(x=sGeneEra,colour=diedWithin28days,y=CT_N))+geom_boxplot()
p2 = ggplot(coxData3,aes(x=sGeneEra,colour=diedWithin28days,y=relativeCopyNumber))+coord_cartesian(ylim=c(0,100))+geom_boxplot()#+facet_wrap(vars(ethnicity_final))
p1+p2+patchwork::plot_annotation(tag_levels = "A")+patchwork::plot_layout(guides="collect")
```
## Cases by IMD

```{r}
ggplot(coxData3 %>% mutate(
    imd_decile=as.factor(as.numeric(as.character(imd_decile)))
  ),aes(fill=sGeneEra,x=imd_decile))+
  geom_bar(stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("cases")+facet_wrap(vars(sGeneEra), scales = "free")
```
## Deaths by IMD

```{r}
ggplot(coxData3 %>% filter(diedWithin28days) %>% mutate(
    imd_decile=as.factor(as.numeric(as.character(imd_decile)))
  ),aes(fill=sGeneEra,x=imd_decile))+
  geom_bar(stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("deaths")+facet_wrap(vars(sGeneEra), scales = "free")
```
## Cases per 1M by Ethnicity

```{r}
ggplot(
  coxData3 %>% group_by(sGeneEra,ethnicity_final,NHSER_name,NHSER_population) %>% summarise(count = n()) %>% group_by(sGeneEra,ethnicity_final) %>% summarise(per1M = sum(count)/sum(NHSER_population)*1000000, pop = sum(NHSER_population)),
  aes(fill=sGeneEra,x=ethnicity_final,y=per1M))+
  geom_bar(stat="identity",position=position_dodge(width=0.7), width = 0.6)+
  ylab("cases per 1M")+facet_wrap(vars(sGeneEra), scales = "free")
```

## Deaths per 1M by Ethnicity

```{r}
ggplot(
  coxData3 %>% filter(diedWithin28days) %>% group_by(sGeneEra,ethnicity_final,NHSER_name,NHSER_population) %>% summarise(count = n()) %>% group_by(sGeneEra,ethnicity_final) %>% summarise(per1M = sum(count)/sum(NHSER_population)*1000000),
  aes(fill=sGeneEra,x=ethnicity_final,y=per1M))+
  geom_bar(stat="identity",position=position_dodge(width=0.7), width = 0.6)+
  ylab("deaths per 1M")+facet_wrap(vars(sGeneEra), scales = "free")
```


```{r}

posSgene = coxData3 %>% filter(sGene=="Positive" & specimen_date.sgene > "2020-10-01" & age > 30)
negSgene = coxData3 %>% filter(sGene=="Negative" & specimen_date.sgene > "2020-10-01" & age > 30)

matches = negSgene %>% inner_join(posSgene, by=c("sex","imd_decile","LTLA_name","ethnicity_final","ageCat"),suffix=c("",".match")) %>% filter(
  abs(age-age.match)<2 & 
  abs(as.numeric(specimen_date.sgene-specimen_date.sgene.match)) < 4
)

set.seed(101)
matchesSelected = matches %>% group_by(FINALID) %>% sample_n(1) %>% ungroup()

matchesSelected %>% summarise(
  ageDifference = mean(age-age.match),
  specimenDateDifference = mean(as.numeric(specimen_date.sgene-specimen_date.sgene.match))
)

#matches = matches %>% group_by(FINALID) %>% mutate(matches=n()) %>% filter(matches>=2) %>% ungroup()
negMatched = matchesSelected %>% select(-ends_with(".match"))
posMatched = matchesSelected %>% select(FINALID = FINALID.match) %>% inner_join(posSgene, by="FINALID")

coxFinal = bind_rows(posMatched,negMatched) %>% mutate(sGene = sGene %>% forcats::fct_drop())

combinedSummary2 = summaryTable(coxFinal) %>% rename_with(~paste0("All ",.x),.cols=c(N,`%age`,`mean (SD)`)) %>%
  left_join(summaryTable(posMatched) %>% rename_with(~paste0("S pos ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value")) %>%
  left_join(summaryTable(negMatched) %>% rename_with(~paste0("S neg ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value")) 

combinedSummary2 %>% group_by(category) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/new-variant/CaseMatchedSummaryTable",defaultFontSize = 8)
```

## Time to admission versus S Gene status


```{r}
ggplot(coxFinal,aes(fill=sGene,x=admissionDelay))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")+
  #geom_density(mapping=aes(colour=sGene),alpha=0.2)+
  #facet_wrap(vars(ethnicity_final),scales="free_y")+
  coord_cartesian(xlim=c(0,15))+facet_wrap(~sGene, scales = "free_y", ncol=1)

ggplot(coxFinal,aes(x=sGene,colour=diedWithin28days,y=CT_N))+geom_boxplot()

ggplot(coxFinal,aes(fill=sGene,x=specimen_date.sgene))+
  geom_bar(stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")+
  facet_wrap(~sGene, scales = "free_y", ncol=1)
```

# summary(admissionToDischargeOrDeath)

# survModel = coxph(Surv(time,status) ~ rcs(age,nk=3)+sex+sGeneEra+imd_decile, coxData3)
# survModel %>% gtsummary::tbl_regression(exp = TRUE) 

```{r}
library(survival)
survModel = survival::coxph(Surv(time,status) ~ sGene, coxFinal)
survminer::ggforest(survModel)

survModel1_5 = survival::coxph(Surv(time2,status) ~ sGene, coxFinal)
survminer::ggforest(survModel1_5)

survModel2 = survival::coxph(Surv(time,status) ~ pspline(admissionDelay,4), coxFinal)
termplot(survModel2, se=TRUE, term=1, col.term=1, col.se=1, xlab="Admission delay")

survModel3 = survival::coxph(Surv(time,status) ~ sGene+age, coxFinal)
survminer::ggforest(survModel3)

survModel4 = survival::coxph(Surv(time2,status) ~ pspline(CT_N,4), coxFinal)
termplot(survModel4, se=TRUE, term=1, col.term=1, col.se=1, xlab="N Gene CT value")

survModel5 = survival::coxph(Surv(time,status) ~ sGene+CT_N, coxFinal)
survminer::ggforest(survModel5)
```