---
title: "Increased hazard of mortality in cases compatible with SARS-CoV-2 variant under investigation 202012/1 - a matched cohort study"
output:
  pdf_document:
    fig_caption: yes
    keep_tex: true
knit: (function(inputFile, encoding,...) {
  rmarkdown::render(
    inputFile,
    encoding = encoding,
    output_dir = "~/Dropbox/covid19/current-rt", output_file=paste0('current-rt-',Sys.Date(),'.pdf'))
  })
# output:
#   word_document :
#     fig_caption: yes
#     fig_width: 7
# knit: (function(inputFile, encoding,...) {
#   rmarkdown::render(
#     inputFile,
#     encoding = encoding,
#     output_dir = "~/Dropbox/covid19/new-variant", output_file=paste0('new-variant-mortality-',Sys.Date(),'.docx'))
#   })
#TODO: https://www.reed.edu/data-at-reed/software/R/markdown_multiple_reports.html
header-includes:
  \usepackage{float}
  \floatplacement{figure}{H}    

fig_width: 7
fig_height: 5
out.width: "100%"
# bibliography: current-rt.bib
# csl: current-rt.csl
# vignette: >
#  %\VignetteIndexEntry{New Variant Mortality}
#  %\VignetteEngine{knitr::rmarkdown}
#  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}

library(tidyverse)
library(patchwork)
library(survival)
library(survminer)

# optional dependency:

if (dir.exists("~/Git/standard-print-output/")) {
  devtools::load_all("~/Git/standard-print-output/")
  standardPrintOutput::setDefaults()
  spo = TRUE
} else {
  spo = FALSE
}

# devtools::load_all("~/Git/uk-covid-datatools/")
# ukcovidtools::reload()

# load functions
source("./new-variant-mortality-2.R")

```

```{r}
## Do the analysis ----
if(!exists("coxData")) 
  coxData = loadData()

ggplot(coxData,aes(x=time,fill=as.factor(status)))+geom_histogram(binwidth=1)+scale_y_log10()
```

## Main analysis

```{r}
if(!exists("run30")) 
  run30 = runAnalysis(coxData, sGeneCtThreshold = 30, ctThreshold = 30,bootstraps = 100,ageTolerance = 3,specimenDateTolerance=5,includeMatched = TRUE)

```


### Median sample KM curve

```{r}
tmp = run30$table4Data %>% group_by(boot) %>% summarise(mean.HR = mean(HR)) %>% ungroup() 
middleBoot = which.min(abs(tmp$mean.HR-median(tmp$mean.HR)))
kmfit = survminer::surv_fit(Surv(time,status) ~ sGene, run30$rawMatchedData %>% filter(boot==middleBoot))
ggs = survminer::ggsurvplot(kmfit, conf.int = TRUE)
fig1 = ggs$plot+coord_cartesian(ylim=c(0.995,1))
if(spo) {
  (fig1+standardPrintOutput::defaultFigureLayout()+standardPrintOutput::narrower())  %>% 
    standardPrintOutput::saveSixthPageFigure("~/Dropbox/covid19/new-variant/KMCurve")
} else {
  fig1
}
# rm(ggs,kmfit,fig1,tmp)
```
### Summary table

```{r}

if(spo) {
  run30$table2Data %>% mutate(
    `S pos N` = as.integer(`S pos N`),
    `S neg N` = as.integer(`S neg N`),
    `Died N` = as.integer(`Died N`),
    ) %>% group_by(category) %>% 
    standardPrintOutput::saveTable("~/Dropbox/covid19/new-variant/CaseMatchedSummaryTable",defaultFontSize = 8)
} else {
  knitr::kable(run30$table2Data)
}
```

### HR table

```{r}
summary = summariseAnalysisRun(run30)
if(spo) {
  prettyPrintSummary(summary) %>% 
    standardPrintOutput::saveTable("~/Dropbox/covid19/new-variant/HRSummaryTable",defaultFontSize = 8)
} else {
  knitr::kable(prettyPrintSummary(summary))
}

```

## Biases in data set

```{r}
#ggplot(summaryAgeTol,aes(x=ageTolerance,y=mean.ageDifference))+geom_bar(stat="identity")

## Check for residual biases in case matched data set ----
# small mean age difference and specimen date difference
# no obvious pattern in admission delays
# CT values lower in sGene negative and in those who died. This is ? feature of sGene neg infection
# specimen dates reasonably distributed in time given that the sGene negative numbers increasing. Could constrain analysis to post Dec only.

# There is a mean age difference of `r sprintf("%1.1f",deltas$ageDifference)` years.
# There is a mean specimen date difference of `r sprintf("%1.1f",deltas$specimenDateDifference)` years.

p1 = ggplot(coxFinal,aes(fill=sGene,x=admissionDelay))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")+
  coord_cartesian(xlim=c(0,15))+facet_wrap(~sGene, scales = "free_y", ncol=1)

p2 = ggplot(coxFinal,aes(x=sGene,colour=diedWithin28days,y=CT_N))+geom_boxplot()+ylab("N Gene CT value")

p3 = ggplot(coxFinal,aes(fill=sGene,x=specimen_date.sgene))+
  geom_bar(stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("count")+
  facet_wrap(~sGene, scales = "free_y", ncol=1)

fig1 = p1+p2+p3+patchwork::plot_annotation(tag_levels = "A")+patchwork::plot_layout(ncol=3,guides="collect")
```

## Sensitivity analysis

```{r}

summaryCT = NULL
for(ct in seq(20,35,1)) {
  run = runAnalysis(coxData, ctThreshold = ct,bootstraps = 25,max=Inf)
  summaryCT = summaryCT %>% bind_rows(summariseAnalysisRun(run))
}
p1 = ggplot(summaryCT %>% filter(model=="S gene only"), aes(x=ctThreshold, y=mean.HR,ymin=mean.HR.lower,ymax=mean.HR.upper))+geom_point()+geom_errorbar()+xlab("CT threshold")+ylab("Hazard rate")#+
  #geom_smooth(method = "lm",formula = y ~ poly(x, 2),se=TRUE,alpha=0.5)
  #geom_smooth(method = "lm",se=TRUE,alpha=0.5)

```

```{r}
summaryDateTol = NULL
for(dateTol in c(0:14)) {
  run = runAnalysis(coxData, specimenDateTolerance = dateTol,bootstraps = 25)
  summaryDateTol = summaryDateTol %>% bind_rows(summariseAnalysisRun(run))
}

p2 = ggplot(summaryDateTol %>% filter(model=="S gene only"), aes(x=specimenDateTolerance, y=mean.HR,ymin=mean.HR.lower,ymax=mean.HR.upper))+geom_point()+geom_errorbar()+xlab("Mismatch tolerance - sample date (days)")+ylab("Hazard rate")+
  #geom_smooth(method = "lm",formula = y ~ poly(x, 2),se=TRUE,alpha=0.5)
  geom_smooth(method = "lm",se=TRUE,alpha=0.5)+scale_x_continuous(breaks = c(0,7,14))

p4 = ggplot(summaryDateTol %>% filter(model=="S gene only"),aes(x=specimenDateTolerance,y=mean.specimenDateDifference))+geom_bar(stat="identity", width=0.7, fill="steelblue") + 
  xlab("Mismatch tolerance - sample date (days)")+ylab("Pairwise bias (days)")+scale_x_continuous(breaks = c(0,7,14))

```

```{r}

summaryAgeTol = NULL
for(ageTol in c(0,1,2,3,4,5,6,7,8,9)) {
  run = runAnalysis(coxData, ageTolerance = ageTol,bootstraps = 25)
  summaryAgeTol = summaryAgeTol %>% bind_rows(summariseAnalysisRun(run))
}

p3 = ggplot(summaryAgeTol %>% filter(model=="S gene only"), aes(x=ageTolerance, y=mean.HR,ymin=mean.HR.lower,ymax=mean.HR.upper))+geom_point()+geom_errorbar()+xlab("Mismatch tolerance - age (years)")+ylab("Hazard rate")+
  #geom_smooth(method = "lm",formula = y ~ poly(x, 2),se=TRUE,alpha=0.5)
  geom_smooth(method = "lm",se=TRUE,alpha=0.5)+scale_x_continuous(breaks = c(0,3,6,9))
```

```{r}
fig3 = p1+p2+p3+p4+patchwork::plot_annotation(tag_levels = "A")+patchwork::plot_layout(ncol=2)
if(spo) {
  fig3 %>% 
    standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/new-variant/SensitivityAnalysis")
} else {
  fig3
}
```



```{r}
# Supplementary materials
## Summary of full unmatched dataset
# combinedSummary = summaryTable(coxData3) %>% rename_with(~paste0("All ",.x),.cols=c(N,`%age`,`mean (SD)`)) %>%
#   left_join(summaryTable(coxData3 %>% filter(!is.na(sGene))) %>% rename_with(~paste0("Tested ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value")) %>%
#   left_join(summaryTable(coxData3 %>% filter(diedWithin28days)) %>% rename_with(~paste0("Died ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value")) %>%
#   left_join(summaryTable(coxData3 %>% filter(diedWithin28days & !is.na(sGene))) %>% rename_with(~paste0("Died + Tested ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value"))


## Outputs ----
# View(table1Data)
# fig1
# summary(survModel)
# summary(survModel2)

# Prep the data set.
# source("./new-variant-mortality.R")

# # grab some ethnicity data. Not going to be used at this point.
# tmp2 = dpc$demog$getDemographicsWithEstimatedEthnicity()
# tmp3 = tmp2 %>% filter(is.finite(count)) %>% inner_join(dpc$codes$getTransitiveClosure() %>% filter(fromCodeType=="LSOA",toCodeType=="NHSER"),by=c("code"="fromCode")) %>% group_by(toCode,ethnicity_final) %>% summarise(count=sum(count))
# popByRegion = tmp3 %>% rename(code = toCode, population = count) #%>% dpc$codes$findNamesByCode()
# coxData3 = coxData3 %>% left_join(popByRegion, by=c("NHSER_code"="code","ethnicity_final"))
# 
# # get overall population by region
# tmp = dpc$demog$getDetailedDemographics()
# popData = tmp %>% semi_join(dpc$codes$getTransitiveClosure() %>% filter(toCode=="E92000001"),by=c("code"="fromCode")) %>% group_by(age) %>% summarise(count = sum(count)) %>% mutate(percent = count/sum(count))
# 
# rm(coxData,coxData2)

# combinedSummary2 %>% group_by(category) %>% standardPrintOutput::saveTable("~/Dropbox/covid19/new-variant/CaseMatchedSummaryTable",defaultFontSize = 8)
# 
# km = survminer::surv_fit(Surv(time,status) ~ sGene, coxFinal)
# tmp = survminer::ggsurvplot(km,conf.int = TRUE,risk.table = "absolute")
# (tmp$plot + coord_cartesian(ylim=c(0.9925,1))) %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/new-variant/KMCurve")


```

# Results

Biases in data set need describing.

* Pillar 2 tests younger
* S Gene negativity proxy for B.1.1.7. Equivocal results
* Rationale for case matching approach.

## Age distributions of cases by sGene status

```{r}



fig1 = ggplot(coxData3,aes(fill=sGeneEra,x=age))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  geom_line(data=popData,aes(x=age,y=percent),colour="grey60",inherit.aes = FALSE)+
  ylab("probability")+facet_wrap(~sGeneEra, scales = "free_y", ncol=1)
  

fig1 %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/new-variant/ProportionByAgeGroup")
  
```

## Age distributions of infections by sGene status who died

```{r}
fig2 = ggplot(coxData3 %>% filter(diedWithin28days),aes(fill=sGeneEra,x=age))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")+facet_wrap(~sGeneEra, scales = "free_y", ncol=1)
  

fig2 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/new-variant/ProportionDeathsByAgeGroup")
```






There is a mean age difference of `r sprintf("%1.1f",deltas$ageDifference)` years.

There is a mean specimen date difference of `r sprintf("%1.1f",deltas$specimenDateDifference)` years.

## Survival models

* Basic model
* Adjusted for age to check for residual biases due to non exact matching on age
* Adjusted for N gene CT values




# Discussion

Its bad.

# Conclusion

Sigh.

# Supplementary materials


## Summary of full unmatched dataset

```{r}

combinedSummary = summaryTable(coxData3) %>% rename_with(~paste0("All ",.x),.cols=c(N,`%age`,`mean (SD)`)) %>%
  left_join(summaryTable(coxData3 %>% filter(!is.na(sGene))) %>% rename_with(~paste0("Tested ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value")) %>%
  left_join(summaryTable(coxData3 %>% filter(diedWithin28days)) %>% rename_with(~paste0("Died ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value")) %>%
  left_join(summaryTable(coxData3 %>% filter(diedWithin28days & !is.na(sGene))) %>% rename_with(~paste0("Died + Tested ",.x),.cols=c(N,`%age`,`mean (SD)`)),by=c("category","value"))

combinedSummary %>% group_by(category) %>% standardPrintOutput::saveTableLandscape("~/Dropbox/covid19/new-variant/CombinedSummaryTable",defaultFontSize = 8)
```

## Cases by IMD

```{r}
s1 = ggplot(coxData3 %>% mutate(
    imd_decile=as.factor(as.numeric(as.character(imd_decile)))
  ),aes(fill=sGeneEra,x=imd_decile))+
  geom_bar(stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("cases")+facet_wrap(vars(sGeneEra), scales = "free")

s1 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/new-variant/SGenePercentByIMD")
```

## Deaths by IMD

```{r}
s2 = ggplot(coxData3 %>% filter(diedWithin28days) %>% mutate(
    imd_decile=as.factor(as.numeric(as.character(imd_decile)))
  ),aes(fill=sGeneEra,x=imd_decile))+
  geom_bar(stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("deaths")+facet_wrap(vars(sGeneEra), scales = "free")

s2 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/new-variant/SGeneDeathsPercentByIMD")
```

## Cases per 1M by Ethnicity

```{r}
s3 = ggplot(
  coxData3 %>% group_by(sGeneEra,ethnicity_final,NHSER_name,population) %>% summarise(count = n()) %>% group_by(sGeneEra,ethnicity_final) %>% summarise(per1M = sum(count)/sum(population)*1000000, pop = sum(population)),
  aes(fill=sGeneEra,x=ethnicity_final,y=per1M))+
  geom_bar(stat="identity",position=position_dodge(width=0.7), width = 0.6)+
  ylab("cases per 1M")+facet_wrap(vars(sGeneEra), scales = "free")

s3 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/new-variant/SGeneCasesPer1MByEthnicity")
```

## Deaths per 1M by Ethnicity

```{r}
s4 = ggplot(
  coxData3 %>% filter(diedWithin28days) %>% group_by(sGeneEra,ethnicity_final,NHSER_name,population) %>% summarise(count = n()) %>% group_by(sGeneEra,ethnicity_final) %>% summarise(per1M = sum(count)/sum(population)*1000000),
  aes(fill=sGeneEra,x=ethnicity_final,y=per1M))+
  geom_bar(stat="identity",position=position_dodge(width=0.7), width = 0.6)+
  ylab("deaths per 1M")+facet_wrap(vars(sGeneEra), scales = "free")

s4 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/new-variant/SGeneDeathsPer1MByEthnicity")
```

## Test reporting delay versus S Gene status


```{r}
s5 = ggplot(coxData3,aes(fill=sGeneEra,x=reportingDelay))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")
  # geom_text(aes( label = scales::percent(..prop..,accuracy=1),
  #                  y= ..prop.. ), stat= "count", vjust = 0.5, hjust=-0.1, position = position_dodge(width=0.6),size=3,angle=90)

s5 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/new-variant/ReportingDelayBySGeneCases")
```

## Time to admission versus S Gene status


```{r}
s6 = ggplot(coxData3,aes(fill=sGeneEra,x=admissionDelay))+
  geom_bar(aes(y = ..prop..), stat="count",position=position_dodge(width=0.7), width = 0.6)+
  ylab("probability")+
  #geom_density(mapping=aes(colour=sGene),alpha=0.2)+
  #facet_wrap(vars(ethnicity_final),scales="free_y")+
  coord_cartesian(xlim=c(0,15))+facet_wrap(~sGeneEra, scales = "free_y", ncol=1)

s6 %>% standardPrintOutput::saveHalfPageFigure("~/Dropbox/covid19/new-variant/AdmissionDelayBySGeneCases")
```

## Death within 28 days versus CT values / relative copy numbers

```{r}
p1 = ggplot(coxData3,aes(x=sGeneEra,colour=diedWithin28days,y=CT_N))+geom_boxplot()
p2 = ggplot(coxData3,aes(x=sGeneEra,colour=diedWithin28days,y=relativeCopyNumber))+coord_cartesian(ylim=c(0,100))+geom_boxplot()#+facet_wrap(vars(ethnicity_final))
s7 = p1+p2+patchwork::plot_annotation(tag_levels = "A")+patchwork::plot_layout(guides="collect")

s7 %>% standardPrintOutput::saveThirdPageFigure("~/Dropbox/covid19/new-variant/NGeneCTBySGeneAndDeathStatus")
```